---
title: "Surrogate Models: Replacing Slow Simulations with Fast Emulators"
subtitle: "Gaussian process emulation lets you run thousands of 'virtual experiments' in seconds. Post 5 in the Building Digital Twin Systems series."
author: "Jong-Hoon Kim"
date: "2026-04-23"
categories: [digital twin, Gaussian process, surrogate model, emulator, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## When simulation is the bottleneck

The Ensemble Kalman Filter in [Post 4](https://www.jonghoonk.com/blog/posts/dt_software_enkf/) ran 300 particles through an SEIR model 90 times — around 27,000 model evaluations. For a compartmental ODE with five states, each evaluation takes milliseconds. Scale up to an agent-based model with 100,000 individuals, or a partial differential equation over a spatial grid, and a single run takes minutes. Running 27,000 of them is no longer feasible in real time.

A **surrogate model** (also called an *emulator*) solves this by learning the input–output map of an expensive simulator from a small number of training runs, then predicting the output for new inputs in microseconds [@sacks1989]. In the context of a digital twin:

- The **simulator** is the expensive mechanistic model (SEIR with spatial structure, IBM, PDE).
- The **surrogate** is a statistical model fitted to the simulator's outputs.
- At runtime, the twin queries the surrogate, not the simulator.

## Gaussian process regression

The **Gaussian process (GP)** [@rasmussen2006] is the dominant surrogate method in computer experiment design [@kennedy2001] because:

1. It provides not just a prediction but a **predictive distribution** — the twin knows how uncertain each emulated output is.
2. Its predictions interpolate the training data exactly (by default) — it is consistent with what the simulator actually produced.
3. The uncertainty grows in regions far from training points, giving a natural signal for when to collect more simulator runs.

A GP is a prior over functions. We write:

$$f(\mathbf{x}) \sim \mathcal{GP}(m(\mathbf{x}),\, k(\mathbf{x}, \mathbf{x}'))$$

where $m$ is the mean function (often zero) and $k$ is the **covariance kernel** — a function that measures how similar two inputs are. The workhorse kernel is the squared-exponential (RBF):

$$k(\mathbf{x}, \mathbf{x}') = \sigma_f^2 \exp\!\left(-\frac{|\mathbf{x} - \mathbf{x}'|^2}{2\ell^2}\right)$$

$\ell$ (length-scale) controls how quickly the function varies; $\sigma_f^2$ (signal variance) controls its amplitude.

Given $n$ training points $(\mathbf{X}, \mathbf{y})$, the GP posterior at a new point $\mathbf{x}^*$ is Gaussian:

$$f(\mathbf{x}^*) \mid \mathbf{X}, \mathbf{y} \;\sim\; \mathcal{N}(\mu^*, {\sigma^*}^2)$$
$$\mu^* = \mathbf{k}^{*\top} (K + \sigma_n^2 I)^{-1} \mathbf{y}$$
$${\sigma^*}^2 = k^{**} - \mathbf{k}^{*\top} (K + \sigma_n^2 I)^{-1} \mathbf{k}^*$$

where $\mathbf{k}^*$ is the vector of covariances between $\mathbf{x}^*$ and training points, $K$ is the training covariance matrix, and $\sigma_n^2$ is nugget (observation noise).

## Implementation in R

### Setting up: a 1-D toy emulator

We first build intuition with a 1-D example: emulate the peak infected count of an SIR model as a function of $\beta$.

```{r}
#| label: setup

library(ggplot2)

# SIR model: returns peak infected fraction
sir_peak <- function(beta, gamma = 0.1, S0 = 9900, I0 = 100) {
  N <- S0 + I0
  dt <- 0.1
  S <- S0; I <- I0; R <- 0
  peak <- I
  for (i in seq_len(round(200 / dt))) {
    inf <- beta * S * I / N * dt
    rec <- gamma * I * dt
    S <- S - inf; I <- I + inf - rec; R <- R + rec
    if (I > peak) peak <- I
  }
  peak / N
}
```

```{r}
#| label: training-data

set.seed(2024)

# Latin hypercube-style training points (evenly spread, small n)
n_train <- 12
beta_train <- seq(0.11, 0.55, length.out = n_train)
y_train    <- sapply(beta_train, sir_peak)

# Dense grid for truth
beta_grid <- seq(0.10, 0.60, length.out = 200)
y_true    <- sapply(beta_grid, sir_peak)
```

### Building a GP from scratch

We implement the squared-exponential GP in base R to make the mechanics transparent. In production you would use `kernlab`, `DiceKriging`, or `GPfit`.

```{r}
#| label: gp-functions

# Squared-exponential kernel
se_kernel <- function(x1, x2, ell, sigma_f) {
  sigma_f^2 * exp(-0.5 * outer(x1, x2, function(a, b) (a - b)^2) / ell^2)
}

# GP fit: return a prediction function
gp_fit <- function(x_train, y_train, ell, sigma_f, sigma_n = 1e-6) {
  K    <- se_kernel(x_train, x_train, ell, sigma_f)
  Kn   <- K + sigma_n^2 * diag(length(x_train))
  Linv <- solve(Kn)            # in production: use Cholesky
  alpha <- Linv %*% y_train

  predict <- function(x_new) {
    Ks  <- se_kernel(x_train, x_new, ell, sigma_f)   # n_train x n_new
    Kss <- se_kernel(x_new,   x_new, ell, sigma_f)   # n_new   x n_new
    mu  <- t(Ks) %*% alpha
    cov <- Kss - t(Ks) %*% Linv %*% Ks
    list(mean = as.vector(mu),
         sd   = sqrt(pmax(0, diag(cov))))
  }
  predict
}

# Fit with hand-tuned hyperparameters (ell = 0.08, sigma_f = 0.35)
gp_pred <- gp_fit(beta_train, y_train, ell = 0.08, sigma_f = 0.35)
pred     <- gp_pred(beta_grid)
```

```{r}
#| label: plot-gp-1d
#| fig-cap: "GP emulator of peak infected fraction vs transmission rate β. Grey ribbon: ±2 posterior SD. The emulator interpolates the 12 training points exactly and quantifies uncertainty between them."

df_pred <- data.frame(
  beta = beta_grid,
  mean = pred$mean,
  lo   = pred$mean - 2 * pred$sd,
  hi   = pred$mean + 2 * pred$sd,
  true = y_true
)

ggplot(df_pred, aes(x = beta)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = mean), colour = "steelblue", linewidth = 1,
            linetype = "solid") +
  geom_line(aes(y = true), colour = "firebrick", linewidth = 0.7,
            linetype = "dashed") +
  geom_point(data = data.frame(beta = beta_train, y = y_train),
             aes(x = beta, y = y), size = 2.5, colour = "black") +
  labs(x = "β (transmission rate)", y = "Peak infected fraction",
       title = "GP emulator: 1-D SIR example",
       subtitle = "Black dots = training runs  |  Dashed = true simulator  |  Blue = emulator ± 2 SD") +
  theme_minimal(base_size = 13)
```

The emulator matches the true simulator closely across the entire input range from just 12 training runs. The uncertainty ribbon widens where training points are sparse.

### 2-D emulation: (β, γ) → peak day

A more realistic scenario: emulate the **day of peak incidence** as a function of two input parameters.

```{r}
#| label: 2d-emulator

# 2-D simulator: returns peak day
sir_peak_day <- function(beta, gamma, S0 = 9900, I0 = 100) {
  N <- S0 + I0
  dt <- 0.2
  S <- S0; I <- I0; R <- 0
  peak_I <- I; peak_day <- 0; t <- 0
  for (i in seq_len(round(300 / dt))) {
    inf <- beta * S * I / N * dt
    rec <- gamma * I * dt
    S <- S - inf; I <- I + inf - rec; R <- R + rec
    t <- t + dt
    if (I > peak_I) { peak_I <- I; peak_day <- t }
  }
  peak_day
}

# Latin hypercube design (manual, 2-D)
set.seed(99)
n2 <- 30
lhs_beta  <- (sample(n2) - runif(n2)) / n2 * (0.5 - 0.15) + 0.15
lhs_gamma <- (sample(n2) - runif(n2)) / n2 * (0.15 - 0.05) + 0.05

y2_train <- mapply(sir_peak_day, lhs_beta, lhs_gamma)

# Standardise inputs to [0, 1] for the kernel
x2_train <- cbind((lhs_beta - 0.15) / 0.35,
                  (lhs_gamma - 0.05) / 0.10)

# 2-D SE kernel (vectorised: uses ||x1-x2||^2 = ||x1||^2 + ||x2||^2 - 2 x1.x2)
se2 <- function(X1, X2, ell, sigma_f) {
  X1s <- X1 / ell;  X2s <- X2 / ell
  d2  <- outer(rowSums(X1s^2), rowSums(X2s^2), "+") - 2 * X1s %*% t(X2s)
  sigma_f^2 * exp(-0.5 * d2)
}

K2    <- se2(x2_train, x2_train, ell = 0.4, sigma_f = 40)
K2n   <- K2 + 1e-4 * diag(n2)
alpha2 <- solve(K2n) %*% y2_train

# Prediction grid
g <- 40
bg <- seq(0.15, 0.50, length.out = g)
gg <- seq(0.05, 0.15, length.out = g)
grid2 <- expand.grid(beta = bg, gamma = gg)
x2_new <- cbind((grid2$beta  - 0.15) / 0.35,
                (grid2$gamma - 0.05) / 0.10)

Ks2  <- se2(x2_train, x2_new, ell = 0.4, sigma_f = 40)
grid2$pred <- as.vector(t(Ks2) %*% alpha2)
```

```{r}
#| label: plot-gp-2d
#| fig-cap: "GP emulator of peak incidence day across (β, γ) space. The emulator was trained on 30 simulator runs (LHS design); it predicts the entire surface in milliseconds."

ggplot(grid2, aes(x = beta, y = gamma, fill = pred)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Peak day") +
  geom_point(data = data.frame(beta = lhs_beta, gamma = lhs_gamma),
             aes(x = beta, y = gamma), colour = "white",
             size = 1.5, inherit.aes = FALSE) +
  labs(x = "β (transmission rate)", y = "γ (recovery rate)",
       title = "GP emulator: peak incidence day") +
  theme_minimal(base_size = 13)
```

### Hyperparameter selection via marginal likelihood

In practice, the length-scale $\ell$ and signal variance $\sigma_f$ are estimated by maximising the log marginal likelihood of the training data — the GP's built-in objective for selecting hyperparameters [@rasmussen2006]:

$$\log p(\mathbf{y} \mid \mathbf{X}, \theta) = -\frac{1}{2}\mathbf{y}^\top K_n^{-1}\mathbf{y} - \frac{1}{2}\log|K_n| - \frac{n}{2}\log 2\pi$$

```{r}
#| label: marginal-likelihood

log_ml <- function(log_ell, log_sf, x, y, sigma_n = 1e-4) {
  ell <- exp(log_ell); sf <- exp(log_sf)
  K  <- se_kernel(x, x, ell, sf) + sigma_n^2 * diag(length(x))
  # Cholesky for numerical stability
  L  <- tryCatch(chol(K), error = function(e) NULL)
  if (is.null(L)) return(-Inf)
  alpha <- backsolve(L, forwardsolve(t(L), y))
  -0.5 * sum(y * alpha) - sum(log(diag(L))) - 0.5 * length(y) * log(2 * pi)
}

# Grid search over log-hyperparameters
ell_seq <- seq(-4, 0,  length.out = 25)
sf_seq  <- seq(-2, 1,  length.out = 25)
lml_mat <- outer(ell_seq, sf_seq,
                 Vectorize(function(le, ls) log_ml(le, ls,
                           beta_train, y_train)))
best_idx <- which(lml_mat == max(lml_mat, na.rm = TRUE), arr.ind = TRUE)
ell_opt  <- exp(ell_seq[best_idx[1]])
sf_opt   <- exp(sf_seq[best_idx[2]])

cat(sprintf("Optimal ell = %.4f,  sigma_f = %.4f\n", ell_opt, sf_opt))

# Refit with optimal hyperparameters
gp_opt  <- gp_fit(beta_train, y_train, ell = ell_opt, sigma_f = sf_opt)
pred_opt <- gp_opt(beta_grid)
```

```{r}
#| label: plot-gp-opt
#| fig-cap: "GP emulator with hyperparameters selected by marginal likelihood maximisation. The length-scale adapts to the curvature of the function, reducing over-smoothing."

df_opt <- data.frame(
  beta = beta_grid,
  mean = pred_opt$mean,
  lo   = pred_opt$mean - 2 * pred_opt$sd,
  hi   = pred_opt$mean + 2 * pred_opt$sd,
  true = y_true
)

ggplot(df_opt, aes(x = beta)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "darkorange", alpha = 0.25) +
  geom_line(aes(y = mean), colour = "darkorange", linewidth = 1) +
  geom_line(aes(y = true), colour = "firebrick", linewidth = 0.7,
            linetype = "dashed") +
  geom_point(data = data.frame(beta = beta_train, y = y_train),
             aes(x = beta, y = y), size = 2.5, colour = "black") +
  labs(x = "β", y = "Peak infected fraction",
       title = "GP emulator: MLE hyperparameters",
       subtitle = sprintf("ell = %.3f,  sigma_f = %.3f", ell_opt, sf_opt)) +
  theme_minimal(base_size = 13)
```

## Active learning: where to run the next simulation

The GP posterior variance tells you exactly where the emulator is most uncertain. **Active learning** (also called sequential design) uses this to choose the next simulator run:

$$\mathbf{x}^*_{\text{next}} = \arg\max_{\mathbf{x}} \,\sigma^*(\mathbf{x})^2$$

```{r}
#| label: active-learning
#| fig-cap: "Active learning adds new training points in the region of highest GP uncertainty (red dashed line). After one step, the emulator uncertainty there shrinks substantially."

# Current uncertainty
sigma_init <- pred_opt$sd

# Pick the most uncertain point
next_idx  <- which.max(sigma_init)
beta_next <- beta_grid[next_idx]
y_next    <- sir_peak(beta_next)

cat(sprintf("Next training point: beta = %.3f, peak = %.4f\n",
            beta_next, y_next))

# Refit with the new point
x_aug <- c(beta_train, beta_next)
y_aug <- c(y_train, y_next)
gp_aug  <- gp_fit(x_aug, y_aug, ell = ell_opt, sigma_f = sf_opt)
pred_aug <- gp_aug(beta_grid)

df_aug <- data.frame(
  beta   = beta_grid,
  before = sigma_init,
  after  = pred_aug$sd
)

ggplot(df_aug, aes(x = beta)) +
  geom_line(aes(y = before, colour = "Before"), linewidth = 1) +
  geom_line(aes(y = after,  colour = "After"),  linewidth = 1,
            linetype = "dashed") +
  geom_vline(xintercept = beta_next, linetype = "dotdash",
             colour = "firebrick") +
  scale_colour_manual(values = c(Before = "steelblue", After = "darkorange")) +
  labs(x = "β", y = "Posterior SD",
       colour = NULL,
       title = "Active learning: uncertainty before and after one new run") +
  theme_minimal(base_size = 13)
```

## Where surrogates fit in a digital twin

A surrogate reduces a slow simulator to a fast lookup. In a deployed digital twin pipeline:

```
Offline (once, or periodically):
  Run simulator at N_design input configurations
        │
        ▼
  Fit GP emulator
        │
        ▼
  Store emulator (kernel matrix + hyperparameters)

Online (real-time):
  New surveillance data
        │
        ▼
  EnKF queries emulator instead of simulator  ← 1000x speedup
        │
        ▼
  Update dashboard
```

For epidemic models with complex spatial or age structure, emulation is often the only way to achieve the sub-second update times a real-time dashboard requires.

## Key design choices

**Training design**. Latin hypercube sampling (LHS) spreads training points efficiently across input space — much better than a regular grid in high dimensions. For $p$ inputs and $n$ training runs, a practical budget is $n \approx 10p$ for smooth functions.

**Kernel choice**. The squared-exponential kernel assumes infinitely differentiable outputs — appropriate for smooth ODE models. For models with discontinuities (agent-based, stochastic), the Matérn-5/2 kernel is more robust.

**Dimension reduction**. GP computational cost scales as $O(n^3)$ in the number of training points. Above $n \approx 2000$, use sparse GPs (inducing point methods) or principal component emulation for high-dimensional outputs.

**Validation**. Always hold out a test set of simulator runs and check that the true outputs fall within the GP's predictive intervals. A 95% interval that contains fewer than 85% of test points signals a mis-specified kernel or inadequate training design.

## References

::: {#refs}
:::
