---
title: "Distribution Matching Methods: From Theory to Practice"
author: "Jong-Hoon Kim"
date: "2024-10-31"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 3
    number-sections: true
    highlight-style: github
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: load-packages
#| include: true

# Load required packages
library(MASS)
library(ks)
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
```

# Introduction

When working with data from different sources or distributions, we often need to transform one distribution to match another. In this post, we'll explore various methods for distribution matching, implement them in R, and discuss their strengths and limitations.

# Distribution Matching Methods

## Quantile Matching

Quantile matching transforms data by mapping corresponding quantiles between distributions. It's a non-parametric approach that preserves the rank order of the original data.

```{r}
#| label: quantile-match
#| code-fold: show

quantile_match <- function(A, B) {
  probs <- seq(0, 1, length.out = length(A))
  sorted_A <- sort(A)
  
  B_transformed <- approx(x = probs, y = sorted_A, 
                          xout = rank(B)/(length(B) + 1),
                          method = "linear")$y
  return(B_transformed)
}
```

## Box-Cox Transformation

The Box-Cox transformation is particularly useful when you want to transform data to approximate normality before matching moments.

```{r}
#| label: boxcox-match
#| code-fold: show

boxcox_match <- function(A, B) {
  # Find optimal lambda for both distributions
  bc_A <- boxcox(A ~ 1, plotit = FALSE)
  lambda_A <- bc_A$x[which.max(bc_A$y)]
  
  # Transform to normality
  transform_boxcox <- function(x, lambda) {
    if (abs(lambda) < 1e-4) {
      log(x)
    } else {
      (x^lambda - 1) / lambda
    }
  }
  
  # Transform both samples
  A_transformed <- transform_boxcox(A, lambda_A)
  B_transformed <- transform_boxcox(B, lambda_A)  # Use same lambda
  
  # Match moments
  B_standardized <- (B_transformed - mean(B_transformed)) / sd(B_transformed)
  B_matched <- B_standardized * sd(A_transformed) + mean(A_transformed)
  
  # Inverse transform
  inverse_boxcox <- function(x, lambda) {
    if (abs(lambda) < 1e-4) {
      exp(x)
    } else {
      (lambda * x + 1)^(1/lambda)
    }
  }
  
  B_final <- inverse_boxcox(B_matched, lambda_A)
  return(B_final)
}
```

## Kernel Density-Based Transformation

This method uses kernel density estimation to transform the distributions.

```{r}
#| label: kernel-match
#| code-fold: show

kernel_density_match <- function(A, B, bw = "nrd0") {
  # Estimate densities
  density_A <- kde(A, h = bw.nrd0(A))
  density_B <- kde(B, h = bw.nrd0(B))
  
  # Calculate CDFs using numerical integration
  cdf_A <- function(x) {
    sapply(x, function(xi) {
      mean(pnorm(xi, A, density_A$h))
    })
  }
  
  cdf_B <- function(x) {
    sapply(x, function(xi) {
      mean(pnorm(xi, B, density_B$h))
    })
  }
  
  # Transform B to match A's distribution
  B_probs <- cdf_B(B)
  
  # Create quantile function for A using interpolation
  A_sorted <- sort(A)
  A_probs <- cdf_A(A_sorted)
  
  B_transformed <- approx(x = A_probs,
                          y = A_sorted,
                          xout = B_probs,
                          yleft = min(A),
                          yright = max(A))$y
  
  return(B_transformed)
}
```

## Moment Matching

A simpler approach that focuses on matching the first two moments of the distributions.

```{r}
#| label: moment-match
#| code-fold: show

moment_match <- function(A, B) {
  # Standardize B
  B_std <- (B - mean(B)) / sd(B)
  
  # Transform to match A's moments
  B_transformed <- B_std * sd(A) + mean(A)
  
  return(B_transformed)
}
```

# Comparing the Methods

Let's generate some sample data and compare all methods:

```{r}
#| label: generate-data
#| code-fold: show

# Set seed for reproducibility
set.seed(123)

# Generate sample data
A <- rnorm(1000, mean = 10, sd = 2)  # Normal distribution
B <- rexp(1000, rate = 0.1)          # Exponential distribution

# Apply all transformations
B_quantile <- quantile_match(A, B)
B_boxcox <- boxcox_match(A, B)
B_kernel <- kernel_density_match(A, B)
B_moment <- moment_match(A, B)
```

## Visual Comparison

Let's create a more elegant visualization using ggplot2:

```{r}
#| label: visualization
#| fig-cap: "Comparison of Distribution Matching Methods"
#| fig-width: 10
#| fig-height: 8

# Create a data frame for plotting
df <- data.frame(
  Original_A = A,
  Original_B = B,
  Quantile = B_quantile,
  BoxCox = B_boxcox,
  Kernel = B_kernel,
  Moment = B_moment
)

# Convert to long format
df_long <- pivot_longer(df, 
                       cols = everything(),
                       names_to = "Method",
                       values_to = "Value")

# Create the plot
ggplot(df_long, aes(x = Value, fill = Method)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~Method, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(x = "Value",
       y = "Density",
       title = "Comparison of Distribution Matching Methods",
       subtitle = "Original A (target) vs Original B and transformed distributions") +
  theme(legend.position = "none")
```

## Numerical Comparison

Let's compare some summary statistics:

```{r}
#| label: summary-stats
#| tbl-cap: "Summary Statistics Comparison"

# Function to calculate summary statistics
get_stats <- function(x) {
  c(Mean = mean(x),
    SD = sd(x),
    Median = median(x),
    Skewness = mean((x - mean(x))^3) / sd(x)^3,
    Kurtosis = mean((x - mean(x))^4) / sd(x)^4)
}

# Calculate statistics for all distributions
stats_df <- data.frame(
  Original_A = get_stats(A),
  Original_B = get_stats(B),
  Quantile = get_stats(B_quantile),
  BoxCox = get_stats(B_boxcox),
  Kernel = get_stats(B_kernel),
  Moment = get_stats(B_moment)
)

# Display the results
knitr::kable(round(stats_df, 3))
```

# Method Selection Guide

Each method has its strengths and appropriate use cases:

## Quantile Matching
- **Pros**: Preserves rank order, works with any distribution
- **Cons**: May not extrapolate well
- **Best for**: General-purpose distribution matching

## Box-Cox Transformation
- **Pros**: Works well for skewed data, preserves relationships
- **Cons**: Requires positive data, assumes underlying normality
- **Best for**: Right-skewed positive data

## Kernel Density-Based
- **Pros**: Highly flexible, handles multimodal distributions
- **Cons**: Computationally intensive, sensitive to bandwidth selection
- **Best for**: Complex, multimodal distributions

## Moment Matching
- **Pros**: Simple, fast, preserves linear relationships
- **Cons**: Only matches first two moments, assumes similar shapes
- **Best for**: Nearly normal distributions or quick approximations

# Performance Comparison

Let's compare the computational performance of these methods:

```{r}
#| label: performance
#| cache: true

library(microbenchmark)

# Benchmark the methods
bench <- microbenchmark(
  Quantile = quantile_match(A, B),
  BoxCox = boxcox_match(A, B),
  Kernel = kernel_density_match(A, B),
  Moment = moment_match(A, B),
  times = 100
)

# Plot results
autoplot(bench) +
  theme_minimal() +
  labs(title = "Performance Comparison",
       subtitle = "Time taken by each method (lower is better)")
```

# Conclusion

While quantile matching is often the default choice for distribution matching, having multiple approaches in your toolkit allows you to handle various scenarios more effectively. The choice of method should depend on:

1. Your data characteristics
2. Computational resources
3. Preservation requirements
4. Desired properties of the transformed distribution

Always validate your transformations through both visual inspection and numerical summaries to ensure the transformed distribution meets your requirements.

# References

- Box, G. E. P., & Cox, D. R. (1964). An analysis of transformations. Journal of the Royal Statistical Society: Series B (Methodological), 26(2), 211-243.
- Silverman, B. W. (1986). Density estimation for statistics and data analysis. CRC press.
- Bolstad, B. M., Irizarry, R. A., Åstrand, M., & Speed, T. P. (2003). A comparison of normalization methods for high density oligonucleotide array data based on variance and bias. Bioinformatics, 19(2), 185-193.
