---
title: "Automated Testing and CI/CD Pipelines for Epidemic Models"
subtitle: "Catching simulation bugs before they reach clients — testthat, GitHub Actions, and the discipline of green builds. Skill 1 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, CI/CD, testing, GitHub Actions, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why this matters for your business

A client's health department runs your epidemic model every morning to inform a 9 AM briefing. On Tuesday, a routine parameter update silently changes the outbreak peak by three days. Nobody notices until Friday, when the Minister of Health asks why the numbers changed.

Automated tests and continuous integration [@wilson2014] prevent this. Every code change triggers a test suite that verifies your model still produces correct output — before it ships. A **CI/CD pipeline** (Continuous Integration / Continuous Deployment) is the engineering practice that makes "green builds" a prerequisite for deployment [@fowler2006].

## The three-level testing pyramid

```
          ┌──────────────┐
          │   E2E tests   │  ← few, slow (test the full API)
          ├──────────────┤
          │ Integration   │  ← some (test model + database)
          ├──────────────┤
          │  Unit tests   │  ← many, fast (test functions)
          └──────────────┘
```

For an epidemic model service, you need:

- **Unit tests** on every model function (does `seir_step()` conserve population?)
- **Integration tests** on the API (does `POST /simulate` return valid JSON?)
- **Regression tests** comparing output to known-good baselines

## Writing tests with testthat

The `testthat` package [@wickham2011] is the standard R testing framework. Tests live in `tests/testthat/` and are discovered automatically.

```{r}
#| label: define-model

# The model function under test
seir_step <- function(S, E, I, R, beta, sigma, gamma, N = NULL) {
  if (is.null(N)) N <- S + E + I + R
  new_exp <- beta * S * I / N
  new_inf <- sigma * E
  new_rec <- gamma * I
  list(
    S = S - new_exp,
    E = E + new_exp - new_inf,
    I = I + new_inf - new_rec,
    R = R + new_rec
  )
}
```

```{r}
#| label: write-tests

# Tests using base R stopifnot() — same logic as testthat, no package needed
run_test <- function(desc, expr) {
  tryCatch({ force(expr); cat("PASS:", desc, "\n") },
           error = function(e) cat("FAIL:", desc, "—", conditionMessage(e), "\n"))
}

# Test 1: population conservation
run_test("SEIR conserves total population", {
  result  <- seir_step(S = 9800, E = 100, I = 80, R = 20,
                       beta = 0.3, sigma = 0.2, gamma = 0.1)
  N_after <- result$S + result$E + result$I + result$R
  stopifnot(abs(N_after - 9980) < 1e-8)
})

# Test 2: no growth when I = 0
run_test("no new exposures when I = 0", {
  result <- seir_step(S = 9000, E = 0, I = 0, R = 1000,
                      beta = 0.5, sigma = 0.2, gamma = 0.1)
  stopifnot(result$E == 0, result$I == 0)
})

# Test 3: subcritical epidemic declines
run_test("epidemic declines when R0 < 1", {
  state <- list(S = 9900, E = 50, I = 50, R = 0)
  for (i in seq_len(50)) {
    state <- seir_step(state$S, state$E, state$I, state$R,
                       beta = 0.05, sigma = 0.2, gamma = 0.1)
  }
  stopifnot(state$I < 50)   # infectious count must have fallen
})
```

```{r}
#| label: test-results-plot
#| fig-cap: "Regression test: model output for fixed parameters should match a stored baseline. Drift from the expected curve signals a breaking change."

library(ggplot2)

# Run a 60-day baseline and compare to stored expected peak
run_seir <- function(beta, sigma, gamma, days = 60, N = 10000, I0 = 10) {
  S <- N - I0; E <- 0; I <- I0; R <- 0
  out <- data.frame(day = 0:days, S = NA, E = NA, I = NA, R = NA)
  out[1, ] <- c(0, S, E, I, R)
  for (d in seq_len(days)) {
    new_exp <- beta * S * I / N
    new_inf <- sigma * E
    new_rec <- gamma * I
    S <- S - new_exp; E <- E + new_exp - new_inf
    I <- I + new_inf - new_rec; R <- R + new_rec
    out[d + 1, ] <- c(d, S, E, I, R)
  }
  out
}

baseline <- run_seir(beta = 0.3, sigma = 0.2, gamma = 0.1)
expected_peak_day <- baseline$day[which.max(baseline$I)]
expected_peak_I   <- max(baseline$I)

# Regression check: peak should be within 1 day and 5% of baseline
run_test("model regression: peak within expected bounds", {
  current <- run_seir(beta = 0.3, sigma = 0.2, gamma = 0.1)
  stopifnot(which.max(current$I) == which.max(baseline$I))
  stopifnot(abs(max(current$I) - expected_peak_I) / expected_peak_I < 0.01)
})

ggplot(baseline, aes(x = day)) +
  geom_line(aes(y = E, colour = "Exposed"), linewidth = 1) +
  geom_line(aes(y = I, colour = "Infectious"), linewidth = 1) +
  geom_vline(xintercept = expected_peak_day, linetype = "dashed",
             colour = "grey50") +
  annotate("text", x = expected_peak_day + 1, y = max(baseline$I) * 0.9,
           label = sprintf("Peak day %d\n(stored baseline)", expected_peak_day),
           hjust = 0, size = 3.2) +
  scale_colour_manual(values = c(Exposed = "orange", Infectious = "firebrick")) +
  labs(x = "Day", y = "Count", colour = NULL,
       title = "Regression baseline: SEIR model output") +
  theme_minimal(base_size = 13)
```

## The GitHub Actions workflow

GitHub Actions runs your tests automatically on every push. Place this file at `.github/workflows/R-tests.yml`:

```yaml
# .github/workflows/R-tests.yml
name: R model tests

on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]

jobs:
  test:
    runs-on: ubuntu-latest

    steps:
      - uses: actions/checkout@v4

      - uses: r-lib/actions/setup-r@v2
        with:
          r-version: "4.3"

      - uses: r-lib/actions/setup-r-dependencies@v2
        with:
          packages: |
            any::testthat
            any::ggplot2

      - name: Run tests
        run: Rscript -e "testthat::test_dir('tests/testthat')"
```

The key elements:

- **Trigger**: runs on every push and pull request to `main`
- **Matrix**: you can extend this to test on multiple R versions
- **Dependencies**: caches installed packages for speed
- **Fail on error**: any failed test returns a non-zero exit code, blocking the merge

## Making CI a client trust signal

A green CI badge in your README signals professionalism to technical buyers. Add it:

```markdown
[![R tests](https://github.com/your-org/dt-api/actions/workflows/R-tests.yml/badge.svg)](...)
```

More importantly, automated tests let you confidently update dependencies, refactor internals, and add features — knowing the client's production output will not change unless you explicitly change it.

**The rule of thumb**: if a bug would embarrass you in front of a client, there should be a test for it.

## References

::: {#refs}
:::
