---
title: "Regularization in statistical models"
author: "Jong-Hoon Kim"
date: "2025-02-02"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: false
    highlight-style: github
    fig-width: 8
    fig-height: 6
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: load-packages

library(tidyverse)
library(glmnet)
library(caret)
library(recipes)
set.seed(123)
```

## Introduction

Regularization is a fundamental technique in statistical modeling that helps prevent overfitting and improves model generalization. This blog post explores different regularization methods, their implementation in R, and their practical applications, with a special focus on scenarios where the number of parameters exceeds the number of observations (p > n).

## Understanding Regularization

Regularization works by adding a penalty term to the model's loss function, effectively constraining the magnitude of model coefficients. The three most common types are:

1. Ridge (L2)
2. Lasso (L1)
3. Elastic Net (combination of L1 and L2)

## Simulation Study: When p > n

Let's first create a scenario where we have more parameters than observations:

```{r}
#| label: simulate-data

# Generate synthetic data
n <- 50  # number of observations
p <- 100 # number of predictors

# Create sparse true coefficients (most are zero)
true_beta <- rep(0, p)
true_beta[1:5] <- c(1.5, -0.8, 1.2, -0.5, 0.9)

# Generate predictor matrix
X <- matrix(rnorm(n * p), nrow = n)
# Generate response with some noise
y <- X %*% true_beta + rnorm(n, sd = 0.5)

# Convert to data frame
data <- as.data.frame(X)
data$y <- y
```

## Ridge Regression (L2)

Ridge regression adds the sum of squared coefficients to the loss function:

```{r}
#| label: ridge

# Prepare data for glmnet
x_matrix <- as.matrix(data[, -ncol(data)])
y_vector <- data$y

# Fit ridge regression
ridge_fit <- glmnet(x_matrix, y_vector, alpha = 0, 
                    lambda = exp(seq(-3, 5, length.out = 100)))

# Cross-validation to find optimal lambda
cv_ridge <- cv.glmnet(x_matrix, y_vector, alpha = 0)

# Plot cross-validation results
plot(cv_ridge)
```

## Lasso Regression (L1)

Lasso regression uses absolute value penalties and can perform variable selection:

```{r}
#| label: lasso

# Fit lasso regression
lasso_fit <- glmnet(x_matrix, y_vector, alpha = 1,
                    lambda = exp(seq(-3, 5, length.out = 100)))

# Cross-validation
cv_lasso <- cv.glmnet(x_matrix, y_vector, alpha = 1)

# Plot solution path
plot(lasso_fit, xvar = "lambda", label = TRUE)
```

## Elastic Net

Elastic Net combines both L1 and L2 penalties:

```{r}
#| label: elastic-net

# Fit elastic net with alpha = 0.5
enet_fit <- glmnet(x_matrix, y_vector, alpha = 0.5,
                   lambda = exp(seq(-3, 5, length.out = 100)))

# Cross-validation
cv_enet <- cv.glmnet(x_matrix, y_vector, alpha = 0.5)

# Compare coefficients at optimal lambda
coef_comparison <- data.frame(
  true = true_beta,
  ridge = as.vector(coef(cv_ridge, s = "lambda.min"))[-1],
  lasso = as.vector(coef(cv_lasso, s = "lambda.min"))[-1],
  elastic = as.vector(coef(cv_enet, s = "lambda.min"))[-1]
)

# Plot first 20 coefficients
head(coef_comparison, 20)
```

## Regularization in p > n Scenarios

When the number of parameters (p) exceeds the number of observations (n), regularization becomes not just useful but essential. Here's why:

1. **Mathematical Necessity**: Without regularization, the problem is ill-posed as there are infinitely many solutions that perfectly fit the training data.

2. **Variance Reduction**: Regularization helps reduce the variance of parameter estimates, which is particularly important when we have limited data.

3. **Feature Selection**: Lasso and Elastic Net can help identify the most important features, which is crucial when dealing with high-dimensional data.

Let's demonstrate this with a comparison of prediction errors:

```{r}
#| label: prediction-comparison

# Generate test data
X_test <- matrix(rnorm(n * p), nrow = n)
y_test <- X_test %*% true_beta + rnorm(n, sd = 0.5)

# Calculate MSE for each method
mse <- data.frame(
  Method = c("Ridge", "Lasso", "Elastic Net"),
  MSE = c(
    mean((predict(cv_ridge, newx = X_test) - y_test)^2),
    mean((predict(cv_lasso, newx = X_test) - y_test)^2),
    mean((predict(cv_enet, newx = X_test) - y_test)^2)
  )
)

print(mse)
```

## Best Practices and Recommendations

When working with p > n scenarios:

1. **Start with Cross-Validation**: Always use cross-validation to select the optimal regularization parameter (λ).

2. **Consider Multiple Methods**: Try different regularization approaches as their performance can vary depending on the underlying data structure.

3. **Scale Features**: Standardize predictors before applying regularization to ensure fair penalization.

4. **Monitor Sparsity**: If interpretability is important, prefer Lasso or Elastic Net as they can produce sparse solutions.

## Conclusion

Regularization is particularly crucial in p > n scenarios, where it helps:
- Prevent perfect but meaningless fits to training data
- Reduce model complexity
- Identify important features
- Improve prediction accuracy on new data

The choice between Ridge, Lasso, and Elastic Net depends on your specific needs regarding sparsity, prediction accuracy, and interpretability.

## References

1. Hastie, T., Tibshirani, R., & Friedman, J. (2009). The Elements of Statistical Learning.
2. Zou, H., & Hastie, T. (2005). Regularization and variable selection via the elastic net.
3. James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013). An Introduction to Statistical Learning.
