---
title: "Usage-Based Billing: Metering API Calls and Compute"
subtitle: "You cannot charge fairly by simulation run without first measuring what each client consumes. Metering infrastructure, pricing models, and Stripe integration. Skill 19 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, billing, pricing, Stripe, SaaS, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why metering matters

You sign a contract with a health department for $2,000/month. A year in, you realise one client is running 5× more EnKF jobs than any other — and consuming 60% of your compute budget. Without usage data you cannot renegotiate, cannot offer volume tiers, and cannot identify your most profitable clients.

**Usage metering** is the instrumentation layer that makes fair, data-driven pricing possible [@stripe2011]. It is also the foundation for the common SaaS pricing models: per-seat, per-API-call, or compute-hour billing.

## Pricing model options

```{r}
#| label: pricing-models
#| fig-cap: "Three SaaS pricing models and their revenue predictability vs. usage alignment trade-offs. Flat subscriptions are easiest to sell; usage-based aligns incentives best; hybrid is most common in practice."

library(ggplot2)

models <- data.frame(
  model       = c("Flat subscription\n$2k/month",
                  "Pure usage-based\n$0.10/EnKF run",
                  "Hybrid\n$500 base +\n$0.05/run"),
  predictability = c(5, 1, 3),
  alignment      = c(1, 5, 4)
)

ggplot(models, aes(x = predictability, y = alignment, label = model)) +
  geom_point(size = 5, colour = "steelblue") +
  geom_text(vjust = -0.9, size = 3.5) +
  scale_x_continuous(name = "Revenue predictability (seller)", limits = c(0, 6)) +
  scale_y_continuous(name = "Usage alignment (buyer)",         limits = c(0, 6)) +
  geom_vline(xintercept = 3, linetype = "dashed", colour = "grey60") +
  geom_hline(yintercept = 3, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 4.5, y = 5.5, label = "Best for\nboth (rare)",
           size = 3, colour = "grey50") +
  labs(title = "Pricing model trade-offs") +
  theme_minimal(base_size = 13)
```

## Building the metering layer

Every API call and simulation job should emit a usage event to the database (from Skill 12, TimescaleDB):

```python
# middleware.py — FastAPI middleware that logs every request
from fastapi import Request
import time

async def metering_middleware(request: Request, call_next):
    start   = time.time()
    response = await call_next(request)
    duration_ms = round((time.time() - start) * 1000)

    # Write usage event (async, non-blocking)
    await write_usage_event(
        tenant_id   = request.state.tenant_id,   # set by auth middleware
        event_type  = "api_call",
        endpoint    = request.url.path,
        duration_ms = duration_ms,
        status_code = response.status_code
    )
    return response
```

For EnKF jobs, record compute units:

```python
@celery_app.task
def run_enkf_task(obs, location_id, n_ens, tenant_id):
    start = time.time()
    result = run_enkf(obs, n_ens)
    compute_seconds = time.time() - start

    # Compute units = particles × steps × time
    compute_units = n_ens * len(obs) * compute_seconds / 1000

    write_usage_event(
        tenant_id     = tenant_id,
        event_type    = "enkf_run",
        compute_units = compute_units,
        n_particles   = n_ens,
        n_steps       = len(obs)
    )
    return result
```

## Monthly billing calculation in R

```{r}
#| label: billing-calc

set.seed(42)

# Simulate one month of usage events across three tenants
n_events <- 200
usage <- data.frame(
  tenant_id    = sample(c("dept_a","dept_b","dept_c"), n_events, replace=TRUE,
                         prob = c(0.5, 0.3, 0.2)),
  event_type   = sample(c("api_call","enkf_run","report_gen"), n_events,
                         replace=TRUE, prob=c(0.7, 0.2, 0.1)),
  compute_units = NA_real_
)
usage$compute_units[usage$event_type == "api_call"]   <- runif(sum(usage$event_type == "api_call"), 0.001, 0.01)
usage$compute_units[usage$event_type == "enkf_run"]   <- runif(sum(usage$event_type == "enkf_run"), 1, 10)
usage$compute_units[usage$event_type == "report_gen"] <- runif(sum(usage$event_type == "report_gen"), 0.1, 0.5)

# Pricing: $500 base + $0.05 per compute unit
BASE_FEE         <- 500
PRICE_PER_UNIT   <- 0.05

billing <- aggregate(compute_units ~ tenant_id, data = usage, sum)
billing$base_fee       <- BASE_FEE
billing$usage_charge   <- round(billing$compute_units * PRICE_PER_UNIT, 2)
billing$total          <- billing$base_fee + billing$usage_charge
billing$compute_units  <- round(billing$compute_units, 1)

knitr::kable(billing[, c("tenant_id","compute_units","base_fee",
                           "usage_charge","total")],
             col.names = c("Tenant","Compute units","Base ($)","Usage ($)","Total ($)"),
             caption = "Monthly invoice — hybrid pricing model")
```

```{r}
#| label: usage-breakdown
#| fig-cap: "Compute unit consumption by tenant and event type. dept_a dominates because it runs more EnKF jobs — the higher-cost event type. This data justifies tier differentiation."

library(ggplot2)
library(dplyr)

usage_summary <- usage |>
  group_by(tenant_id, event_type) |>
  summarise(total_cu = sum(compute_units, na.rm=TRUE), .groups="drop")

ggplot(usage_summary, aes(x = tenant_id, y = total_cu, fill = event_type)) +
  geom_col(position = "stack", width = 0.6) +
  scale_fill_manual(values = c(api_call   = "steelblue",
                                enkf_run   = "firebrick",
                                report_gen = "darkorange"),
                    name = "Event type") +
  labs(x = "Tenant", y = "Compute units",
       title = "Monthly compute consumption by tenant and type") +
  theme_minimal(base_size = 13)
```

## Stripe integration

Stripe [@stripe2011] handles the payment side. For usage-based billing, use Stripe's Metered Billing:

```python
import stripe
stripe.api_key = os.environ["STRIPE_SECRET_KEY"]

# At billing period end, report usage to Stripe
def report_usage_to_stripe(tenant_id: str, compute_units: float):
    subscription_item_id = get_stripe_sub_item(tenant_id)
    stripe.SubscriptionItem.create_usage_record(
        subscription_item_id,
        quantity  = round(compute_units * 1000),   # Stripe uses integers
        timestamp = int(time.time()),
        action    = "increment"
    )
```

Stripe invoices the customer automatically at the end of the billing cycle and handles payment retries, failed payments, and dunning.

## Tiers: a practical starting point

| Tier | Price | Included | Overage |
|------|-------|----------|---------|
| Starter | $500/mo | 100 compute units | $0.10/unit |
| Professional | $1,500/mo | 500 units | $0.05/unit |
| Enterprise | $4,000/mo | 2,000 units | $0.03/unit |

Track which tier each tenant is on and apply the correct overage rate automatically.

## References

::: {#refs}
:::
