---
title: "Hybrid Forecasting: Combining Mechanistic Models with Machine Learning"
subtitle: "Pure SEIR models shine at long horizons; ML wins at short ones. Ensembling both gives clients the best of both worlds. Skill 5 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, forecasting, machine learning, ensemble, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The forecasting horizon problem

Mechanistic SEIR models are powerful at **long horizons** (4–12 weeks) because they encode biological constraints: the epidemic cannot grow faster than $R_0$ allows, and it must eventually peak and decline. But at **short horizons** (1–7 days), the mechanistic model's forecast is often less accurate than a simple time-series extrapolation that just follows the recent trend.

The US CDC's collaborative influenza forecasting project found that ensemble methods combining multiple model types outperformed any single model [@reich2019], a result replicated for COVID-19 [@ray2022]. Combining your mechanistic digital twin with a lightweight statistical learner is therefore both technically sound and a strong product story.

## Three forecasting components

We build three forecasters and combine them:

1. **Mechanistic**: project the EnKF state estimate forward using the SEIR model
2. **Statistical**: exponential smoothing on recent case counts (captures trend + noise)
3. **Ensemble**: weighted average, weights learned from past forecast errors

```{r}
#| label: generate-truth

set.seed(42)
n_total <- 90
N_pop   <- 100000

# Simulate true SEIR trajectory
seir_step <- function(S, E, I, R, beta, sigma = 0.2, gamma = 0.12) {
  N   <- S + E + I + R
  new_e <- beta * S * I / N
  new_i <- sigma * E
  new_r <- gamma * I
  list(S = S - new_e, E = E + new_e - new_i,
       I = I + new_i - new_r, R = R + new_r,
       new_cases = new_i)
}

state <- list(S = N_pop - 30, E = 20, I = 10, R = 0)
true_cases <- numeric(n_total)
for (d in seq_len(n_total)) {
  state <- seir_step(state$S, state$E, state$I, state$R, beta = 0.32)
  true_cases[d] <- rpois(1, max(1, state$new_cases))
}

# Training window: days 1-75; evaluate on days 76-90
n_train  <- 75
n_eval   <- n_total - n_train
obs      <- true_cases[seq_len(n_train)]
actuals  <- true_cases[(n_train + 1):n_total]
```

```{r}
#| label: forecasters

# --- Mechanistic forecast: project SEIR forward from current state ---
seir_forecast <- function(obs_vec, horizon, beta = 0.32,
                           sigma = 0.2, gamma = 0.12, N = 100000) {
  # Rough state initialisation from last 14 days
  recent_inf <- mean(tail(obs_vec, 14))
  I_est <- recent_inf / sigma
  E_est <- I_est * sigma / sigma  # E ~ I at quasi-steady state
  R_est <- sum(obs_vec) * 0.9
  S_est <- N - E_est - I_est - R_est
  state <- list(S = max(S_est, 1), E = E_est, I = I_est, R = max(R_est, 0))
  fcst  <- numeric(horizon)
  for (h in seq_len(horizon)) {
    state <- seir_step(state$S, state$E, state$I, state$R, beta, sigma, gamma)
    fcst[h] <- state$new_cases
  }
  fcst
}

# --- Statistical: exponential smoothing (Holt-Winters level + trend) ---
hw_forecast <- function(obs_vec, horizon) {
  # Holt two-parameter (level + trend)
  alpha <- 0.25; beta_hw <- 0.1
  n  <- length(obs_vec)
  L  <- obs_vec[1]; T <- 0
  for (i in 2:n) {
    L_prev <- L; L <- alpha * obs_vec[i] + (1 - alpha) * (L + T)
    T <- beta_hw * (L - L_prev) + (1 - beta_hw) * T
  }
  fcst <- pmax(0, L + T * seq_len(horizon))
  fcst
}

# Generate 15-day forecasts
horizon    <- n_eval
fcst_seir  <- seir_forecast(obs, horizon)
fcst_hw    <- hw_forecast(obs, horizon)
```

```{r}
#| label: ensemble
#| fig-cap: "Hybrid ensemble (purple) vs individual forecasters: mechanistic SEIR (blue) and exponential smoothing (orange). The ensemble minimises error by blending the biological constraints of SEIR with the data-driven trend of HW."

# Simple equal-weight ensemble (in production, learn weights from holdout)
fcst_ensemble <- 0.55 * fcst_seir + 0.45 * fcst_hw

# Evaluate
rmse <- function(pred, obs) sqrt(mean((pred - obs)^2))
mae  <- function(pred, obs) mean(abs(pred - obs))

scores <- data.frame(
  Model = c("SEIR", "Holt-Winters", "Ensemble"),
  RMSE  = round(c(rmse(fcst_seir, actuals),
                  rmse(fcst_hw,   actuals),
                  rmse(fcst_ensemble, actuals)), 1),
  MAE   = round(c(mae(fcst_seir, actuals),
                  mae(fcst_hw,   actuals),
                  mae(fcst_ensemble, actuals)), 1)
)
knitr::kable(scores, caption = "Forecast accuracy (15-day horizon)")

library(ggplot2)
days_eval <- (n_train + 1):n_total
df_plot <- data.frame(
  day      = rep(days_eval, 4),
  cases    = c(actuals, fcst_seir, fcst_hw, fcst_ensemble),
  model    = rep(c("Actual", "SEIR", "Holt-Winters", "Ensemble"),
                 each = n_eval)
)
df_plot$model <- factor(df_plot$model,
                        levels = c("Actual","SEIR","Holt-Winters","Ensemble"))

ggplot(df_plot, aes(x = day, y = cases, colour = model, linetype = model)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c(Actual        = "grey30",
                                 SEIR          = "steelblue",
                                 `Holt-Winters` = "darkorange",
                                 Ensemble      = "mediumpurple")) +
  scale_linetype_manual(values = c(Actual        = "solid",
                                   SEIR          = "dashed",
                                   `Holt-Winters` = "dashed",
                                   Ensemble      = "solid")) +
  labs(x = "Day", y = "New cases", colour = NULL, linetype = NULL,
       title = "15-day forecast: mechanistic vs statistical vs ensemble") +
  theme_minimal(base_size = 13)
```

## Learning ensemble weights from past errors

Equal weights are a starting point. Optimal weights minimise squared forecast error on a holdout window:

```{r}
#| label: optimal-weights

# Optimise weights w such that w*SEIR + (1-w)*HW minimises RMSE
opt <- optimise(
  f = function(w) rmse(w * fcst_seir + (1 - w) * fcst_hw, actuals),
  interval = c(0, 1)
)
cat(sprintf("Optimal SEIR weight: %.2f  |  HW weight: %.2f\n",
            opt$minimum, 1 - opt$minimum))
cat(sprintf("Ensemble RMSE at optimal weights: %.1f\n", opt$objective))
```

With more models (random forest, ARIMA, Facebook Prophet), stack them using non-negative least squares or Bayesian model averaging. The `forecastHybrid` R package automates this pipeline [@hyndman2008].

## The product pitch

"Our forecast combines a calibrated SEIR model with a data-driven component. At 1–3 day horizons it tracks the current trend; at 2–8 week horizons the mechanistic biology takes over. Neither approach alone achieves both."

This narrative is technically accurate, scientifically grounded, and differentiates you from both pure statisticians and pure modellers.

## References

::: {#refs}
:::
