---
title: "Data Validation: Catching Bad Surveillance Data Before It Corrupts Your Model"
subtitle: "Schema checks, range guards, and anomaly flags at the ingestion boundary — because garbage in means garbage forecasts out. Skill 4 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, data validation, data quality, R, surveillance]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The silent failure mode

The EnKF from Post 4 receives daily case counts and updates its state estimate. What happens when the surveillance system sends a negative count (database export error), a missing row (reporting gap), or a value of 99999 (sentinel for "unknown")? The filter silently produces a nonsensical state estimate, which propagates into the 14-day forecast, which the health minister uses to decide whether to close schools.

Data validation [@stodden2014] is the defensive layer between raw surveillance feeds and your model engine. It is not glamorous, but it is the difference between a system that fails loudly and safely versus one that fails silently and expensively.

## The validation contract

Every data ingestion point should enforce:

1. **Schema** — required columns present, correct types
2. **Range** — values within physically plausible bounds
3. **Completeness** — no unexpected gaps in the time series
4. **Consistency** — values coherent with recent history (no 10× jumps without a flag)
5. **Provenance** — source, submission timestamp, data version recorded

```{r}
#| label: setup

library(ggplot2)
library(dplyr)

# Simulate incoming surveillance data with deliberate errors
set.seed(42)
n <- 30
dates <- seq(as.Date("2024-03-01"), by = "day", length.out = n)
cases <- c(rpois(10, 45), rpois(9, 80), -5, rpois(5, 120),
           NA, 99999, rpois(3, 90))

df_raw <- data.frame(
  date        = dates,
  location_id = "district_a",
  new_cases   = cases,
  source      = "surveillance_v1"
)
```

## Building a validation function

```{r}
#| label: validator

validate_surveillance <- function(df,
                                  max_plausible = 50000,
                                  max_jump_ratio = 10,
                                  sentinel_values = c(99999, -1, 9999)) {

  issues <- list()

  # 1. Required columns
  required <- c("date", "location_id", "new_cases")
  missing_cols <- setdiff(required, names(df))
  if (length(missing_cols) > 0) {
    issues[["missing_columns"]] <- missing_cols
  }

  # 2. Missing values
  na_rows <- which(is.na(df$new_cases))
  if (length(na_rows) > 0) {
    issues[["missing_values"]] <- data.frame(
      row = na_rows, date = df$date[na_rows]
    )
  }

  # 3. Negative counts
  neg_rows <- which(!is.na(df$new_cases) & df$new_cases < 0)
  if (length(neg_rows) > 0) {
    issues[["negative_counts"]] <- data.frame(
      row = neg_rows, date = df$date[neg_rows],
      value = df$new_cases[neg_rows]
    )
  }

  # 4. Sentinel values
  sentinel_rows <- which(df$new_cases %in% sentinel_values)
  if (length(sentinel_rows) > 0) {
    issues[["sentinel_values"]] <- data.frame(
      row = sentinel_rows, date = df$date[sentinel_rows],
      value = df$new_cases[sentinel_rows]
    )
  }

  # 5. Implausibly large values
  big_rows <- which(!is.na(df$new_cases) & df$new_cases > max_plausible)
  if (length(big_rows) > 0) {
    issues[["implausible_values"]] <- data.frame(
      row = big_rows, date = df$date[big_rows],
      value = df$new_cases[big_rows]
    )
  }

  # 6. Anomalous jumps (consecutive-day ratio)
  clean_cases <- df$new_cases
  clean_cases[is.na(clean_cases) | clean_cases < 0 |
                clean_cases %in% sentinel_values] <- NA
  ratio <- clean_cases[-1] / pmax(clean_cases[-length(clean_cases)], 1)
  jump_rows <- which(!is.na(ratio) & ratio > max_jump_ratio) + 1
  if (length(jump_rows) > 0) {
    issues[["anomalous_jumps"]] <- data.frame(
      row = jump_rows, date = df$date[jump_rows],
      value = df$new_cases[jump_rows],
      ratio = round(ratio[jump_rows - 1], 1)
    )
  }

  list(n_rows = nrow(df), n_issues = length(issues), issues = issues)
}

result <- validate_surveillance(df_raw)
cat(sprintf("Rows: %d | Issue types found: %d\n",
            result$n_rows, result$n_issues))
for (nm in names(result$issues)) {
  cat("\n Issue:", nm, "\n")
  print(result$issues[[nm]])
}
```

```{r}
#| label: clean-data
#| fig-cap: "Raw (red) vs validated-and-cleaned (blue) surveillance data. Negative values, sentinels, and anomalous spikes are flagged; clean rows are passed to the model."

# Apply cleaning: replace flagged values with NA for imputation
df_clean <- df_raw
flag_rows <- c(
  result$issues$missing_values$row,
  result$issues$negative_counts$row,
  result$issues$sentinel_values$row,
  result$issues$anomalous_jumps$row
)
flag_rows <- unique(flag_rows[!is.na(flag_rows)])
df_clean$new_cases[flag_rows] <- NA
df_clean$flagged <- seq_len(nrow(df_clean)) %in% flag_rows

# Simple linear interpolation for missing values
df_clean$new_cases_imputed <- df_clean$new_cases
na_idx <- which(is.na(df_clean$new_cases_imputed))
if (length(na_idx) > 0) {
  df_clean$new_cases_imputed <- approx(
    x   = seq_len(nrow(df_clean))[!is.na(df_clean$new_cases)],
    y   = df_clean$new_cases[!is.na(df_clean$new_cases)],
    xout = seq_len(nrow(df_clean))
  )$y
}

ggplot(df_clean, aes(x = date)) +
  geom_col(aes(y = pmax(0, new_cases, na.rm = TRUE)),
           fill = "firebrick", alpha = 0.4, width = 0.8) +
  geom_line(aes(y = new_cases_imputed),
            colour = "steelblue", linewidth = 1.2) +
  geom_point(data = df_clean[df_clean$flagged, ],
             aes(y = new_cases_imputed),
             shape = 4, size = 3, colour = "firebrick", stroke = 1.5) +
  labs(x = NULL, y = "Daily cases",
       title = "Data validation: flagged values (✕) replaced by interpolation",
       subtitle = "Red bars = raw input  |  Blue line = cleaned/imputed  |  ✕ = flagged row") +
  theme_minimal(base_size = 13)
```

## Validation in the Python ingestion pipeline

For the production pipeline (FastAPI + TimescaleDB), use Pydantic to enforce schema and ranges at the API boundary:

```python
from pydantic import BaseModel, Field, field_validator
from datetime import date
from typing import Optional

class ObservationRecord(BaseModel):
    date:        date
    location_id: str   = Field(min_length=1, max_length=64)
    new_cases:   int   = Field(ge=0, le=50000)
    source:      Optional[str] = None

    @field_validator("new_cases")
    @classmethod
    def not_sentinel(cls, v):
        sentinels = {99999, 9999, -1}
        if v in sentinels:
            raise ValueError(f"Sentinel value {v} not allowed")
        return v
```

Any record that fails validation returns HTTP 422 with a clear error message — it never touches the database.

## Validation as a service feature

Data quality reporting is itself a product feature clients value. Include in your dashboard:

- **Completeness score**: % of expected daily records received
- **Anomaly log**: timestamped record of all flagged values and how they were handled
- **Imputation transparency**: where were gaps filled, and how

This turns a defensive engineering practice into a visible trust signal.

## References

::: {#refs}
:::
