---
title: "Real-Time Simulation Progress Streaming with WebSockets"
subtitle: "Long-running EnKF jobs need live feedback. WebSocket push turns a waiting spinner into an informative progress stream. Skill 13 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, WebSockets, real-time, FastAPI, R, API]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The problem with long-running simulations

A standard HTTP request has a request–response cycle: client asks, server computes, server replies. If the computation takes 30 seconds, the client waits with a blank spinner. If it takes 5 minutes, most HTTP clients time out.

For epidemic simulations — parameter sweeps, scenario ensembles, full Bayesian MCMC — computation time is measured in minutes, not milliseconds. Clients need three things:

1. **Progress indication**: "On iteration 45 of 200"
2. **Partial results**: display preliminary estimates as they arrive
3. **Graceful cancellation**: stop the run if the client navigates away

**WebSockets** [@fette2011] solve this. Unlike HTTP, a WebSocket connection stays open after the initial handshake. The server can push messages to the client at any time — progress updates, intermediate results, error notifications.

## WebSocket vs Server-Sent Events

| | WebSocket | Server-Sent Events (SSE) |
|--|-----------|--------------------------|
| Direction | Bidirectional | Server → client only |
| Protocol | ws:// or wss:// | HTTP |
| Browser support | Excellent | Excellent |
| Reconnect | Manual | Automatic |
| **Best for** | Interactive, two-way | Progress streams |

For simulation progress (server → client only), SSE is simpler. For interactive dashboards where the client sends slider updates, WebSocket is needed. We cover WebSocket as it is more general [@fastapi_ws].

## FastAPI WebSocket implementation

```python
# server.py
from fastapi import FastAPI, WebSocket
import asyncio, json

app = FastAPI()

async def run_enkf_with_progress(obs: list, ws: WebSocket):
    """Run EnKF and send progress updates over WebSocket."""
    n_steps  = len(obs)
    beta_est = 0.30
    I_est    = obs[0] if obs else 10

    for t, y_t in enumerate(obs):
        # Simulate one EnKF update step
        await asyncio.sleep(0.05)   # represents real computation
        innov   = y_t - I_est
        K       = 0.3
        I_est   = max(0, I_est + K * innov)
        beta_est = max(0.05, beta_est + 0.001 * innov / max(y_t, 1))

        # Push progress to client
        await ws.send_json({
            "type":     "progress",
            "step":     t + 1,
            "total":    n_steps,
            "pct":      round((t + 1) / n_steps * 100, 1),
            "I_est":    round(I_est, 1),
            "beta_est": round(beta_est, 4)
        })

    # Final result
    await ws.send_json({
        "type":     "complete",
        "I_final":  round(I_est, 1),
        "beta_final": round(beta_est, 4)
    })

@app.websocket("/ws/enkf")
async def enkf_websocket(websocket: WebSocket):
    await websocket.accept()
    data = await websocket.receive_json()
    obs  = data.get("obs", [])
    await run_enkf_with_progress(obs, websocket)
    await websocket.close()
```

## JavaScript client (browser)

```javascript
// dashboard.js
const ws = new WebSocket("wss://api.your-domain.com/ws/enkf");

ws.onopen = () => {
  ws.send(JSON.stringify({ obs: dailyCaseCounts }));
};

ws.onmessage = (event) => {
  const msg = JSON.parse(event.data);
  if (msg.type === "progress") {
    updateProgressBar(msg.pct);
    updateChart(msg.step, msg.I_est, msg.beta_est);
  } else if (msg.type === "complete") {
    showFinalResult(msg.I_final, msg.beta_final);
    ws.close();
  }
};

ws.onerror = (err) => console.error("WebSocket error:", err);
```

## Simulating progress streaming in R

We simulate the same progress stream in R to illustrate the message structure and visualise the real-time update pattern:

```{r}
#| label: progress-simulation

set.seed(42)
n_steps <- 60
obs     <- rpois(n_steps, lambda = c(seq(10, 150, length.out = 30),
                                      seq(150, 30, length.out = 30)))

# Simulate EnKF progress messages
simulate_progress <- function(obs) {
  beta_est <- 0.30; I_est <- obs[1]
  messages <- vector("list", length(obs))
  for (t in seq_along(obs)) {
    innov    <- obs[t] - I_est
    I_est    <- max(0, I_est + 0.3 * innov)
    beta_est <- max(0.05, min(0.8, beta_est + 0.001 * innov / max(obs[t], 1)))
    messages[[t]] <- list(
      step     = t, total = length(obs),
      pct      = round(t / length(obs) * 100, 1),
      I_est    = round(I_est, 1),
      beta_est = round(beta_est, 4)
    )
  }
  do.call(rbind, lapply(messages, as.data.frame))
}

progress_df <- simulate_progress(obs)
head(progress_df, 3)
```

```{r}
#| label: progress-viz
#| fig-cap: "Simulated WebSocket progress stream: top panel shows EnKF state estimate updating in real time; bottom panel shows transmission rate β converging. Each row is one WebSocket message pushed to the client."

library(ggplot2)
library(tidyr)

progress_df$obs <- obs

p_data <- pivot_longer(progress_df[, c("step", "I_est", "obs")],
                        -step, names_to = "series")

ggplot(p_data, aes(x = step, y = value, colour = series)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c(I_est = "steelblue", obs = "grey50"),
                      labels = c(I_est = "EnKF estimate", obs = "Observed"),
                      name = NULL) +
  labs(x = "Step (WebSocket message number)", y = "Cases",
       title = "Real-time progress stream: state estimate vs observations") +
  theme_minimal(base_size = 13)
```

## Reconnection and heartbeats

WebSocket connections drop — mobile networks, server restarts, load balancer timeouts. Handle this gracefully:

```python
# Server: send a heartbeat every 30 seconds to keep the connection alive
async def run_with_heartbeat(ws: WebSocket):
    heartbeat_task = asyncio.create_task(send_heartbeat(ws))
    try:
        await run_enkf_with_progress(obs, ws)
    finally:
        heartbeat_task.cancel()

async def send_heartbeat(ws: WebSocket):
    while True:
        await asyncio.sleep(30)
        await ws.send_json({"type": "heartbeat"})
```

```javascript
// Client: reconnect on disconnect
ws.onclose = () => {
  console.log("Connection lost, reconnecting in 3s...");
  setTimeout(() => connectWebSocket(), 3000);
};
```

## When to use WebSockets vs polling

Use WebSockets when: computation takes > 10 seconds, you want sub-second latency updates, or you need bidirectional communication.

Use polling (repeated HTTP GET every 5s) when: simplicity matters more than latency, or your infrastructure does not support persistent connections (some serverless platforms).

For epidemic models with 30–120 second run times, WebSockets are the right choice.

## References

::: {#refs}
:::
