---
title: "Self-Attention and Transformers: The Architecture Behind LLMs"
subtitle: "Attention is a learned weighted average over past time steps. Build it from scratch in R by solving a concrete prediction task, then see how the full transformer scales the idea."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [machine learning, transformers, attention, deep learning, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why transformers matter for epidemic intelligence

In 2017, Vaswani et al. [@vaswani2017] published "Attention Is All You Need" and rendered recurrent networks obsolete for sequence modeling. The transformer architecture they introduced is now the foundation of every large language model — GPT-4, Claude, Gemini — and of the time-series forecasting models emerging in computational epidemiology, including Temporal Fusion Transformer [@lim2021] and TimeGPT.

Understanding transformers has two practical payoffs for a digital twin platform:

1. **Time-series forecasting**: attention-based models can learn which past epidemic weeks are most informative for the next-week forecast — something fixed moving averages cannot do.
2. **LLM integration**: LLMs attend to every part of your prompt when generating each output token. Understanding this mechanism makes you a better prompt engineer.

## The prediction problem

An epidemic time series has temporal structure: next week's case count depends on recent weeks, but not all weeks equally. The key question is **which past weeks matter most**, and by how much.

The simplest forecaster is a **moving average** (MA): predict next week as the mean of the last $k$ weeks. This treats all lags equally. Attention learns a *weighted* average where the weights are optimised to minimise prediction error.

To make this concrete, we generate a 30-week epidemic curve and set up a **one-step-ahead forecasting task**: given the last $k = 5$ weeks, predict the next week's case count.

```{r}
#| label: attn-data

set.seed(42)

# 30-week epidemic curve (rise to peak at week 15, then decline)
t_seq <- 1:30
cases <- round(200 * exp(-((t_seq - 15)^2) / 20) + rnorm(30, 0, 8))
cases <- pmax(1, cases)

k <- 5   # look-back window (how many past weeks to use)

# Sliding-window dataset: each row = one prediction step
n_pred <- length(cases) - k
X_mat  <- do.call(rbind, lapply(
  (k + 1):length(cases),
  function(i) cases[(i - k):(i - 1)]
))                                      # (n_pred, k) — each row is a window
y_vec  <- cases[(k + 1):length(cases)] # (n_pred,)   — next-week targets

cat("Forecasting steps:", n_pred, "\n")
cat("Observed range:   ", min(cases), "–", max(cases), "cases/week\n")
```

## Attention as a learned weighted average

A moving average computes $\hat{y}_t = \frac{1}{k} \sum_{j=1}^{k} x_{t-j}$. Attention generalises this to a *softmax-weighted* sum:

$$\hat{y}_t = \sum_{j=1}^{k} w_j \, x_{t-j}, \quad \mathbf{w} = \text{softmax}(\boldsymbol{\alpha})$$

where $\boldsymbol{\alpha} \in \mathbb{R}^k$ are **unnormalized attention logits** — the parameters we learn. The softmax ensures weights are non-negative and sum to 1. We find $\boldsymbol{\alpha}$ by minimising mean squared prediction error with gradient descent.

This is the core idea behind every transformer: learn *which positions to attend to* rather than treating all positions equally.

```{r}
#| label: attn-train

softmax <- function(alpha) {
  e <- exp(alpha - max(alpha))   # numerically stable
  e / sum(e)
}

# Loss: MSE between attention-weighted prediction and true next-week cases
loss_fn <- function(alpha) {
  w    <- softmax(alpha)
  pred <- as.vector(X_mat %*% w)
  mean((pred - y_vec)^2)
}

# Numerical gradient (central differences)
num_grad <- function(f, x, eps = 1e-6) {
  vapply(seq_along(x), function(j) {
    x1 <- x2 <- x
    x1[j] <- x1[j] + eps
    x2[j] <- x2[j] - eps
    (f(x1) - f(x2)) / (2 * eps)
  }, numeric(1))
}

# Gradient descent on alpha
alpha <- rep(0, k)   # start: uniform weights (= moving average)
lr    <- 0.5
for (iter in 1:2000) {
  alpha <- alpha - lr * num_grad(loss_fn, alpha)
}

w_learned <- softmax(alpha)

# Predictions
ma_pred   <- rowMeans(X_mat)                 # baseline: uniform MA
attn_pred <- as.vector(X_mat %*% w_learned)  # attention

# Evaluation
rmse_ma   <- sqrt(mean((ma_pred   - y_vec)^2))
rmse_attn <- sqrt(mean((attn_pred - y_vec)^2))

cat("Learned attention weights (lag 1 = last week, lag 5 = 5 weeks ago):\n")
cat(paste0("  lag ", seq_len(k), ": ", round(w_learned, 3)), sep = "\n")
cat("\nMoving average RMSE:", round(rmse_ma,   1), "\n")
cat("Attention RMSE:     ", round(rmse_attn, 1), "\n")
```

## What the learned weights reveal

```{r}
#| label: attn-plot
#| fig-cap: "Top: one-step-ahead forecasts. Attention (blue) follows the epidemic curve more closely than the moving average (orange) because it up-weights the most recent lag. Bottom: learned attention weights by lag. Lag 1 (last week) receives the highest weight — consistent with short-term autocorrelation in epidemic incidence. The weights decay with lag, a pattern a fixed MA cannot represent."

library(ggplot2)

week_idx <- (k + 1):length(cases)

df_pred <- data.frame(
  week    = rep(week_idx, 3),
  cases   = c(y_vec, ma_pred, attn_pred),
  model   = rep(c("Observed", "Moving average", "Attention (learned)"),
                each = n_pred)
)
df_pred$model <- factor(df_pred$model,
  levels = c("Observed", "Moving average", "Attention (learned)"))

p_pred <- ggplot(df_pred, aes(x = week, y = cases,
                               colour = model, linetype = model)) +
  geom_line(linewidth = 0.9) +
  geom_point(data = subset(df_pred, model == "Observed"),
             size = 1.8) +
  scale_colour_manual(values = c("Observed"           = "black",
                                  "Moving average"     = "orange",
                                  "Attention (learned)"= "steelblue"),
                      name = NULL) +
  scale_linetype_manual(values = c("Observed"           = "solid",
                                    "Moving average"     = "dashed",
                                    "Attention (learned)"= "solid"),
                        name = NULL) +
  labs(x = "Week", y = "Cases",
       title = "A: One-step-ahead forecast comparison") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

lag_labels <- paste0("lag ", seq_len(k),
                     c(" (last wk)", rep("", k - 1)))
df_weights <- data.frame(
  lag    = factor(lag_labels, levels = lag_labels),
  weight = w_learned
)

p_weights <- ggplot(df_weights, aes(x = lag, y = weight)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 1 / k, linetype = "dashed", colour = "firebrick") +
  annotate("text", x = 1.5, y = 1 / k + 0.01,
           label = "uniform (MA)", colour = "firebrick", size = 3.5) +
  labs(x = "Lag", y = "Attention weight",
       title = "B: Learned attention weights") +
  theme_minimal(base_size = 12)

gridExtra_available <- requireNamespace("gridExtra", quietly = TRUE)
if (gridExtra_available) {
  gridExtra::grid.arrange(p_pred, p_weights, ncol = 2)
} else {
  print(p_pred)
  print(p_weights)
}
```

The dashed red line in panel B is the uniform weight $1/k$ that the moving average uses. The learned attention concentrates weight on recent lags, which is the correct inductive bias for epidemic incidence: last week explains most of next week's count.

## From weighted average to full self-attention

The predictor above has one limitation: the weights $\mathbf{w}$ are *fixed* once trained — they do not adapt to the current value of the sequence. If the epidemic is accelerating versus plateauing, the optimal lag weights differ.

The transformer's self-attention solves this by making the weights **content-dependent**. Each position produces a query $\mathbf{q}$ ("what context am I looking for?") and keys $\mathbf{k}_j$ ("what does position $j$ offer?"). The attention weight from position $i$ to position $j$ is:

$$w_{ij} = \frac{\exp(\mathbf{q}_i \cdot \mathbf{k}_j / \sqrt{d_k})}{\sum_j \exp(\mathbf{q}_i \cdot \mathbf{k}_j / \sqrt{d_k})}$$

When the epidemic is growing steeply, the query for the current position will be similar to keys of recent high-growth periods, concentrating weight there. When the epidemic plateaus, the query changes and the attention pattern shifts. This is impossible with a fixed $\mathbf{w}$.

The full transformer [@vaswani2017] extends further:

| Component | Purpose |
|-----------|---------|
| **Multi-head attention** | Run $h$ attention patterns in parallel; different heads may learn short-range vs long-range dependencies |
| **Causal masking** | For forecasting, mask future positions so position $t$ only attends to positions $\leq t$ |
| **Feed-forward sublayer** | Per-position MLP applied after attention (nonlinear feature transform) |
| **Layer normalisation + residuals** | Stable training through depth |
| **Positional encoding** | Inject position information since attention itself is permutation-invariant |

Stack 6–96 such blocks, pre-train on token prediction at scale, and the result is a model that reasons across arbitrarily long contexts — the same mechanism that allows Claude or GPT to follow a 10,000-word prompt consistently.

## Practical takeaway

- **Attention = learned weighted average over context**, adapted to content at each position.
- For epidemic forecasting, recent lags deserve higher attention — a fixed MA cannot express this, but learned attention can.
- You do not need to implement a transformer from scratch in production. Use `neuralforecast` (Python) or Nixtla's TimeGPT API for transformer-based epidemic forecasting.
- Understanding the mechanism tells you *why* these models work and *what their attention patterns mean* when you inspect them.

## References

::: {#refs}
:::
