---
title: "Self-Attention and Transformers: The Architecture Behind LLMs"
subtitle: "GPT, BERT, and every modern language model are built on one idea: letting every position in a sequence attend to every other. Implement scaled dot-product attention in R."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [machine learning, transformers, attention, deep learning, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why transformers matter for epidemic intelligence

In 2017, Vaswani et al. [@vaswani2017] published "Attention Is All You Need" and rendered recurrent networks obsolete for sequence modeling. The transformer architecture they introduced is now the foundation of every large language model — GPT-4, Claude, Gemini, and the text-to-forecast models emerging in computational epidemiology.

Understanding transformers has two practical payoffs for a digital twin platform:

1. **LLM integration**: you will embed LLM API calls into your reporting pipeline (Skill 5 in this series). Understanding the mechanism helps you prompt more effectively and interpret model behaviour.
2. **Time-series forecasting**: transformer-based models like Temporal Fusion Transformer [@lim2021] now match or beat classical methods on epidemic time series. Understanding attention tells you *why*.

## The core problem with sequences

An epidemic time series has temporal dependencies: this week's case count depends on last week's. RNNs handle this by maintaining a hidden state, but they pass information through a bottleneck — each step must compress the entire past into a fixed-size vector. Long-range dependencies (a seasonal pattern 52 weeks ago informing today's forecast) get lost.

Self-attention sidesteps the bottleneck: every time step attends directly to every other time step. The weight of each connection is learned.

## Scaled dot-product attention

For an input sequence of $n$ tokens, each represented as a $d_k$-dimensional vector, we compute three matrices:

- $Q = X W_Q$ (queries: "what am I looking for?")
- $K = X W_K$ (keys: "what do I offer?")
- $V = X W_V$ (values: "what do I contribute if selected?")

The attention output is:

$$\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{Q K^\top}{\sqrt{d_k}}\right) V$$

The $Q K^\top$ matrix is an $n \times n$ similarity matrix — each row is the query for one position, each column is the key for another. After softmax, each row sums to 1 and acts as a weighted average over the values.

## Implementing attention in R

We apply self-attention to a 12-week epidemic time series to see which weeks attend to which.

```{r}
#| label: attn-setup

set.seed(42)

# Twelve weeks of epidemic cases (rise and fall)
cases  <- c(5, 12, 28, 64, 120, 180, 200, 165, 110, 60, 25, 10)
n_seq  <- length(cases)
d_model <- 8   # embedding dimension

# Simple linear embedding: scale the scalar into d_model dims
x_norm <- cases / max(cases)
set.seed(7)
W_emb <- matrix(rnorm(d_model, 0, 0.5), 1, d_model)
X_emb <- x_norm %*% W_emb    # broadcast: (n_seq, d_model)

# Positional encoding (sinusoidal)
pos_enc <- matrix(0, n_seq, d_model)
for (pos in seq_len(n_seq)) {
  for (i in seq(1, d_model, 2)) {
    pos_enc[pos, i]   <- sin(pos / 10000^((i - 1) / d_model))
    if (i + 1 <= d_model)
      pos_enc[pos, i + 1] <- cos(pos / 10000^((i - 1) / d_model))
  }
}
X <- X_emb + pos_enc   # (n_seq, d_model) with position information
```

```{r}
#| label: attn-compute

# Q, K, V projection matrices
d_k <- d_model
set.seed(3)
W_Q <- matrix(rnorm(d_model * d_k, 0, 0.3), d_model, d_k)
W_K <- matrix(rnorm(d_model * d_k, 0, 0.3), d_model, d_k)
W_V <- matrix(rnorm(d_model * d_k, 0, 0.3), d_model, d_k)

Q <- X %*% W_Q   # (n_seq, d_k)
K <- X %*% W_K
V <- X %*% W_V

# Numerically stable row-wise softmax
row_softmax <- function(M) {
  t(apply(M, 1, function(row) {
    e <- exp(row - max(row))
    e / sum(e)
  }))
}

# Scaled dot-product attention
scores       <- Q %*% t(K) / sqrt(d_k)   # (n_seq, n_seq)
attn_weights <- row_softmax(scores)       # (n_seq, n_seq)
attn_output  <- attn_weights %*% V        # (n_seq, d_k)

cat("Attention weight matrix dimension:", dim(attn_weights), "\n")
cat("Each row sums to:", round(rowSums(attn_weights)[1:3], 6), "...\n")
```

```{r}
#| label: attn-heatmap
#| fig-cap: "Self-attention weight matrix for a 12-week epidemic time series. Each cell (row i, column j) shows how much week i attends to week j when computing its contextual representation. The diagonal dominates for many positions (each week attends to itself), but peak weeks (5–7) attend broadly to their neighbours, capturing the plateau structure."

library(ggplot2)

week_labels <- paste0("W", seq_len(n_seq))

attn_df <- expand.grid(query = week_labels, key = week_labels,
                        stringsAsFactors = FALSE)
attn_df$query <- factor(attn_df$query, levels = week_labels)
attn_df$key   <- factor(attn_df$key,   levels = week_labels)
attn_df$weight <- as.vector(t(attn_weights))   # row = query

ggplot(attn_df, aes(x = key, y = query, fill = weight)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "white", high = "steelblue",
                      name = "Attention\nweight") +
  scale_y_discrete(limits = rev(week_labels)) +
  labs(x = "Key (attends to)", y = "Query (from position)",
       title = "Self-attention weights: 12-week epidemic series") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Multi-head attention

A single attention head learns one type of relationship. Multi-head attention runs $h$ heads in parallel, each with its own $W_Q, W_K, W_V$, then concatenates and projects the outputs:

$$\text{MultiHead}(Q,K,V) = \text{Concat}(\text{head}_1, \ldots, \text{head}_h) W^O$$

In practice, $d_k = d_{\text{model}} / h$ so total compute stays constant. Different heads learn different relationships — one may capture local context, another seasonal patterns.

## The full transformer block

One transformer layer wraps multi-head attention in a residual connection, followed by a feedforward sublayer:

```
x → MultiHeadAttention(x) → Add & LayerNorm → FFN → Add & LayerNorm → output
```

Stack 12–96 such layers (GPT-4 uses 96) and pre-train on token prediction across trillions of words, and you get a model that can reason, translate, and generate coherent epidemic summaries.

## What this means for your platform

You do not need to train a transformer from scratch — foundation models already exist. What matters is understanding:

- **Why attention captures long-range patterns** better than moving averages or RNNs
- **What the attention weights mean** when a model focuses heavily on certain weeks in a forecast
- **How to structure inputs** for transformer-based forecasting APIs (e.g., Nixtla's TimeGPT)

The attention mechanism is also the reason LLMs can follow long, structured prompts reliably — the model is literally attending to every part of your prompt when generating each output token.

## References

::: {#refs}
:::
