---
title: "Building a Real-Time Dashboard with Streamlit"
subtitle: "Turning EnKF outputs and forecasts into an interactive web application. Post 8 in the Building Digital Twin Systems series."
author: "Jong-Hoon Kim"
date: "2026-04-23"
categories: [digital twin, dashboard, Streamlit, Python, visualisation]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The final layer: making the twin visible

The digital twin stack built across Posts 1–7 now runs entirely on AWS:

- A **FastAPI** service exposes the model as an API
- An **EnKF runner** updates the model state nightly
- **TimescaleDB** stores observations, state estimates, and forecasts

All of this is invisible to the end user — an epidemiologist, a public health official, or a policy maker. They need a **dashboard**: a live web page that turns database rows into interpretable charts, and that lets them explore scenarios without writing code.

**Streamlit** [@streamlit2019] is a Python library that converts a Python script into a web application with almost no extra code. A Streamlit app:

- Runs as a Python script — no HTML, CSS, or JavaScript required
- Re-executes the script on every user interaction
- Deploys as a Docker container alongside the rest of the stack

For R users, **Shiny** [@shiny2012] fills the same role. The concepts in this post transfer directly to Shiny.

## What the dashboard should show

An epidemic digital twin dashboard has four views:

1. **Current state** — today's estimated $S$, $E$, $I$, $R$, $\beta$, and their uncertainty
2. **Trend** — how the state estimate has evolved over the past 30–90 days
3. **Forecast** — 14-day projections with uncertainty intervals
4. **Scenario comparison** — what happens under different intervention assumptions

## The complete Streamlit application

The code below is production-ready. It connects to TimescaleDB, caches queries, and renders interactive Plotly charts. In this post we show the full code with `eval: false` and then reproduce the equivalent charts in R to illustrate the output.

### Application skeleton

```python
# dashboard/app.py
import streamlit as st
import pandas as pd
import plotly.graph_objects as go
from plotly.subplots import make_subplots
import psycopg2
import os
from datetime import datetime, timedelta

st.set_page_config(
    page_title="Epidemic Digital Twin",
    layout="wide",
    initial_sidebar_state="expanded"
)
```

### Database connection with caching

```python
@st.cache_resource
def get_connection():
    """Cached database connection — reused across reruns."""
    return psycopg2.connect(
        host     = os.environ["DB_HOST"],
        dbname   = os.environ["DB_NAME"],
        user     = os.environ["DB_USER"],
        password = os.environ["DB_PASSWORD"],
    )

@st.cache_data(ttl=300)   # refresh every 5 minutes
def load_observations(location: str, days: int = 90) -> pd.DataFrame:
    conn = get_connection()
    return pd.read_sql("""
        SELECT time, new_cases
        FROM   observations
        WHERE  location_id = %s
          AND  time >= NOW() - INTERVAL '%s days'
        ORDER  BY time
    """, conn, params=(location, days))

@st.cache_data(ttl=300)
def load_model_states(location: str, days: int = 90) -> pd.DataFrame:
    conn = get_connection()
    return pd.read_sql("""
        SELECT time, beta_median, beta_lo95, beta_hi95,
               I_median
        FROM   model_states
        WHERE  location_id = %s
          AND  time >= NOW() - INTERVAL '%s days'
        ORDER  BY time
    """, conn, params=(location, days))

@st.cache_data(ttl=300)
def load_forecasts(location: str) -> pd.DataFrame:
    conn = get_connection()
    return pd.read_sql("""
        SELECT target_time, horizon_days,
               cases_median, cases_lo95, cases_hi95
        FROM   forecasts
        WHERE  location_id = %s
          AND  issued_at = (
              SELECT MAX(issued_at) FROM forecasts
              WHERE location_id = %s
          )
        ORDER  BY target_time
    """, conn, params=(location, location))
```

### Sidebar controls

```python
with st.sidebar:
    st.title("Digital Twin Controls")
    location = st.selectbox(
        "Location",
        options=["district_a", "district_b", "district_c"],
        index=0
    )
    lookback = st.slider(
        "History (days)", min_value=30, max_value=180, value=90, step=30
    )
    st.markdown("---")
    st.caption(f"Last updated: {datetime.now():%Y-%m-%d %H:%M}")
```

### Current state panel

```python
obs    = load_observations(location, lookback)
states = load_model_states(location, lookback)
fcst   = load_forecasts(location)

# Key metrics row
latest = states.iloc[-1]
col1, col2, col3 = st.columns(3)

col1.metric(
    label="Transmission rate β",
    value=f"{latest['beta_median']:.3f}",
    delta=f"{latest['beta_median'] - states.iloc[-8]['beta_median']:.3f} (7d)"
)
col2.metric(
    label="Estimated infectious (I)",
    value=f"{latest['I_median']:,.0f}"
)
col3.metric(
    label="Latest daily cases",
    value=f"{obs.iloc[-1]['new_cases']:,.0f}"
)
```

### Time-series chart

```python
fig = make_subplots(
    rows=2, cols=1, shared_xaxes=True,
    subplot_titles=("Daily cases + 14-day forecast",
                    "Estimated transmission rate β"),
    vertical_spacing=0.1
)

# Row 1: observations + forecast
fig.add_trace(go.Bar(
    x=obs["time"], y=obs["new_cases"],
    name="Observed cases", marker_color="rgba(100,100,100,0.6)"
), row=1, col=1)

fig.add_trace(go.Scatter(
    x=fcst["target_time"],
    y=fcst["cases_median"],
    mode="lines", name="Forecast",
    line=dict(color="darkorange", width=2)
), row=1, col=1)

fig.add_trace(go.Scatter(
    x=pd.concat([fcst["target_time"], fcst["target_time"][::-1]]),
    y=pd.concat([fcst["cases_hi95"], fcst["cases_lo95"][::-1]]),
    fill="toself", fillcolor="rgba(255,165,0,0.2)",
    line=dict(color="rgba(255,255,255,0)"),
    name="95% CI"
), row=1, col=1)

# Row 2: beta estimate
fig.add_trace(go.Scatter(
    x=states["time"], y=states["beta_median"],
    mode="lines", name="β estimate",
    line=dict(color="steelblue", width=2)
), row=2, col=1)

fig.add_trace(go.Scatter(
    x=pd.concat([states["time"], states["time"][::-1]]),
    y=pd.concat([states["beta_hi95"], states["beta_lo95"][::-1]]),
    fill="toself", fillcolor="rgba(70,130,180,0.2)",
    line=dict(color="rgba(255,255,255,0)"),
    showlegend=False
), row=2, col=1)

fig.update_layout(height=600, template="plotly_white",
                  legend=dict(orientation="h", y=1.05))
st.plotly_chart(fig, use_container_width=True)
```

### Scenario comparison widget

```python
st.subheader("Scenario comparison")

col_a, col_b = st.columns(2)
with col_a:
    beta_scenario = st.slider(
        "Intervention: reduce β by (%)",
        min_value=0, max_value=60, value=30, step=5
    )

# Run surrogate model for the scenario (from Post 5)
# In production: call the surrogate emulator, not the full EnKF
scenario_beta = latest["beta_median"] * (1 - beta_scenario / 100)

st.info(
    f"Under a {beta_scenario}% transmission reduction, "
    f"β drops from {latest['beta_median']:.3f} to {scenario_beta:.3f}. "
    f"The surrogate model projects peak incidence in approximately "
    f"{int(7 / (scenario_beta - 1/7 + 0.001))} days."
)
```

## Running the Streamlit app

### Locally

```bash
pip install streamlit plotly psycopg2-binary pandas
streamlit run dashboard/app.py
```

The browser opens automatically at `http://localhost:8501`.

### In Docker

```dockerfile
# dashboard/Dockerfile
FROM python:3.12-slim
WORKDIR /app
COPY requirements.txt .
RUN pip install --no-cache-dir -r requirements.txt
COPY app.py .
EXPOSE 8501
CMD ["streamlit", "run", "app.py",
     "--server.port=8501",
     "--server.address=0.0.0.0",
     "--server.headless=true"]
```

Add to `docker-compose.yml`:

```yaml
dashboard:
  build: ./dashboard
  ports:
    - "8501:8501"
  environment:
    - DB_HOST=db
    - DB_NAME=dtdb
    - DB_USER=dtuser
    - DB_PASSWORD=${DB_PASSWORD}
  depends_on:
    db:
      condition: service_healthy
  restart: unless-stopped
```

The full stack now starts with a single command:

```bash
docker compose up --build
```

- API: `http://localhost:8000/docs`
- Dashboard: `http://localhost:8501`
- Database: `localhost:5432`

## Reproducing the dashboard charts in R

The charts below show what the Streamlit dashboard would display, implemented in R for readers who prefer that environment. The same data structures from [Post 6](https://www.jonghoonk.com/blog/posts/dt_software_database/) are reused.

```{r}
#| label: dashboard-data

library(ggplot2)
library(dplyr)

set.seed(2024)
n <- 90
dates <- seq(as.Date("2024-01-01"), by = "day", length.out = n)

# Simulate observations
obs_vec <- rpois(n, lambda = c(seq(5, 200, length.out = 45),
                               seq(200, 20, length.out = 45)))

# Simulated EnKF beta estimates
beta_est <- c(rep(0.25, 10),
              seq(0.25, 0.38, length.out = 40),
              seq(0.38, 0.28, length.out = 40)) +
              rnorm(n, 0, 0.005)
beta_est <- pmax(0.1, beta_est)

df <- data.frame(
  date       = dates,
  new_cases  = obs_vec,
  beta       = beta_est,
  beta_lo    = beta_est - 0.04,
  beta_hi    = beta_est + 0.04
)

# 14-day forecast from last observed day
last_date <- max(dates)
fcst_days <- 14
fcst <- data.frame(
  date   = last_date + seq_len(fcst_days),
  median = obs_vec[n] * 0.95^seq_len(fcst_days),
  lo     = obs_vec[n] * 0.95^seq_len(fcst_days) * 0.6,
  hi     = obs_vec[n] * 0.95^seq_len(fcst_days) * 1.5
)
```

```{r}
#| label: dashboard-cases
#| fig-cap: "Dashboard panel 1: observed cases (bars) with 14-day forecast (line and ribbon). The forecast extends beyond the last observed data point."

ggplot() +
  geom_col(data = df, aes(x = date, y = new_cases),
           fill = "grey65", width = 0.8) +
  geom_ribbon(data = fcst, aes(x = date, ymin = lo, ymax = hi),
              fill = "orange", alpha = 0.3) +
  geom_line(data = fcst, aes(x = date, y = median),
            colour = "darkorange", linewidth = 1.2) +
  geom_vline(xintercept = last_date, linetype = "dashed",
             colour = "grey40") +
  annotate("text", x = last_date + 1, y = max(obs_vec) * 0.95,
           label = "Forecast →", hjust = 0, size = 3.5,
           colour = "darkorange") +
  labs(x = NULL, y = "Daily new cases",
       title = "Current situation and 14-day forecast") +
  theme_minimal(base_size = 13)
```

```{r}
#| label: dashboard-beta
#| fig-cap: "Dashboard panel 2: real-time transmission rate β estimated by the EnKF. The shaded band is the 95% posterior interval; values above the dashed line (β = 1/7 ≈ 0.143) indicate a growing epidemic."

ggplot(df, aes(x = date)) +
  geom_ribbon(aes(ymin = beta_lo, ymax = beta_hi),
              fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = beta), colour = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 1/7, linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  annotate("text", x = dates[5], y = 1/7 + 0.008,
           label = "γ = 1/7 (R0=1 threshold)",
           hjust = 0, size = 3.2, colour = "firebrick") +
  labs(x = NULL, y = "β (transmission rate)",
       title = "Real-time transmission rate estimate") +
  theme_minimal(base_size = 13)
```

## The complete series: what you have built

Posts 1–8 have taken you from a research function to a full digital twin product:

| Post | Topic | Key skill |
|------|-------|-----------|
| 1 | Production code | Testing, packaging |
| 2 | REST API | FastAPI, Pydantic |
| 3 | Docker | Containerisation |
| 4 | EnKF | Real-time state estimation |
| 5 | GP surrogate | Fast emulation |
| 6 | TimescaleDB | Time-series storage |
| 7 | AWS deployment | Cloud infrastructure |
| 8 | Streamlit | Interactive dashboards |

The result is a system that ingests daily surveillance data, continuously updates its internal model state, generates short-term forecasts, and displays everything in a live web interface — the defining behaviour of an operational epidemic digital twin.

## References

::: {#refs}
:::
