---
title: "Multi-Tenant Architecture: Serving Multiple Health Departments on One Platform"
subtitle: "Row-level security, schema isolation, and cost allocation across clients who must not see each other's data. Skill 11 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, multi-tenancy, database, architecture, software engineering]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The scaling inflection point

Your first client is running on a dedicated instance — one database, one API container. You sign a second client. Do you:

**Option A**: Clone the entire stack — two separate databases, two API deployments, two sets of infrastructure.

**Option B**: Add a `tenant_id` column to every table and run everything on shared infrastructure.

Option A is simple but costs 2× at two clients, 10× at ten. Option B is a **multi-tenant** architecture [@guo2007; @fowler2002]: shared infrastructure, logically isolated data. It is the model used by every successful SaaS business.

## Three isolation models

```{r}
#| label: isolation-models
#| fig-cap: "Three multi-tenancy isolation models on the security vs. cost spectrum. Most digital twin businesses start at Row-Level Security and move to Schema Isolation if compliance requires it."

library(ggplot2)

df <- data.frame(
  model      = c("Shared table\n+ RLS", "Schema per\ntenant",
                 "Database per\ntenant"),
  cost       = c(1, 2, 4),
  isolation  = c(2, 3.5, 5),
  label_cost = c("Lowest cost\nHighest density", "Medium cost",
                 "Highest cost\nFull isolation")
)

ggplot(df, aes(x = cost, y = isolation)) +
  geom_point(size = 6, colour = "steelblue") +
  geom_text(aes(label = model), vjust = -1, size = 3.5) +
  geom_text(aes(label = label_cost), vjust = 2.5, size = 2.8,
            colour = "grey50") +
  scale_x_continuous(name = "Infrastructure cost (relative)",
                     limits = c(0.5, 5)) +
  scale_y_continuous(name = "Data isolation strength",
                     limits = c(1, 6)) +
  labs(title = "Multi-tenancy isolation models") +
  theme_minimal(base_size = 13)
```

## Row-Level Security in PostgreSQL

The recommended starting point: add `tenant_id` to every table, then use PostgreSQL's **Row-Level Security (RLS)** to enforce that each database session can only see its own tenant's rows.

```sql
-- 1. Add tenant_id to existing tables
ALTER TABLE observations  ADD COLUMN tenant_id TEXT NOT NULL;
ALTER TABLE model_states  ADD COLUMN tenant_id TEXT NOT NULL;
ALTER TABLE forecasts     ADD COLUMN tenant_id TEXT NOT NULL;

-- 2. Enable RLS on each table
ALTER TABLE observations  ENABLE ROW LEVEL SECURITY;
ALTER TABLE model_states  ENABLE ROW LEVEL SECURITY;
ALTER TABLE forecasts     ENABLE ROW LEVEL SECURITY;

-- 3. Create policy: session variable must match tenant_id
CREATE POLICY tenant_isolation ON observations
  USING (tenant_id = current_setting('app.current_tenant'));

CREATE POLICY tenant_isolation ON model_states
  USING (tenant_id = current_setting('app.current_tenant'));

-- 4. Set the tenant on connection
SET app.current_tenant = 'district_health_dept';
-- Now ALL queries on this connection see only that tenant's data
```

The API sets the tenant variable from the JWT claim before executing any query. Each database connection is locked to one tenant.

## Simulating RLS in R

```{r}
#| label: rls-simulation

# Simulate the multi-tenant data store in R
all_observations <- data.frame(
  tenant_id   = c(rep("dept_a", 5), rep("dept_b", 5), rep("dept_c", 5)),
  date        = rep(Sys.Date() - 4:0, 3),
  location_id = c(rep("loc_1", 5), rep("loc_2", 5), rep("loc_3", 5)),
  new_cases   = c(rpois(5, 45), rpois(5, 80), rpois(5, 30))
)

# Simulate the RLS filter — what each tenant sees
query_as_tenant <- function(data, tenant) {
  data[data$tenant_id == tenant, setdiff(names(data), "tenant_id")]
}

# Each department sees only its own data
cat("dept_a sees:\n"); print(query_as_tenant(all_observations, "dept_a"))
cat("\ndept_b sees:\n"); print(query_as_tenant(all_observations, "dept_b"))
```

```{r}
#| label: tenant-metrics
#| fig-cap: "Per-tenant case counts in a shared multi-tenant store. RLS ensures dept_a never sees dept_b's data, even though they share the same physical table."

library(ggplot2)
library(dplyr)

# Show what an admin (all-tenant) view looks like
all_observations |>
  ggplot(aes(x = date, y = new_cases, colour = tenant_id, group = tenant_id)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(dept_a = "steelblue",
                                  dept_b = "firebrick",
                                  dept_c = "darkgreen"),
                      name = "Tenant") +
  labs(x = NULL, y = "Daily cases",
       title = "Shared table — admin view across all tenants",
       subtitle = "Each tenant's session sees only their own row via RLS") +
  theme_minimal(base_size = 13)
```

## Tenant onboarding workflow

When a new health department signs a contract, the onboarding script does:

```r
onboard_tenant <- function(con, tenant_id, contact_email) {
  # 1. Insert tenant record
  DBI::dbExecute(con,
    "INSERT INTO tenants (tenant_id, contact_email, created_at)
     VALUES ($1, $2, NOW())",
    params = list(tenant_id, contact_email)
  )

  # 2. Create API key (hashed)
  raw_key   <- paste0(sample(c(letters, 0:9), 32, replace=TRUE), collapse="")
  key_hash  <- digest::digest(raw_key, algo = "sha256")
  DBI::dbExecute(con,
    "INSERT INTO api_keys (tenant_id, key_hash, created_at)
     VALUES ($1, $2, NOW())",
    params = list(tenant_id, key_hash)
  )

  # 3. Seed initial model state
  # ... insert baseline EnKF posterior for this tenant

  message(sprintf("Tenant '%s' onboarded. API key: %s", tenant_id, raw_key))
  raw_key   # return only once; never stored in plaintext
}
```

**The key rule**: the raw API key is shown only at creation and never stored — only its hash is in the database, like a password.

## Cost attribution

With shared infrastructure, you need to attribute costs to tenants for:

- **Usage-based billing**: charge by API call, EnKF run, or data stored
- **Resource planning**: identify which tenants drive the most compute

Add a `usage_events` table:

```sql
CREATE TABLE usage_events (
  tenant_id    TEXT        NOT NULL,
  event_type   TEXT        NOT NULL,  -- 'api_call', 'enkf_run', 'report_gen'
  occurred_at  TIMESTAMPTZ NOT NULL DEFAULT NOW(),
  duration_ms  INTEGER,
  cost_units   NUMERIC     -- whatever unit you bill in
);
SELECT create_hypertable('usage_events', 'occurred_at');
```

Monthly billing is then a simple aggregate query per tenant.

## References

::: {#refs}
:::
