---
title: "The Production Gap: From Research Code to Software Product"
subtitle: "Why your epidemic model works in R but breaks in production — and how to fix it. Post 1 in the Building Digital Twin Systems series."
author: "Jong-Hoon Kim"
date: "2026-04-23"
categories: [digital twin, software engineering, Python, R, reproducibility]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The gap nobody talks about

You have built an epidemic model that works. It fits data, recovers parameters, and produces credible forecasts. You have written it carefully in R or Python, and the results appear in a paper.

Now someone asks you to *run it for them*. They want to give it their data, get results, and pay for the service.

This is where almost every research model breaks — not because the science is wrong, but because research code is built to be run once by the person who wrote it, not to run reliably at any time for any user.

This post describes the gap between research code and production software, and provides a concrete set of habits to close it. It is the foundation for everything in this series. Wilson et al. [@wilson2014] and Sandve et al. [@sandve2013] provide comprehensive treatments; this post is a distilled, applied version focused on simulation software.

## What breaks when research code meets production

### Hard-coded paths and environments

Research code is full of paths like `/Users/yourname/Desktop/data/outbreak.csv`. The moment someone else runs it, or you run it on a cloud server, everything breaks. The same problem appears with R library paths, Python virtual environments, and system-level dependencies.

```r
# Research code — breaks immediately on any other machine
data <- read.csv("/Users/jonghoon/Desktop/POLYMOD_2017.csv")
model_output <- run_sir(data, beta = 0.4, gamma = 0.2)
write.csv(model_output, "/Users/jonghoon/results/sir_output.csv")
```

```r
# Production-ready — paths are arguments, not constants
run_model <- function(input_path, beta, gamma, output_path) {
  data <- read.csv(input_path)
  model_output <- run_sir(data, beta = beta, gamma = gamma)
  write.csv(model_output, output_path)
  invisible(model_output)
}
```

### No input validation

Research code trusts itself. Production code must distrust everything it receives. A user who sends a negative beta, a data frame with missing columns, or dates in the wrong timezone will produce silent wrong answers — or a crash with an unhelpful error message.

```r
# Fragile — crashes or silently wrong if inputs are bad
run_sir <- function(S0, I0, R0, beta, gamma, days) {
  # ... model code
}

# Robust — validates before computing
run_sir <- function(S0, I0, R0, beta, gamma, days) {
  stopifnot(
    is.numeric(beta), beta > 0, beta < 10,
    is.numeric(gamma), gamma > 0, gamma < 1,
    is.numeric(days), days > 0, days == round(days),
    S0 > 0, I0 >= 0, R0 >= 0
  )
  # ... model code
}
```

### No tests

If you change one function, how do you know nothing else broke? Research workflows rely on the scientist visually checking outputs. Production software requires automated tests that run in seconds and catch regressions before they reach users.

### State in global variables

Interactive R sessions accumulate state: objects in `.GlobalEnv`, options set with `options()`, random seeds set somewhere in the session. Code that works interactively fails when run fresh in a clean environment because it depends on something that was set five steps earlier.

```r
# Hidden state dependency — works in an interactive session, fails elsewhere
gamma <- 0.2           # set somewhere earlier in the script
forecast_sir <- function(beta, days) {
  # Uses gamma from the global environment — invisible dependency
  run_sir(beta = beta, gamma = gamma, days = days)
}

# Self-contained — all dependencies explicit
forecast_sir <- function(beta, gamma, days) {
  run_sir(beta = beta, gamma = gamma, days = days)
}
```

### No versioning of model and data together

A research model evolves. Parameters change, equations are corrected, new compartments are added. Without versioning, the model that produced last month's report is gone. Reproducibility requires that the code, data, and environment that produced any result can be recovered exactly [@sandve2013].

## The minimal production-ready SIR model in R

Here is a self-contained, validated, testable SIR model that illustrates the principles above. Nothing is hard-coded, all inputs are validated, and the function is easy to test.

```{r}
#| label: sir-model

sir_model <- function(S0, I0, R0 = 0, beta, gamma, days,
                      dt = 0.1) {
  # --- input validation ---
  stopifnot(
    is.numeric(S0), S0 > 0,
    is.numeric(I0), I0 >= 0,
    is.numeric(beta), beta > 0,
    is.numeric(gamma), gamma > 0, gamma < 1,
    is.numeric(days), days >= 1
  )

  N <- S0 + I0 + R0
  steps <- round(days / dt)
  out <- matrix(0, nrow = steps + 1, ncol = 4,
                dimnames = list(NULL, c("time", "S", "I", "R")))
  out[1, ] <- c(0, S0, I0, R0)

  S <- S0; I <- I0; R <- R0
  for (i in seq_len(steps)) {
    new_inf <- beta * S * I / N * dt
    new_rec <- gamma * I * dt
    S <- S - new_inf
    I <- I + new_inf - new_rec
    R <- R + new_rec
    out[i + 1, ] <- c(i * dt, S, I, R)
  }
  as.data.frame(out)
}

# Run it
result <- sir_model(S0 = 9900, I0 = 100, beta = 0.4,
                    gamma = 0.1, days = 60)
head(result, 4)
```

```{r}
#| label: sir-plot
#| fig-cap: "SIR model trajectories. The I compartment peaks around day 30."

library(ggplot2)
library(tidyr)

result_long <- pivot_longer(result, cols = c(S, I, R),
                            names_to = "compartment",
                            values_to = "count")

ggplot(result_long, aes(x = time, y = count,
                        colour = compartment)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c(S = "steelblue",
                                 I = "firebrick",
                                 R = "darkgreen")) +
  labs(x = "Days", y = "Individuals", colour = NULL) +
  theme_minimal(base_size = 13)
```

## Writing tests for your model

A test checks that known inputs produce known outputs. For a simulation, tests are often:

1. **Boundary tests**: R0 = 1 means no net growth; setting gamma = beta should stabilise the epidemic.
2. **Conservation tests**: S + I + R should equal N at every step.
3. **Regression tests**: a reference scenario whose output is saved and checked against future runs.

```{r}
#| label: tests

# Conservation: N must be constant
test_conservation <- function() {
  r <- sir_model(S0 = 9900, I0 = 100, beta = 0.3,
                 gamma = 0.1, days = 100)
  N <- 9900 + 100
  max_deviation <- max(abs(rowSums(r[, c("S","I","R")]) - N))
  if (max_deviation > 0.01) {
    stop(paste("Conservation violated: max deviation =",
               round(max_deviation, 4)))
  }
  message("Conservation test passed")
}

# Boundary: I(∞) → 0 for R0 < 1
test_subcritical <- function() {
  r <- sir_model(S0 = 9990, I0 = 10, beta = 0.05,
                 gamma = 0.2, days = 200)
  final_I <- tail(r$I, 1)
  if (final_I > 1) {
    stop(paste("Subcritical epidemic did not die out; I =",
               round(final_I, 2)))
  }
  message("Subcritical test passed")
}

test_conservation()
test_subcritical()
```

Both tests pass. In a real project these would live in a `tests/` directory and be run automatically every time code changes, using `testthat` in R or `pytest` in Python.

## Python packaging: structuring code to be imported

The same model in Python, structured as a proper package rather than a script:

```python
# File: sir_model/model.py

def sir_model(S0: float, I0: float, beta: float,
              gamma: float, days: int, dt: float = 0.1,
              R0: float = 0.0) -> list[dict]:
    """
    Deterministic SIR model via Euler integration.

    Parameters
    ----------
    S0, I0, R0 : initial compartment sizes
    beta       : transmission rate (per day)
    gamma      : recovery rate (per day)
    days       : simulation length
    dt         : time step (days)

    Returns
    -------
    List of dicts with keys time, S, I, R
    """
    if not (beta > 0 and gamma > 0 and days >= 1):
        raise ValueError("beta > 0, gamma > 0, days >= 1 required")

    N = S0 + I0 + R0
    steps = round(days / dt)
    results = [{"time": 0.0, "S": S0, "I": I0, "R": R0}]

    S, I, R = S0, I0, R0
    for i in range(1, steps + 1):
        new_inf = beta * S * I / N * dt
        new_rec = gamma * I * dt
        S -= new_inf
        I += new_inf - new_rec
        R += new_rec
        results.append({"time": round(i * dt, 6),
                         "S": S, "I": I, "R": R})
    return results
```

```python
# File: sir_model/__init__.py
from .model import sir_model   # makes  `from sir_model import sir_model` work
```

```python
# File: sir_model/tests/test_model.py
import pytest
from sir_model import sir_model

def test_conservation():
    N = 10_000
    result = sir_model(S0=9900, I0=100, beta=0.3, gamma=0.1, days=100)
    for row in result:
        total = row["S"] + row["I"] + row["R"]
        assert abs(total - N) < 0.01, f"N drift at t={row['time']}: {total}"

def test_subcritical():
    result = sir_model(S0=9990, I0=10, beta=0.05, gamma=0.2, days=200)
    assert result[-1]["I"] < 1.0, "Subcritical epidemic did not fade"

def test_invalid_input():
    with pytest.raises(ValueError):
        sir_model(S0=100, I0=10, beta=-0.3, gamma=0.1, days=50)
```

The directory structure looks like this:

```
sir_model/
├── __init__.py
├── model.py
└── tests/
    └── test_model.py
pyproject.toml       ← package metadata and dependencies
README.md
```

Running `pytest` from the project root executes all tests in under a second. If you change `model.py` and a test fails, you know immediately before any code reaches a user.

## The `pyproject.toml`: declaring what your code needs

```toml
[project]
name = "sir-model"
version = "0.1.0"
description = "Production SIR epidemic model"
requires-python = ">=3.11"
dependencies = [
    "numpy>=1.26",
    "pandas>=2.0",
]

[project.optional-dependencies]
dev = ["pytest", "ruff", "mypy"]

[build-system]
requires = ["hatchling"]
build-backend = "hatchling.build"
```

Anyone can now install your model with `pip install .` and the correct dependencies are fetched automatically. There are no "it works on my machine" problems because all dependencies are explicit.

## The ten-minute checklist

Before putting any simulation code in front of users, check these items [@taschuk2017]:

| Check | Research code | Production code |
|---|---|---|
| Paths | Hard-coded | Arguments or config file |
| Dependencies | Implicit | Listed in `requirements.txt` / `pyproject.toml` |
| Inputs | Trusted | Validated at boundaries |
| Random seed | Set interactively | Explicit argument |
| Tests | None | Conservation, boundary, regression |
| Error messages | R traceback / Python traceback | Informative user-facing message |
| Version | "latest" | Tagged release |

## What comes next

This post established the discipline needed to write software that others can depend on. The next post builds directly on it: we take the validated `sir_model` function and expose it as a REST API using FastAPI, so that any program — not just R or Python scripts — can call your model over HTTP.

## References

::: {#refs}
:::
