---
title: "Time-Series Data Storage with TimescaleDB"
subtitle: "Persisting simulation outputs, observations, and model states in a PostgreSQL extension built for time-series workloads. Post 6 in the Building Digital Twin Systems series."
author: "Jong-Hoon Kim"
date: "2026-04-23"
categories: [digital twin, database, TimescaleDB, PostgreSQL, R, data engineering]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why simulation data needs a database

After [Post 4](https://www.jonghoonk.com/blog/posts/dt_software_enkf/) (EnKF) and [Post 5](https://www.jonghoonk.com/blog/posts/dt_software_surrogate/) (GP surrogate), your digital twin produces several types of time-stamped output every day:

- **Observations**: daily case counts from surveillance
- **State estimates**: EnKF posterior for $S$, $E$, $I$, $R$, $\beta$
- **Forecasts**: 14-day forward projections with uncertainty
- **Scenario runs**: counterfactuals under alternative interventions

Saving these to CSV files works for one outbreak. It breaks when you need to:

- Query "all estimates for district X between March and June"
- Join surveillance data with model output by timestamp
- Write to the database from a Python API while reading from an R dashboard
- Retain 3 years of daily forecasts without running out of disk or memory

A relational database handles all of this. **TimescaleDB** [@freedman2019] is a PostgreSQL [@momjian2001] extension that adds time-series optimisations — automatic partitioning by time, efficient time-range queries, and continuous aggregates — while keeping full SQL compatibility.

## Core concepts

### PostgreSQL refresher

PostgreSQL is a relational database: data lives in **tables** (rows and columns), queries use SQL, and the database enforces data types and constraints. The key operations are:

```sql
-- Create a table
CREATE TABLE observations (
    time      TIMESTAMPTZ NOT NULL,
    location  TEXT        NOT NULL,
    cases     INTEGER     NOT NULL
);

-- Insert a row
INSERT INTO observations VALUES ('2024-03-01', 'district_a', 42);

-- Query with a time filter
SELECT * FROM observations
WHERE time >= '2024-03-01'
  AND time <  '2024-04-01'
ORDER BY time;
```

### What TimescaleDB adds

TimescaleDB converts a regular table into a **hypertable** with one command:

```sql
SELECT create_hypertable('observations', 'time');
```

Under the hood it partitions the table into **chunks** by time interval (default: 7 days). This means:

- Time-range queries touch only the relevant chunks — much faster than scanning the full table
- Old chunks can be compressed or moved to cheaper storage automatically
- Chunk pruning is transparent: you write the same SQL

## Schema design for a digital twin

A practical schema for an epidemic digital twin has four tables:

```sql
-- Raw surveillance data
CREATE TABLE observations (
    time          TIMESTAMPTZ NOT NULL,
    location_id   TEXT        NOT NULL,
    new_cases     INTEGER     NOT NULL,
    source        TEXT
);
SELECT create_hypertable('observations', 'time');

-- EnKF posterior state estimates
CREATE TABLE model_states (
    time          TIMESTAMPTZ NOT NULL,
    location_id   TEXT        NOT NULL,
    run_id        UUID        NOT NULL,
    S_median      REAL,
    E_median      REAL,
    I_median      REAL,
    R_median      REAL,
    beta_median   REAL,
    beta_lo95     REAL,
    beta_hi95     REAL
);
SELECT create_hypertable('model_states', 'time');

-- Short-term forecasts
CREATE TABLE forecasts (
    issued_at     TIMESTAMPTZ NOT NULL,  -- when the forecast was made
    target_time   TIMESTAMPTZ NOT NULL,  -- what time it predicts
    location_id   TEXT        NOT NULL,
    horizon_days  SMALLINT    NOT NULL,
    cases_median  REAL,
    cases_lo95    REAL,
    cases_hi95    REAL
);
SELECT create_hypertable('forecasts', 'issued_at');

-- Scenario / counterfactual runs
CREATE TABLE scenarios (
    run_id        UUID PRIMARY KEY DEFAULT gen_random_uuid(),
    created_at    TIMESTAMPTZ NOT NULL DEFAULT NOW(),
    description   TEXT,
    parameters    JSONB       -- flexible key-value store for run config
);
```

The `JSONB` column on `scenarios` stores arbitrary key-value metadata (beta, gamma, intervention timing) without requiring a schema change when parameters evolve.

## Simulating the workflow in R

We do not need a live database to work through the patterns — we can simulate the full workflow using R data frames with the same column structure. This lets you develop and test logic locally before wiring up a real TimescaleDB instance.

```{r}
#| label: setup
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)
```

### Generating synthetic data

```{r}
#| label: generate-data

set.seed(2024)

# Simulate 90 days of observations for two locations
n_days <- 90
dates  <- seq(as.Date("2024-01-01"), by = "day", length.out = n_days)

# SEIR peak parameters per location
make_obs <- function(beta, seed_val) {
  set.seed(seed_val)
  N <- 100000; S <- N - 50; E <- 30; I <- 20; R <- 0
  sigma <- 1/5; gamma <- 1/7; out <- numeric(n_days)
  for (d in seq_len(n_days)) {
    new_e <- beta * S * I / N
    new_i <- sigma * E
    new_r <- gamma * I
    S <- S - new_e; E <- E + new_e - new_i
    I <- I + new_i - new_r; R <- R + new_r
    out[d] <- rpois(1, max(1, new_i))
  }
  out
}

obs_a <- make_obs(0.35, 101)
obs_b <- make_obs(0.28, 202)

# "observations" table
df_obs <- data.frame(
  time        = rep(dates, 2),
  location_id = rep(c("district_a", "district_b"), each = n_days),
  new_cases   = c(obs_a, obs_b),
  source      = "surveillance"
)
```

### Simulating EnKF posterior storage

```{r}
#| label: enkf-states

# Simulate what the EnKF would write to model_states
# (using a simple running mean + noise for illustration)
make_states <- function(obs_vec, beta_true) {
  n <- length(obs_vec)
  beta_est  <- cumsum(c(0.25, rnorm(n - 1, 0, 0.005)))
  beta_est  <- pmin(pmax(beta_est, 0.1), 0.6)
  I_est     <- stats::filter(obs_vec, rep(1/7, 7), circular = TRUE)
  data.frame(
    beta_median = beta_est,
    beta_lo95   = beta_est - 0.05,
    beta_hi95   = beta_est + 0.05,
    I_median    = as.numeric(I_est)
  )
}

states_a <- make_states(obs_a, 0.35)
states_b <- make_states(obs_b, 0.28)

df_states <- data.frame(
  time        = rep(dates, 2),
  location_id = rep(c("district_a", "district_b"), each = n_days),
  run_id      = "run-2024-01-01",
  rbind(states_a, states_b)
)
```

### Generating forecasts

```{r}
#| label: forecasts

# Simple 14-day ahead forecast issued every 7 days
issue_days <- seq(21, n_days - 14, by = 7)

df_fcst <- do.call(rbind, lapply(issue_days, function(d) {
  base   <- df_obs$new_cases[df_obs$location_id == "district_a"][d]
  growth <- 0.97   # slight decline
  do.call(rbind, lapply(1:14, function(h) {
    mu <- base * growth^h
    data.frame(
      issued_at    = dates[d],
      target_time  = dates[d + h],
      location_id  = "district_a",
      horizon_days = h,
      cases_median = round(mu),
      cases_lo95   = round(mu * 0.6),
      cases_hi95   = round(mu * 1.5)
    )
  }))
}))
```

## Querying patterns in SQL and R

The value of a database is its query language. Here are the four most common patterns for a digital twin dashboard, shown as equivalent R `dplyr` operations [@wickham2019] alongside their SQL counterparts.

### Pattern 1: time-range filter

```sql
-- SQL
SELECT time, new_cases
FROM   observations
WHERE  location_id = 'district_a'
  AND  time BETWEEN '2024-02-01' AND '2024-03-31';
```

```{r}
#| label: query-time-range

df_obs |>
  filter(location_id == "district_a",
         time >= as.Date("2024-02-01"),
         time <= as.Date("2024-03-31")) |>
  select(time, new_cases) |>
  head(5)
```

### Pattern 2: join observations with model state

```sql
-- SQL
SELECT o.time, o.new_cases, s.beta_median, s.I_median
FROM   observations o
JOIN   model_states s
  ON   o.time = s.time AND o.location_id = s.location_id
WHERE  o.location_id = 'district_a';
```

```{r}
#| label: query-join
#| fig-cap: "Observed cases (bars) joined with the EnKF β estimate (line). This is the core diagnostic plot for an operational digital twin."

joined <- df_obs |>
  filter(location_id == "district_a") |>
  inner_join(
    df_states |> filter(location_id == "district_a"),
    by = c("time", "location_id")
  )

ggplot(joined, aes(x = time)) +
  geom_col(aes(y = new_cases), fill = "grey70", width = 0.8) +
  geom_line(aes(y = beta_median * 800), colour = "steelblue",
            linewidth = 1) +
  geom_ribbon(aes(ymin = beta_lo95 * 800, ymax = beta_hi95 * 800),
              fill = "steelblue", alpha = 0.2) +
  scale_y_continuous(
    name  = "Daily new cases",
    sec.axis = sec_axis(~ . / 800, name = "β estimate")
  ) +
  labs(x = NULL, title = "Observed cases and EnKF β estimate: district A") +
  theme_minimal(base_size = 13)
```

### Pattern 3: forecast vs actuals (calibration)

```sql
-- SQL: compare 7-day ahead forecasts to what actually happened
SELECT f.issued_at, f.target_time,
       f.cases_median, f.cases_lo95, f.cases_hi95,
       o.new_cases AS actual
FROM   forecasts f
JOIN   observations o
  ON   f.target_time = o.time AND f.location_id = o.location_id
WHERE  f.location_id = 'district_a'
  AND  f.horizon_days = 7;
```

```{r}
#| label: query-calibration
#| fig-cap: "Seven-day-ahead forecast (line and ribbon) vs actuals (points). Coverage of the 95% interval can be read directly from how many points fall inside the ribbon."

cal <- df_fcst |>
  filter(horizon_days == 7) |>
  inner_join(
    df_obs |> filter(location_id == "district_a") |>
      rename(actual = new_cases),
    by = c("target_time" = "time", "location_id")
  )

ggplot(cal, aes(x = target_time)) +
  geom_ribbon(aes(ymin = cases_lo95, ymax = cases_hi95),
              fill = "orange", alpha = 0.3) +
  geom_line(aes(y = cases_median), colour = "darkorange", linewidth = 1) +
  geom_point(aes(y = actual), colour = "grey30", size = 1.5) +
  labs(x = NULL, y = "New cases",
       title = "7-day ahead forecast vs actuals") +
  theme_minimal(base_size = 13)
```

### Pattern 4: TimescaleDB continuous aggregate (weekly totals)

In TimescaleDB, continuous aggregates pre-compute summaries and update incrementally:

```sql
-- Create a weekly aggregate (runs automatically as data arrives)
CREATE MATERIALIZED VIEW weekly_cases
WITH (timescaledb.continuous) AS
SELECT time_bucket('7 days', time) AS week,
       location_id,
       SUM(new_cases)              AS total_cases,
       AVG(new_cases)              AS avg_daily
FROM   observations
GROUP BY week, location_id;

-- Query it like a regular table
SELECT * FROM weekly_cases
WHERE week >= '2024-02-01'
ORDER BY week, location_id;
```

The R equivalent for local development:

```{r}
#| label: weekly-agg
#| fig-cap: "Weekly case totals per district — equivalent to a TimescaleDB continuous aggregate query. In production this query reads a pre-computed materialised view and returns in milliseconds regardless of the raw table size."

df_obs |>
  mutate(week = as.Date(cut(time, "week"))) |>
  group_by(week, location_id) |>
  summarise(total_cases = sum(new_cases), .groups = "drop") |>
  ggplot(aes(x = week, y = total_cases, fill = location_id)) +
  geom_col(position = "dodge", width = 5) +
  scale_fill_manual(values = c(district_a = "steelblue",
                               district_b = "firebrick"),
                    name = "Location") +
  labs(x = NULL, y = "Weekly cases",
       title = "Weekly case totals by district") +
  theme_minimal(base_size = 13)
```

## Connecting from R and Python

In a deployed system, both the EnKF runner (R) and the dashboard (Python) connect to the same TimescaleDB instance using standard database drivers.

**From R** using the `DBI` and `RPostgres` packages:

```r
library(DBI)
library(RPostgres)

con <- dbConnect(
  Postgres(),
  host     = "localhost",
  port     = 5432,
  dbname   = "dtdb",
  user     = Sys.getenv("DB_USER"),
  password = Sys.getenv("DB_PASSWORD")
)

# Write EnKF output
dbWriteTable(con, "model_states", df_states, append = TRUE)

# Read for dashboard
df <- dbGetQuery(con, "
  SELECT time, beta_median, beta_lo95, beta_hi95
  FROM   model_states
  WHERE  location_id = 'district_a'
  ORDER  BY time DESC
  LIMIT  30
")

dbDisconnect(con)
```

**From Python** using `psycopg2` or `SQLAlchemy`:

```python
import psycopg2
import pandas as pd

conn = psycopg2.connect(
    host="localhost", dbname="dtdb",
    user=os.environ["DB_USER"],
    password=os.environ["DB_PASSWORD"]
)

df = pd.read_sql("""
    SELECT time, new_cases
    FROM   observations
    WHERE  location_id = %s
    ORDER  BY time
""", conn, params=("district_a",))
```

Using `Sys.getenv()` / `os.environ` for credentials keeps passwords out of code and out of git.

## Running TimescaleDB locally with Docker

The Docker Compose file from [Post 3](https://www.jonghoonk.com/blog/posts/dt_software_docker/) already included a TimescaleDB service:

```yaml
db:
  image: timescale/timescaledb:latest-pg16
  environment:
    POSTGRES_USER: dtuser
    POSTGRES_PASSWORD: secret
    POSTGRES_DB: dtdb
  ports:
    - "5432:5432"
  volumes:
    - pgdata:/var/lib/postgresql/data
  healthcheck:
    test: ["CMD-SHELL", "pg_isready -U dtuser -d dtdb"]
    interval: 5s
    retries: 10
```

After `docker compose up`, connect with `psql -h localhost -U dtuser -d dtdb` and run the schema creation SQL above. The data volume `pgdata` persists the database across container restarts.

## Design rules for production

**Choose the right time column**. `issued_at` on forecasts (when the forecast was made) versus `target_time` (what it predicts) serve different queries. Store both — `issued_at` for hypertable partitioning, `target_time` for calibration joins.

**Index location and run IDs**. Add `CREATE INDEX ON model_states (location_id, time DESC)` so per-location queries skip unneeded chunks.

**Compress old chunks**. TimescaleDB's native compression can reduce storage by 90% for old time ranges: `SELECT add_compression_policy('observations', INTERVAL '30 days')`.

**Use connection pooling**. Open a new database connection per API request is expensive. Use PgBouncer or SQLAlchemy's connection pool in production.

## References

::: {#refs}
:::
