---
title: "Secrets Management: Keeping Credentials Out of Your Code"
subtitle: "Hardcoded passwords are a security vulnerability and a compliance failure. AWS Secrets Manager, environment variables, and secret rotation done right. Skill 15 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, security, secrets management, AWS, DevOps]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why secrets management is non-negotiable

A developer pushes a database connection string to a public GitHub repository by accident. Within 15 minutes, a bot discovers it and begins exfiltrating patient data. This is not a hypothetical — it happens dozens of times per day across GitHub [@owasp2021].

Any health department client will conduct at minimum a basic security review before signing a contract. "Where are your database passwords stored?" is question one. "In the code" or "in a `.env` file in the repository" ends the conversation.

Proper secrets management is both a security practice and a commercial requirement. It is also straightforward to implement.

## The hierarchy of bad to good

```
❌ Hardcoded in source code:   password = "secret123"
❌ In .env committed to git:   DATABASE_URL=postgresql://...
⚠  In .env not committed:     ok for local dev only
✓  Environment variable at runtime: -e DB_PASSWORD=...
✓  AWS Secrets Manager:        retrieved at startup, rotated automatically
✓  HashiCorp Vault:            enterprise-grade, audited, dynamic secrets
```

## Environment variables: the baseline

The minimum safe approach: never put secrets in source code. Pass them as environment variables at runtime.

**Docker run:**
```bash
docker run -p 8000:8000 \
  -e DATABASE_URL="postgresql://user:pass@host:5432/db" \
  -e API_SECRET_KEY="$(openssl rand -hex 32)" \
  dt-api:latest
```

**R code that reads them:**
```r
db_url  <- Sys.getenv("DATABASE_URL",   unset = stop("DATABASE_URL not set"))
api_key <- Sys.getenv("API_SECRET_KEY", unset = stop("API_SECRET_KEY not set"))
```

```{r}
#| label: env-demo

# Demonstrate safe secret reading pattern in R
safe_get_env <- function(key, required = TRUE) {
  val <- Sys.getenv(key, unset = NA_character_)
  if (is.na(val) || nchar(val) == 0) {
    if (required) stop(sprintf("Required environment variable '%s' not set.", key))
    return(NULL)
  }
  val
}

# Show what the pattern produces
tryCatch(
  safe_get_env("DATABASE_URL"),   # this env var is not set
  error = function(e) cat("Error (expected):", conditionMessage(e), "\n")
)

# Simulate a set variable
Sys.setenv(DT_API_KEY_TEST = "fake-key-for-demo-only")
cat("Key found:", nchar(safe_get_env("DT_API_KEY_TEST")), "chars\n")
Sys.unsetenv("DT_API_KEY_TEST")
```

## AWS Secrets Manager

For production, AWS Secrets Manager [@aws_sm] stores secrets encrypted at rest, controls access via IAM, and supports automatic rotation.

```bash
# Store a secret (one-time)
aws secretsmanager create-secret \
  --name "dt/prod/db-password" \
  --description "TimescaleDB password for prod environment" \
  --secret-string '{"username":"dtuser","password":"S3cur3P@ss"}'

# Retrieve at runtime (in startup script)
SECRET=$(aws secretsmanager get-secret-value \
  --secret-id "dt/prod/db-password" \
  --query SecretString --output text)

DB_USER=$(echo $SECRET | python3 -c "import sys,json; print(json.load(sys.stdin)['username'])")
DB_PASS=$(echo $SECRET | python3 -c "import sys,json; print(json.load(sys.stdin)['password'])")

export DATABASE_URL="postgresql://${DB_USER}:${DB_PASS}@localhost:5432/dtdb"
```

**In R:**
```r
library(aws.secretsmanager)

get_db_creds <- function(secret_name = "dt/prod/db-password") {
  raw <- get_secret(secret_name)
  creds <- jsonlite::fromJSON(raw$SecretString)
  list(user = creds$username, password = creds$password)
}
```

## Automatic rotation

Secrets Manager can rotate the database password automatically every 30–90 days. Set it up once:

```bash
aws secretsmanager rotate-secret \
  --secret-id "dt/prod/db-password" \
  --rotation-rules AutomaticallyAfterDays=30
```

When rotation fires, Secrets Manager calls a Lambda function that updates the RDS user's password and stores the new value. Your application always calls `get_secret()` at startup — it never needs to know a rotation happened.

## Scanning for leaked secrets

Add `git-secrets` or `gitleaks` as a pre-commit hook to prevent accidental commits:

```bash
# Install gitleaks
brew install gitleaks    # or download binary

# Scan the whole repo history
gitleaks detect --source . --log-opts HEAD

# Add as pre-commit hook
gitleaks protect --staged
```

## The secrets audit checklist

```{r}
#| label: audit-checklist
#| echo: false

df <- data.frame(
  Item = c(
    "No secrets in source code (grep for common patterns)",
    ".env files in .gitignore",
    "Secrets passed as environment variables or Secrets Manager",
    "Different credentials per environment (dev ≠ prod)",
    "Database users have minimum required permissions",
    "API keys have expiry dates and can be revoked",
    "Rotation policy configured for long-lived secrets",
    "Audit trail: who accessed which secret and when"
  ),
  Priority = c("Critical","Critical","Critical","High",
               "High","High","Medium","Medium")
)
knitr::kable(df, caption = "Secrets management audit checklist")
```

Every item marked "Critical" will be raised in a security review by a health department procurement team. Tick them off before your first enterprise client conversation.

## References

::: {#refs}
:::
