---
title: "A Lightweight Data Lake for Epidemic Surveillance Ingestion"
subtitle: "Hospital data, wastewater, and syndromic surveillance land in different formats. Parquet + DuckDB unifies them into one queryable store. Skill 12 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, data engineering, Parquet, DuckDB, Arrow, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The heterogeneous data problem

A real-time epidemic digital twin ingests from multiple sources simultaneously:

- Hospital: daily admissions CSV emailed at 8 AM
- Wastewater: JSON from lab API, updated 3×/week
- Syndromic surveillance: nightly SQL dump from the health information system
- Seroprevalence surveys: Excel files, irregular timing

Routing all of these directly into TimescaleDB requires a schema for each, an ingestion script for each, and a way to handle delays, duplicates, and format changes. The classic data engineering pattern is to land raw data in a **data lake** first — a cheap, schema-flexible storage layer — and then push cleaned, normalised records to the model database.

## The Parquet + DuckDB stack

**Parquet** [@richardson2023] is a column-oriented file format with efficient compression. A 100MB CSV typically compresses to 10–20MB Parquet with dramatically faster analytical queries.

**DuckDB** [@raasveldt2019] is an in-process analytical database that queries Parquet files directly — no server required. Together they implement a lightweight data lake that runs on a single EC2 instance.

```
Raw files (S3 or local)
  hospital/2024-04-01.csv
  wastewater/2024-04-01.json
  syndromic/2024-04-01.parquet
         │
         ▼  (ingestion scripts)
Landing zone  → Parquet partitioned by source + date
         │
         ▼  (DuckDB transform query)
Cleaned layer → Normalised Parquet schema
         │
         ▼
TimescaleDB   → observations table
```

## Building the ingestion pipeline in R

```{r}
#| label: simulate-raw-data

library(dplyr)

# Simulate raw data from three heterogeneous sources
set.seed(2024)
n_days <- 14
dates  <- seq(as.Date("2024-04-01"), by = "day", length.out = n_days)

# Source 1: hospital admissions (CSV-like: district, date, admissions)
df_hospital <- data.frame(
  District    = rep(c("North", "South"), each = n_days),
  ReportDate  = rep(dates, 2),
  Admissions  = rpois(n_days * 2, 30),
  DataVersion = "v2.1"
)

# Source 2: wastewater (JSON-like: different naming, units)
df_ww <- data.frame(
  district_code = rep(c("N", "S"), each = n_days),
  sample_date   = rep(dates, 2),
  gc_per_ml     = round(runif(n_days * 2, 100, 5000)),
  lab_id        = "WW_LAB_001"
)

# Source 3: syndromic surveillance (SQL dump: ICD-coded ED visits)
df_syndromic <- data.frame(
  location  = rep(c("North", "South"), each = n_days),
  visit_dt  = rep(dates, 2),
  icd_flu   = rpois(n_days * 2, 15),
  icd_resp  = rpois(n_days * 2, 45)
)

cat("Hospital rows:", nrow(df_hospital),
    " | WW rows:", nrow(df_ww),
    " | Syndromic rows:", nrow(df_syndromic), "\n")
```

```{r}
#| label: normalise

# Normalise to a common schema: date, location_id, source, metric, value
normalise <- function(df, source_name) {
  switch(source_name,
    hospital = df |>
      transmute(
        date        = ReportDate,
        location_id = tolower(gsub(" ", "_", District)),
        source      = "hospital",
        metric      = "admissions",
        value       = as.numeric(Admissions)
      ),
    wastewater = df |>
      transmute(
        date        = sample_date,
        location_id = ifelse(district_code == "N", "north", "south"),
        source      = "wastewater",
        metric      = "gc_per_ml",
        value       = as.numeric(gc_per_ml)
      ),
    syndromic = df |>
      transmute(
        date        = visit_dt,
        location_id = tolower(location),
        source      = "syndromic",
        metric      = "flu_ed_visits",
        value       = as.numeric(icd_flu)
      )
  )
}

df_lake <- bind_rows(
  normalise(df_hospital,  "hospital"),
  normalise(df_ww,        "wastewater"),
  normalise(df_syndromic, "syndromic")
)

cat("Unified lake rows:", nrow(df_lake), "\n")
head(df_lake, 4)
```

```{r}
#| label: parquet-duckdb
#| eval: false

# In production: write to Parquet and query with DuckDB
library(arrow)
library(duckdb)

# Write partitioned Parquet (partitioned by source + date)
write_dataset(df_lake, path = "data_lake/cleaned",
              format = "parquet",
              partitioning = c("source", "date"))

# Query with DuckDB — no need to load into memory
con <- dbConnect(duckdb::duckdb())
dbExecute(con, "
  CREATE VIEW lake AS
  SELECT * FROM parquet_scan('data_lake/cleaned/**/*.parquet',
                              hive_partitioning = TRUE)
")
# Aggregate across sources
result <- dbGetQuery(con, "
  SELECT date, location_id, source, SUM(value) AS total
  FROM lake
  WHERE metric IN ('admissions', 'flu_ed_visits')
  GROUP BY date, location_id, source
  ORDER BY date, location_id
")
dbDisconnect(con)
```

```{r}
#| label: lake-viz
#| fig-cap: "Unified surveillance lake: three heterogeneous data sources normalised to a common schema and visualised together. This is the view your EnKF ingestion script queries each morning."

library(ggplot2)

ggplot(df_lake |> filter(metric %in% c("admissions","flu_ed_visits")),
       aes(x = date, y = value, colour = source)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ location_id, ncol = 2) +
  scale_colour_manual(values = c(hospital   = "steelblue",
                                  syndromic  = "firebrick"),
                      name = "Source") +
  labs(x = NULL, y = "Count", title = "Unified surveillance lake: two sources, two districts") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Partitioning strategy

Partition your Parquet lake by `source` and `date`:

```
data_lake/cleaned/source=hospital/date=2024-04-01/part-0.parquet
data_lake/cleaned/source=wastewater/date=2024-04-01/part-0.parquet
```

DuckDB's partition pruning means a query filtered to `source='hospital' AND date='2024-04-01'` reads only a few kilobytes, even if the full lake is hundreds of gigabytes.

## From lake to model

The EnKF runner queries the lake each morning:

```r
# Pull yesterday's unified case counts for all districts
obs_today <- df_lake |>
  filter(
    date   == Sys.Date() - 1,
    metric == "admissions",
    source == "hospital"
  ) |>
  arrange(location_id)

# Feed into EnKF
run_enkf(obs_today$value, obs_today$location_id)
```

The lake is the contract between the messy world of surveillance data and the clean model pipeline that depends on it.

## References

::: {#refs}
:::
