---
title: "Gradient Descent: The Optimization Engine Behind Every ML Model"
subtitle: "All of machine learning reduces to minimizing a loss function. Gradient descent does this iteratively — and understanding it unlocks every other ML technique."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [machine learning, optimization, R, epidemic modeling]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why optimization is the heart of ML

Every machine learning algorithm — from linear regression to GPT-4 — is doing the same thing: finding parameter values that minimize a loss function. The loss measures how wrong the model is. Gradient descent is the algorithm that walks downhill on that loss surface to find the minimum.

For an epidemic modeler, this is familiar territory. Fitting a transmission rate $\beta$ to case data by minimizing sum-of-squared errors *is* gradient descent — most implementations just call it `optim()` and hide the mechanics. Understanding those mechanics is what separates someone who *uses* ML tools from someone who can *build and debug* them.

## The algorithm

Given a loss function $L(\theta)$ over parameters $\theta$, gradient descent updates parameters each iteration:

$$\theta_{t+1} = \theta_t - \eta \nabla_\theta L(\theta_t)$$

where $\eta$ is the **learning rate** (step size) and $\nabla_\theta L$ is the gradient — the direction of steepest ascent. Stepping opposite to the gradient moves downhill.

Three variants differ in how much data each step uses:

| Variant | Data per step | Noise | Typical use |
|---------|--------------|-------|-------------|
| Batch GD | Full dataset | Low | Small data, convex losses |
| Stochastic GD (SGD) | 1 sample | High | Online learning |
| Mini-batch GD | 32–512 samples | Medium | Neural networks |

## From concept to code: fitting an epidemic growth curve

Early epidemic growth is approximately exponential: $I(t) = I_0 \, e^{rt}$. Given noisy case counts, we want to recover $I_0$ and $r$. This is a 2-parameter optimization problem — a clean place to watch gradient descent work.

```{r}
#| label: gd-setup

set.seed(42)

# Simulate early epidemic (14 days)
t_obs   <- 0:14
I0_true <- 10
r_true  <- 0.18
I_true  <- I0_true * exp(r_true * t_obs)
I_obs   <- round(pmax(1, I_true * exp(rnorm(length(t_obs), 0, 0.15))))

# Loss: mean squared error on log scale
loss <- function(params) {
  I_hat <- params[1] * exp(params[2] * t_obs)
  mean((log(I_obs) - log(pmax(1e-9, I_hat)))^2)
}

# Numerical gradient (central differences)
grad <- function(f, p, eps = 1e-6) {
  vapply(seq_along(p), function(i) {
    p_hi <- p_lo <- p
    p_hi[i] <- p_hi[i] + eps
    p_lo[i] <- p_lo[i] - eps
    (f(p_hi) - f(p_lo)) / (2 * eps)
  }, numeric(1))
}

# Run gradient descent from a naive starting point
params  <- c(I0 = 5, r = 0.05)
lr      <- 0.01
n_iter  <- 300

history <- data.frame(
  iter = seq_len(n_iter),
  loss = NA_real_,
  I0   = NA_real_,
  r    = NA_real_
)

for (k in seq_len(n_iter)) {
  g       <- grad(loss, params)
  params  <- params - lr * g
  history[k, ] <- c(k, loss(params), params["I0"], params["r"])
}

cat("Recovered I0:", round(params["I0"], 1), " (true:", I0_true, ")\n")
cat("Recovered r: ", round(params["r"],  4), " (true:", r_true,  ")\n")
```

```{r}
#| label: gd-plots
#| fig-cap: "Gradient descent convergence. Left: loss falls rapidly in the first 50 iterations then plateaus. Right: the fitted curve closely tracks the observed counts after 300 steps."

library(ggplot2)

# Panel A: loss curve
p_loss <- ggplot(history, aes(x = iter, y = loss)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  labs(x = "Iteration", y = "Log-scale MSE",
       title = "A: Loss curve") +
  theme_minimal(base_size = 12)

# Panel B: fitted vs observed
t_fine  <- seq(0, 14, length.out = 100)
I_fit   <- params["I0"] * exp(params["r"] * t_fine)
df_obs  <- data.frame(t = t_obs, cases = I_obs)
df_fit  <- data.frame(t = t_fine, cases = I_fit)

p_fit <- ggplot() +
  geom_point(data = df_obs, aes(x = t, y = cases),
             colour = "firebrick", size = 2) +
  geom_line(data = df_fit,  aes(x = t, y = cases),
            colour = "steelblue", linewidth = 1) +
  labs(x = "Day", y = "Infectious cases",
       title = "B: Fitted vs observed") +
  theme_minimal(base_size = 12)

# Side by side with base R layout
gridExtra_available <- requireNamespace("gridExtra", quietly = TRUE)
if (gridExtra_available) {
  gridExtra::grid.arrange(p_loss, p_fit, ncol = 2)
} else {
  print(p_loss)
  print(p_fit)
}
```

## The loss landscape: why learning rate matters

The loss surface $L(I_0, r)$ is a bowl in this case (convex). But in general neural networks have many saddle points and local minima. The learning rate $\eta$ controls step size:

- Too large: parameters overshoot the minimum, loss diverges
- Too small: converges correctly but takes thousands of unnecessary iterations
- Just right: rapid initial descent, smooth convergence

```{r}
#| label: loss-landscape
#| fig-cap: "Loss surface for the epidemic growth parameters (log scale). The gradient descent path (white dots) converges from the starting point (top-left) to the true minimum (center). The contours reveal the elongated valley that makes a fixed learning rate suboptimal — the reason adaptive optimisers like Adam exist."

# Compute loss on a grid
I0_grid <- seq(2, 25, length.out = 60)
r_grid  <- seq(0.05, 0.35, length.out = 60)

grid_df <- expand.grid(I0 = I0_grid, r = r_grid)
grid_df$loss <- vapply(
  seq_len(nrow(grid_df)),
  function(i) loss(c(grid_df$I0[i], grid_df$r[i])),
  numeric(1)
)

# Subsample path for readability
path_sub <- history[seq(1, n_iter, by = 10), ]

ggplot(grid_df, aes(x = I0, y = r, z = log(loss + 1e-6))) +
  geom_contour_filled(bins = 20, alpha = 0.85) +
  geom_point(data = path_sub, aes(x = I0, y = r, z = NULL),
             colour = "white", size = 1.2) +
  geom_point(aes(x = I0_true, y = r_true, z = NULL),
             colour = "yellow", shape = 4, size = 4, stroke = 2) +
  scale_fill_viridis_d(option = "plasma", guide = "none") +
  labs(x = expression(I[0]), y = "r",
       title = "Loss landscape: gradient descent path (white) to true minimum (×)") +
  theme_minimal(base_size = 12)
```

## Beyond vanilla gradient descent: Adam

The Adam optimizer [@kingma2014] adapts the learning rate per parameter using estimates of the first (momentum) and second (variance) moments of the gradient:

```python
# Adam update (conceptual — not executed)
m = beta1 * m + (1 - beta1) * grad        # 1st moment
v = beta2 * v + (1 - beta2) * grad**2     # 2nd moment
m_hat = m / (1 - beta1**t)                # bias correction
v_hat = v / (1 - beta2**t)
theta = theta - lr * m_hat / (sqrt(v_hat) + eps)
```

Adam is the default in PyTorch and TensorFlow because it handles elongated loss landscapes (like the one above) much better than fixed-step gradient descent. The default `beta1 = 0.9`, `beta2 = 0.999`, `lr = 1e-3` work well for most neural networks.

## What to take away

Gradient descent is not a black box — it is an iterative walk down a loss surface, one parameter update at a time. The practical lessons:

1. **Log-transform count data** before computing MSE — raw counts produce loss surfaces that are hard to descend.
2. **Plot your loss curve**. If it doesn't fall, your learning rate is too high or the gradient is broken.
3. **Adam is the safe default** for neural networks; `optim(method = "L-BFGS-B")` is the safe default for low-dimensional epidemic fitting.
4. **Numerical gradients** (central differences) let you debug autodiff implementations — if numerical and analytical gradients disagree, your backprop is wrong.

## References

::: {#refs}
:::
