---
title: "Global Sensitivity Analysis as a Billable Model Audit Service"
subtitle: "Which parameters drive your epidemic forecast? Sobol indices and Morris screening turn model uncertainty into a professional deliverable. Skill 2 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, sensitivity analysis, Sobol, epidemic modeling, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The business case

A health department client asks: *"You say $R_0 = 1.8$, but how sensitive is your 14-day forecast to that estimate? What if it's actually 2.2?"* Without sensitivity analysis you can only shrug. With it, you can hand them a one-page report quantifying exactly how much each uncertain parameter moves the forecast needle.

Global sensitivity analysis (SA) [@saltelli2002; @marino2008] answers: **which inputs explain most of the output variance?** This is a different question from "what happens if I change $\beta$ by 10%" (local sensitivity). Global SA samples the entire input space and decomposes variance into contributions from each parameter and their interactions.

It is also a distinct, defensible deliverable — a model audit that commands consulting fees separate from the core subscription.

## Variance-based (Sobol) indices

For a model $Y = f(X_1, \ldots, X_k)$, the first-order Sobol index for parameter $X_i$ is:

$$S_i = \frac{V_{X_i}[\,E_{\mathbf{X}_{\sim i}}(Y \mid X_i)\,]}{V(Y)}$$

It measures the fraction of output variance explained by $X_i$ alone. The total-order index $S_{T_i}$ captures $X_i$'s contribution including all interactions with other parameters:

$$S_{T_i} = 1 - \frac{V_{X_{\sim i}}[\,E_{X_i}(Y \mid \mathbf{X}_{\sim i})\,]}{V(Y)}$$

When $S_{T_i} \gg S_i$, interactions dominate. Both indices lie in $[0,1]$ and $\sum_i S_i \leq 1$ [@saltelli2010].

## Implementing Sobol indices in R

We compute Sobol indices for peak infected count in an SEIR model with three uncertain inputs: $\beta$, $\sigma$ (incubation rate), and $\gamma$ (recovery rate).

```{r}
#| label: seir-model

# SEIR model returning peak infectious count
seir_peak <- function(beta, sigma, gamma,
                      N = 100000, I0 = 50, days = 120) {
  S <- N - I0; E <- 0; I <- I0; R <- 0
  peak <- I
  for (d in seq_len(days)) {
    new_e <- beta * S * I / N
    new_i <- sigma * E
    new_r <- gamma * I
    S <- S - new_e; E <- E + new_e - new_i
    I <- I + new_i - new_r; R <- R + new_r
    if (I > peak) peak <- I
  }
  peak / N   # return as fraction
}
```

```{r}
#| label: sobol-mc

set.seed(2024)

# Parameter ranges (uniform priors)
ranges <- list(
  beta  = c(0.20, 0.55),
  sigma = c(1/8,  1/3),
  gamma = c(1/10, 1/5)
)

# Saltelli (2010) estimator needs two independent base samples A, B
n <- 4000
A <- sapply(ranges, function(r) runif(n, r[1], r[2]))
B <- sapply(ranges, function(r) runif(n, r[1], r[2]))

# Evaluate model on A and B
yA <- mapply(seir_peak, A[,"beta"], A[,"sigma"], A[,"gamma"])
yB <- mapply(seir_peak, B[,"beta"], B[,"sigma"], B[,"gamma"])

# For each parameter i, create AB_i: A with column i replaced by B
sobol_indices <- function(yA, yB, A, B, ranges) {
  k  <- ncol(A)
  nms <- colnames(A)
  S  <- numeric(k); ST <- numeric(k)
  Vy <- var(c(yA, yB))

  for (i in seq_len(k)) {
    AB_i <- A
    AB_i[, i] <- B[, i]
    yABi <- mapply(seir_peak,
                   AB_i[,"beta"], AB_i[,"sigma"], AB_i[,"gamma"])

    # Jansen estimator (more robust than Saltelli original)
    S[i]  <- (var(yB) - mean((yB - yABi)^2) / 2) / Vy
    ST[i] <- mean((yA - yABi)^2) / (2 * Vy)
  }
  data.frame(parameter = nms,
             S1 = pmax(0, S),
             ST = pmax(0, ST))
}

idx <- sobol_indices(yA, yB, A, B, ranges)
print(idx)
```

```{r}
#| label: plot-sobol
#| fig-cap: "First-order (S1) and total-order (ST) Sobol indices for peak infectious fraction. Parameters with large ST but small S1 have significant interaction effects."

library(ggplot2)

idx_long <- rbind(
  data.frame(parameter = idx$parameter, index = "First-order (S1)",
             value = idx$S1),
  data.frame(parameter = idx$parameter, index = "Total-order (ST)",
             value = idx$ST)
)

ggplot(idx_long, aes(x = reorder(parameter, -value),
                     y = value, fill = index)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_manual(values = c("First-order (S1)" = "steelblue",
                               "Total-order (ST)"  = "firebrick"),
                    name = NULL) +
  geom_hline(yintercept = 0.05, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.6, y = 0.07,
           label = "5% threshold", hjust = 0, size = 3.2, colour = "grey40") +
  labs(x = "Parameter", y = "Sobol index",
       title = "Global sensitivity: SEIR peak infectious fraction") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

## Morris screening for quick triage

When you have many parameters (20+), full Sobol estimation is expensive. The **Morris method** [@marino2008] is a cheap rank-order screen: run the model $r(k+1)$ times (rather than $2nk$), and compute the mean $\mu^*$ and standard deviation $\sigma$ of elementary effects.

- High $\mu^*$: parameter has large influence
- High $\sigma$: parameter has nonlinear effects or interactions

```{r}
#| label: morris
#| fig-cap: "Morris plot: parameters with high mean effect (μ*) are important; high standard deviation (σ) signals interactions or nonlinearity. Parameters in the bottom-left are unimportant and can be fixed."

morris_screen <- function(model_fn, ranges, r = 20, delta = 0.5) {
  k   <- length(ranges)
  nms <- names(ranges)
  EE  <- matrix(0, nrow = r, ncol = k)

  for (rep in seq_len(r)) {
    # Random base point in [0, 1-delta]
    x0 <- sapply(seq_len(k), function(i) runif(1, 0, 1 - delta))
    # Evaluate at base
    params0 <- mapply(function(x, r) r[1] + x * diff(r), x0, ranges)
    y0 <- do.call(model_fn, as.list(setNames(params0, nms)))

    # Randomly permute dimensions and apply +delta one at a time
    perm <- sample(k)
    x_current <- x0
    for (j in perm) {
      x_new <- x_current
      x_new[j] <- x_current[j] + delta
      params_new <- mapply(function(x, r) r[1] + x * diff(r),
                           x_new, ranges)
      y_new <- do.call(model_fn, as.list(setNames(params_new, nms)))
      EE[rep, j] <- (y_new - y0) / delta
      y0 <- y_new
      x_current <- x_new
    }
  }

  data.frame(
    parameter = nms,
    mu_star   = colMeans(abs(EE)),
    sigma_ee  = apply(EE, 2, sd)
  )
}

morris_res <- morris_screen(seir_peak, ranges, r = 50)

ggplot(morris_res, aes(x = mu_star, y = sigma_ee,
                       label = parameter)) +
  geom_point(size = 4, colour = "steelblue") +
  geom_text(size = 4, nudge_x = 0.005, hjust = 0) +
  labs(x = "μ* (mean |elementary effect|)",
       y = "σ (SD of elementary effects)",
       title = "Morris sensitivity screening: SEIR model") +
  theme_minimal(base_size = 13)
```

## Packaging as a deliverable

A sensitivity analysis report for a client should include:

1. **Parameter ranges** — what assumptions went into the SA
2. **Sobol indices table** with confidence intervals (bootstrap)
3. **Ranking narrative** — "transmission rate $\beta$ explains 78% of forecast variance"
4. **Implications** — which parameters are worth measuring better vs. fixing at nominal values

The `sensitivity` R package provides production-ready estimators with confidence intervals. The `SALib` Python library is the equivalent for Python-based models.

## References

::: {#refs}
:::
