---
title: "HIPAA-Aware Architecture Patterns for Health Data on AWS"
subtitle: "Any US health department client will ask about HIPAA. Design for compliance from the start — the right AWS services, the right configurations, the right documentation. Skill 20 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, HIPAA, compliance, AWS, security, health data]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## HIPAA and your business

The Health Insurance Portability and Accountability Act [@hipaa1996] governs Protected Health Information (PHI) in the United States. If your digital twin processes surveillance data that could identify individual patients — hospital admissions by age/sex/zip code, for example — you are likely handling PHI.

HIPAA compliance is not a product feature; it is a prerequisite for US government health contracts. A health department procurement team will ask three questions:

1. Do you have a Business Associate Agreement (BAA)?
2. Which AWS services do you use, and do they cover PHI under the BAA?
3. How do you enforce minimum necessary access?

Answering yes, accurately, to all three unlocks contracts that competitors without HIPAA posture cannot win.

## What HIPAA requires technically

The HIPAA Security Rule [@nist800] specifies three safeguard categories:

**Administrative**: who is authorised, what training they have, what your incident response plan is.

**Physical**: where your servers are, who can access them (less relevant for cloud deployments — AWS handles this under the BAA).

**Technical**: encryption at rest and in transit, access controls, audit logs, automatic session timeout.

The technical safeguards map directly to AWS service configurations.

## HIPAA-eligible AWS services

Not all AWS services are covered under the BAA. The key ones for your digital twin stack:

```{r}
#| label: hipaa-services
#| echo: false

df <- data.frame(
  Service = c("EC2", "RDS (PostgreSQL)", "S3", "Secrets Manager",
              "CloudWatch Logs", "VPC", "IAM", "KMS", "ELB/ALB",
              "Lambda"),
  HIPAA_Eligible = c("Yes","Yes","Yes","Yes",
                     "Yes","Yes","Yes","Yes","Yes","Yes"),
  Use_in_DT = c("API + EnKF server", "TimescaleDB",
                "Data lake / backups", "Credentials",
                "Audit logs", "Network isolation",
                "Access control", "Encryption keys",
                "HTTPS termination", "Scheduled jobs"),
  Encryption_Default = c("No (add EBS encryption)","Yes","Yes","Yes",
                          "Yes","N/A","N/A","N/A","Yes","N/A")
)

knitr::kable(df, caption = "Key AWS services: HIPAA eligibility and digital twin use")
```

Enable encryption at rest explicitly for EC2 EBS volumes — it is not on by default.

## Architecture checklist

```{r}
#| label: arch-checklist
#| fig-cap: "HIPAA-compliant digital twin architecture. PHI never leaves the private subnet; all external access flows through the ALB and is authenticated. Audit logs capture every access event."

library(ggplot2)

# Architecture diagram as a text figure
df_nodes <- data.frame(
  x     = c(1,   3,   3,   5,   5,   7),
  y     = c(3,   5,   1,   5,   1,   3),
  label = c("Internet\n(HTTPS only)",
            "ALB\n(TLS termination)",
            "CloudWatch\nLogs",
            "EC2\n(API + EnKF)\nPrivate subnet",
            "RDS\nTimescaleDB\nPrivate subnet",
            "S3\n(Parquet lake)\nEncrypted"),
  colour = c("grey60","steelblue","orange","steelblue","steelblue","darkgreen")
)

df_edges <- data.frame(
  x    = c(1.3, 3.3, 4.3, 3.3),
  xend = c(2.7, 4.7, 4.7, 4.7),
  y    = c(3,   5,   5,   1),
  yend = c(5,   5,   1,   1)
)

ggplot() +
  geom_segment(data = df_edges,
               aes(x=x, xend=xend, y=y, yend=yend),
               arrow = arrow(length=unit(0.2,"cm")),
               colour = "grey50", linewidth = 0.8) +
  geom_label(data = df_nodes, aes(x=x, y=y, label=label, fill=colour),
             colour = "white", size = 3, label.size = 0, fontface = "bold") +
  scale_fill_identity() +
  xlim(0.5, 7.5) + ylim(0, 6) +
  labs(title = "HIPAA-compliant digital twin architecture") +
  theme_void(base_size = 13)
```

## The 10 technical controls

```{r}
#| label: controls
#| echo: false

controls <- data.frame(
  Control = c(
    "Encryption in transit (TLS 1.2+)",
    "Encryption at rest (AES-256)",
    "Network isolation (VPC, no public DB)",
    "Minimum privilege IAM roles",
    "Multi-factor authentication for console",
    "Audit logging (CloudTrail + CloudWatch)",
    "Automatic session timeout (15 min idle)",
    "Database access logging",
    "Backup with 7-year retention",
    "Incident response plan documented"
  ),
  AWS_Mechanism = c(
    "ALB + ACM certificate",
    "RDS encryption, S3 SSE-KMS, EBS encryption",
    "Private subnets, security groups",
    "IAM roles with least-privilege policies",
    "AWS IAM Identity Center",
    "CloudTrail data events, VPC flow logs",
    "Cognito/application-level enforcement",
    "PostgreSQL pg_audit extension",
    "RDS automated backups + S3 lifecycle",
    "Written policy document + tested procedure"
  ),
  Priority = c(rep("Required", 6), rep("Required", 4))
)

knitr::kable(controls, caption = "HIPAA technical safeguards mapped to AWS")
```

## The Business Associate Agreement

Before processing any PHI, sign a **BAA** with AWS — this is a legal agreement where AWS acknowledges their role as your Business Associate and commits to HIPAA-compliant handling of PHI on their infrastructure.

Sign it in the AWS console under Account → Business Associate Addendum. It covers all HIPAA-eligible services automatically once signed.

You will also need to sign a BAA with each client — you are their Business Associate when you process their surveillance data. Use a standard template reviewed by a health IT attorney; it is a one-time cost that protects you and signals professionalism.

## Synthetic data for development

Never use real patient data in development or testing. Generate **synthetic surveillance data** that has the same statistical properties as real data but cannot be traced to individuals. The simulation code from throughout this series is already doing this — use the same approach for test fixtures.

This practice — using synthetic data in non-production environments — is explicitly recommended by NIST 800-53 [@nist800] and eliminates most PHI risk from your development workflow.

## Beyond the checklist

HIPAA compliance is not a one-time audit — it is ongoing. Key practices after go-live:

- **Annual security review**: re-run the checklist and update documentation
- **Penetration testing**: hire an external firm to test your API annually
- **Employee training**: anyone who touches production data needs annual HIPAA training
- **Breach response**: you have 60 days to notify HHS of a breach affecting 500+ individuals

A compliance posture that is documented, tested, and maintained is worth far more to a health department procurement team than a checkbox in a sales deck.

## References

::: {#refs}
:::
