---
title: "Database Schema Migrations for Production TimescaleDB"
subtitle: "Your schema will evolve. Alembic and disciplined migration scripts make schema changes safe, reviewable, and reversible. Skill 18 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, database, migrations, Alembic, PostgreSQL, DevOps]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The schema evolution problem

Your digital twin has been running in production for two months. You want to add a `severity_level` column to the `model_states` table — a field that did not exist when the schema was designed. How do you:

1. Add the column to the production database without downtime?
2. Ensure your staging environment matches production?
3. Roll back if something goes wrong?
4. Communicate the change to teammates who run their own local databases?

Manual `ALTER TABLE` statements run in the console answer question 1 but fail questions 2–4. **Database migrations** [@alembic2010; @humble2010] are versioned, executable scripts that transform the schema from state A to state B — and, critically, back again.

## Alembic: the Python migration tool

Alembic is the standard migration library for SQLAlchemy-based Python projects (FastAPI, Django). It generates migration scripts that:

- Track the current database revision
- Apply pending migrations in order
- Reverse them on rollback

```bash
# Initialise Alembic in your project
alembic init migrations

# Auto-generate a migration from model changes
alembic revision --autogenerate -m "add severity level to model states"

# Apply all pending migrations
alembic upgrade head

# Roll back one step
alembic downgrade -1

# Show current revision
alembic current
```

## A migration script

Alembic generates this file; you review and edit it:

```python
# migrations/versions/a1b2c3_add_severity_level.py
"""add severity level to model states

Revision ID: a1b2c3d4e5f6
Revises: 9f8e7d6c5b4a
Create Date: 2024-04-15 09:00:00
"""
from alembic import op
import sqlalchemy as sa

def upgrade():
    """Add severity_level column with default NULL."""
    op.add_column(
        "model_states",
        sa.Column("severity_level", sa.String(16), nullable=True)
    )
    # Backfill existing rows with a computed value
    op.execute("""
        UPDATE model_states
        SET    severity_level = CASE
                   WHEN I_median > 1000 THEN 'high'
                   WHEN I_median > 200  THEN 'medium'
                   ELSE 'low'
               END
    """)
    # Now make NOT NULL after backfill
    op.alter_column("model_states", "severity_level", nullable=False)

def downgrade():
    """Remove the column — data is lost."""
    op.drop_column("model_states", "severity_level")
```

## Migration safety rules

```{r}
#| label: migration-rules

library(ggplot2)

rules <- data.frame(
  Rule = c(
    "Always write a downgrade() function",
    "Backfill before adding NOT NULL constraints",
    "Never rename a column (add new + copy + drop old)",
    "Test on a copy of prod data before applying",
    "Apply migrations before deploying new code",
    "Store migration scripts in version control",
    "Never edit a migration that has been applied to prod"
  ),
  Category = c("Rollback","Safety","Compatibility",
               "Testing","Deployment","Process","Integrity"),
  Risk = c("High","High","High","Medium","High","Medium","Critical")
)

knitr::kable(rules[, c("Rule","Category","Risk")],
             caption = "Database migration safety rules")
```

## Simulating migration state tracking in R

```{r}
#| label: migration-tracker
#| fig-cap: "Migration history: each revision builds on the previous one, forming a linear chain. Alembic tracks which revision each environment is on. When dev is ahead of prod, the next deployment applies the gap."

# Simulate migration history
set.seed(1)
migrations <- data.frame(
  revision    = paste0("rev_", seq_len(8)),
  description = c("initial schema",
                  "add observations table",
                  "add model_states table",
                  "add hypertables",
                  "add forecasts table",
                  "add tenant_id columns",
                  "add severity_level",
                  "add usage_events table"),
  applied_dev  = c(rep(TRUE, 8)),
  applied_prod = c(rep(TRUE, 6), FALSE, FALSE),
  applied_date = seq(as.Date("2024-01-01"), by = "week", length.out = 8)
)

migrations$colour <- ifelse(migrations$applied_prod, "steelblue", "firebrick")

ggplot(migrations, aes(x = seq_len(nrow(migrations)),
                        y = 1, fill = applied_prod)) +
  geom_tile(colour = "white", linewidth = 1, height = 0.6) +
  geom_text(aes(label = gsub("_", "\n", description)),
            size = 2.5, colour = "white") +
  scale_fill_manual(values = c("TRUE"  = "steelblue",
                                "FALSE" = "firebrick"),
                    labels  = c("TRUE" = "Applied to prod",
                                "FALSE" = "Pending"),
                    name = NULL) +
  scale_x_continuous(name = "Migration sequence", breaks = 1:8) +
  labs(y = NULL, title = "Migration state: prod is 2 revisions behind dev") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "top")
```

## Zero-downtime migrations

For tables with millions of rows (TimescaleDB hypertables), a naive `ADD COLUMN NOT NULL` locks the table and causes downtime. The safe sequence:

```sql
-- Step 1 (deploy immediately, no lock): add nullable column with default
ALTER TABLE model_states
  ADD COLUMN severity_level TEXT DEFAULT NULL;

-- Step 2 (background job, hours): backfill existing rows in small batches
UPDATE model_states
  SET severity_level = 'medium'
  WHERE severity_level IS NULL
  LIMIT 10000;  -- repeat until done

-- Step 3 (deploy next version): add NOT NULL constraint after all rows are filled
ALTER TABLE model_states
  ALTER COLUMN severity_level SET NOT NULL;
```

Each step is a separate migration file. The deployment is spread across multiple releases — each release is safe independently.

## The CI/CD integration

Add migration tests to your GitHub Actions workflow:

```yaml
- name: Run database migrations
  run: |
    alembic upgrade head
    python -m pytest tests/test_migrations.py
```

The test file verifies that the schema matches the SQLAlchemy models after migration, catching drift before it reaches production.

## References

::: {#refs}
:::
