---
title: "Securing Your API with JWT Authentication and Role-Based Access"
subtitle: "Multi-tenant SaaS requires per-client tokens, role enforcement, and key revocation. JWT + OAuth2 is the industry standard. Skill 17 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, security, authentication, JWT, FastAPI, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why authentication is non-negotiable

Your epidemic model API is live. Without authentication, anyone who discovers the URL can call `/simulate` without limit, read your clients' outbreak data, and flood the server with requests. More practically: every enterprise client will ask "how do we authenticate?" before signing a contract.

**JWT (JSON Web Token)** [@jones2015] is the standard token format for API authentication. It encodes identity and permissions as a digitally signed JSON object. Combined with **OAuth2** [@hardt2012], it supports delegated access: a health department can issue tokens to their own staff without giving out your master API key.

## How JWT works

```
Client                           Server
──────                           ──────
POST /auth/token                 Verify username + password
{username, password}    ──►      Create JWT:
                                   header.payload.signature
                        ◄──      {"access_token": "eyJ...", "expires_in": 3600}

GET /simulate           ──►      Decode JWT
Authorization: Bearer eyJ...     Verify signature
                                 Check expiry
                                 Extract tenant_id, role
                        ◄──      Return result (or 401)
```

The server never stores the token. The JWT's digital signature means the server can verify it was issued by them — without a database lookup on every request.

## JWT structure

A JWT has three parts separated by dots: `header.payload.signature`

```json
// Header (base64-encoded)
{"alg": "HS256", "typ": "JWT"}

// Payload (base64-encoded, NOT encrypted — visible to anyone)
{
  "sub": "dept_a_user",
  "tenant_id": "district_health_dept_a",
  "role": "analyst",
  "exp": 1735689600
}

// Signature (HMAC-SHA256 of header + payload using secret key)
HMAC-SHA256(base64(header) + "." + base64(payload), SECRET_KEY)
```

**Important**: the payload is visible but tamper-proof. Never put passwords or sensitive data in it.

## FastAPI implementation

```python
from fastapi import FastAPI, Depends, HTTPException, status
from fastapi.security import OAuth2PasswordBearer, OAuth2PasswordRequestForm
from jose import JWTError, jwt
from datetime import datetime, timedelta
import bcrypt

SECRET_KEY = "your-256-bit-secret"   # from environment variable
ALGORITHM  = "HS256"
TOKEN_TTL_MINUTES = 60

oauth2_scheme = OAuth2PasswordBearer(tokenUrl="/auth/token")

def create_token(sub: str, tenant_id: str, role: str) -> str:
    payload = {
        "sub":       sub,
        "tenant_id": tenant_id,
        "role":      role,
        "exp":       datetime.utcnow() + timedelta(minutes=TOKEN_TTL_MINUTES)
    }
    return jwt.encode(payload, SECRET_KEY, algorithm=ALGORITHM)

def verify_token(token: str = Depends(oauth2_scheme)) -> dict:
    try:
        payload = jwt.decode(token, SECRET_KEY, algorithms=[ALGORITHM])
        if payload.get("sub") is None:
            raise HTTPException(status_code=401, detail="Invalid token")
        return payload
    except JWTError:
        raise HTTPException(status_code=401, detail="Invalid token")

def require_role(required_role: str):
    def checker(payload: dict = Depends(verify_token)):
        if payload.get("role") != required_role and \
           payload.get("role") != "admin":
            raise HTTPException(status_code=403,
                                detail=f"Role '{required_role}' required")
        return payload
    return checker

app = FastAPI()

@app.post("/auth/token")
def get_token(form: OAuth2PasswordRequestForm = Depends()):
    user = lookup_user(form.username)   # query your users table
    if not user or not bcrypt.checkpw(form.password.encode(),
                                       user["password_hash"].encode()):
        raise HTTPException(status_code=401, detail="Invalid credentials")
    token = create_token(user["username"], user["tenant_id"], user["role"])
    return {"access_token": token, "token_type": "bearer"}

@app.post("/simulate")
def simulate(req: SimulateRequest,
             user: dict = Depends(verify_token)):
    # user["tenant_id"] is now available for RLS (Skill 11)
    return run_model(req, tenant_id=user["tenant_id"])

@app.delete("/admin/users/{user_id}")
def delete_user(user_id: str,
                admin: dict = Depends(require_role("admin"))):
    # Only admins reach here
    return delete_user_record(user_id)
```

## Roles and permissions in R

The R SDK verifies the token and extracts roles:

```{r}
#| label: jwt-demo

# Illustrate JWT payload structure in R (no encoding library needed)
# A real JWT payload is base64url-encoded JSON — we show it decoded.

jwt_payload <- list(
  sub       = "alice",
  tenant_id = "district_health_dept_a",
  role      = "analyst",
  iat       = as.integer(Sys.time()),
  exp       = as.integer(Sys.time()) + 3600L
)

cat("JWT payload (decoded):\n")
cat("  sub      :", jwt_payload$sub,       "\n")
cat("  tenant_id:", jwt_payload$tenant_id, "\n")
cat("  role     :", jwt_payload$role,      "\n")
cat("  issued   :", format(as.POSIXct(jwt_payload$iat), "%Y-%m-%d %H:%M:%S"), "\n")
cat("  expires  :", format(as.POSIXct(jwt_payload$exp), "%Y-%m-%d %H:%M:%S"), "\n")

# Verify the token has not expired
is_valid <- jwt_payload$exp > as.integer(Sys.time())
cat("\nToken valid:", is_valid, "\n")
cat("Note: the payload is base64url-encoded but NOT encrypted;",
    "never put passwords or PHI in it.\n")
```

```{r}
#| label: role-matrix
#| fig-cap: "Role-permission matrix for a digital twin API. Each cell shows whether the role can perform the action. Analyst can query but not admin; Admin can do everything."

library(ggplot2)

roles   <- c("Read-only", "Analyst", "Manager", "Admin")
actions <- c("View forecasts", "Run simulations", "Export data",
             "Upload observations", "Manage users", "Billing access")

perms <- matrix(c(
  TRUE,  FALSE, FALSE, FALSE, FALSE, FALSE,
  TRUE,  TRUE,  TRUE,  FALSE, FALSE, FALSE,
  TRUE,  TRUE,  TRUE,  TRUE,  FALSE, FALSE,
  TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE
), nrow = 4, byrow = TRUE)

df_perm <- expand.grid(role = roles, action = actions)
df_perm$allowed <- as.vector(perms)

ggplot(df_perm, aes(x = factor(action, levels = rev(actions)),
                    y = factor(role, levels = roles),
                    fill = allowed)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(allowed, "✓", "✗")),
            size = 5, colour = "white") +
  scale_fill_manual(values = c("FALSE" = "firebrick",
                                "TRUE"  = "steelblue"),
                    guide = "none") +
  scale_x_discrete(name = NULL) +
  scale_y_discrete(name = NULL) +
  labs(title = "Role-based access control matrix") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Token refresh and revocation

Access tokens should be short-lived (1 hour). Issue a **refresh token** (long-lived, stored in a database) so clients can get new access tokens without re-entering a password.

To revoke a compromised token: store a **denylist** in Redis. The `verify_token` function checks the token's `jti` (unique ID) against the denylist before accepting it.

## References

::: {#refs}
:::
