---
title: "Real-Time State Estimation with the Ensemble Kalman Filter"
subtitle: "Updating an epidemic model as data arrive — from the classic KF to the nonlinear ensemble version. Post 4 in the Building Digital Twin Systems series."
author: "Jong-Hoon Kim"
date: "2026-04-23"
categories: [digital twin, Kalman filter, data assimilation, R, epidemic modeling]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The gap between a model and a digital twin

An epidemic model fed fixed parameters produces a single trajectory. A **digital twin** does something more: it continuously updates its internal state as real data arrive, so that at every point in time the model reflects the current best estimate of reality.

The standard Kalman filter [@kalman1960] achieves this for **linear** systems. An SIR model is nonlinear — transmission depends on the product $SI$, not a sum. The **Ensemble Kalman Filter** (EnKF) [@evensen1994; @evensen2003] extends the idea to nonlinear models by representing uncertainty through a cloud of particles (an *ensemble*) rather than an analytical covariance matrix. It was first applied to epidemic forecasting for influenza by Shaman et al. [@shaman2013] and is now a core tool for real-time surveillance.

## The Kalman filter idea

The Kalman filter alternates two steps.

**Forecast step** — propagate the current state estimate forward using the model:
$$\hat{x}_t^- = f(\hat{x}_{t-1})$$
$$P_t^- = F P_{t-1} F^\top + Q$$
where $P$ is the error covariance, $F$ is the linearised model Jacobian, and $Q$ is process noise.

**Update step** — correct the forecast using a new observation $y_t$:
$$K_t = P_t^- H^\top (H P_t^- H^\top + R)^{-1}$$
$$\hat{x}_t = \hat{x}_t^- + K_t(y_t - H \hat{x}_t^-)$$
$$P_t = (I - K_t H) P_t^-$$

$H$ maps the model state to the observation space, $R$ is observation noise, and $K$ is the **Kalman gain** — a weight that balances model uncertainty against observation uncertainty.

The EnKF replaces the analytical $P$ with a sample covariance computed from an ensemble of $N$ model runs.

## The EnKF algorithm

```
Initialise: draw N particles from the prior
For each observation time t:
  1. FORECAST: propagate each particle through the model
  2. ANALYSE:
     a. Compute ensemble mean and sample covariance P
     b. For each particle i, add observation noise: ỹᵢ = y_t + ε_i, ε_i ~ N(0, R)
     c. Compute Kalman gain: K = P Hᵀ (H P Hᵀ + R)⁻¹
     d. Update each particle: xᵢ ← xᵢ + K(ỹᵢ - H xᵢ)
```

The perturbed observations in step 2b prevent ensemble collapse (all particles converging to the same point) and preserve the theoretical covariance relationship [@evensen2003].

## Implementation in R

We use an SEIR model: $S \to E \to I \to R$, where $E$ is a latent exposed compartment with mean duration $1/\sigma$. The observed variable is the new infectious count $\Delta I$ (newly symptomatic cases per day), not the total $I$ — matching real surveillance data.

### The process model

```{r}
#| label: process-model

seir_step <- function(state, beta, sigma, gamma, dt = 1.0) {
  S <- max(0, state["S"],  na.rm = TRUE)
  E <- max(0, state["E"],  na.rm = TRUE)
  I <- max(0, state["I"],  na.rm = TRUE)
  R <- max(0, state["R"],  na.rm = TRUE)
  if (!is.finite(beta) || beta <= 0) beta <- 0.001
  N <- S + E + I + R
  if (!is.finite(N) || N <= 0) {
    return(c(S = 0, E = 0, I = 0, R = 0, deltaI = 0))
  }

  new_exp <- beta * S * I / N * dt
  new_inf <- sigma * E * dt
  new_rec <- gamma * I * dt

  c(
    S     = S - new_exp,
    E     = E + new_exp - new_inf,
    I     = I + new_inf - new_rec,
    R     = R + new_rec,
    deltaI = new_inf           # observed: new infections
  )
}
```

### Generating synthetic observations

We simulate a "true" outbreak and observe daily new cases with Poisson noise — a common model for reported case counts.

```{r}
#| label: synthetic-data

set.seed(42)

# True parameters
N_pop   <- 100000
beta_t  <- 0.35
sigma_t <- 1 / 5   # incubation 5 days
gamma_t <- 1 / 7   # infectious 7 days

# Simulate true trajectory
true_state <- c(S = N_pop - 50, E = 30, I = 20, R = 0, deltaI = 0)
n_days <- 90
true_traj <- matrix(0, nrow = n_days + 1, ncol = 5)
colnames(true_traj) <- names(true_state)
true_traj[1, ] <- true_state

for (d in seq_len(n_days)) {
  true_traj[d + 1, ] <- seir_step(true_traj[d, ],
                                   beta_t, sigma_t, gamma_t)
}

# Observed: Poisson noise on new infections
obs <- rpois(n_days, lambda = pmax(1, true_traj[-1, "deltaI"]))
```

```{r}
#| label: plot-truth
#| fig-cap: "True SEIR trajectory (lines) and observed daily cases (points). Poisson noise makes the observations stochastic."

library(ggplot2)

days <- 0:n_days
df_truth <- data.frame(
  day   = days,
  S     = true_traj[, "S"],
  E     = true_traj[, "E"],
  I     = true_traj[, "I"],
  R     = true_traj[, "R"]
)

ggplot(df_truth, aes(x = day)) +
  geom_line(aes(y = E, colour = "Exposed")) +
  geom_line(aes(y = I, colour = "Infectious")) +
  geom_point(data = data.frame(day = seq_len(n_days), obs = obs),
             aes(x = day, y = obs, colour = "Observed"),
             size = 0.8, alpha = 0.7) +
  scale_colour_manual(values = c(Exposed = "orange",
                                 Infectious = "firebrick",
                                 Observed = "grey40")) +
  labs(x = "Day", y = "Count", colour = NULL) +
  theme_minimal(base_size = 13)
```

### The EnKF function

```{r}
#| label: enkf

enkf_seir <- function(obs,           # vector of daily observations
                      N_pop,         # population size
                      N_ens = 200,   # ensemble size
                      sigma = 1/5,   # fixed incubation rate
                      gamma = 1/7,   # fixed recovery rate
                      beta_mean = 0.3,
                      beta_sd   = 0.1,
                      obs_var   = NULL) {

  n_obs   <- length(obs)
  # State vector: (S, E, I, R, log_beta)
  # log_beta is estimated alongside states
  n_state <- 5

  # Initialise ensemble from prior
  ens <- matrix(0, nrow = N_ens, ncol = n_state)
  colnames(ens) <- c("S", "E", "I", "R", "log_beta")

  # Spread initial infections across S and I
  I_init      <- round(rnorm(N_ens, 20, 5))
  I_init      <- pmax(I_init, 1)
  ens[, "I"]  <- I_init
  ens[, "S"]  <- N_pop - I_init
  ens[, "E"]  <- 0
  ens[, "R"]  <- 0
  ens[, "log_beta"] <- rnorm(N_ens, log(beta_mean), beta_sd)

  # Containers for posterior summaries
  beta_post <- matrix(NA, nrow = n_obs, ncol = 3)  # median, q2.5, q97.5
  I_post    <- matrix(NA, nrow = n_obs, ncol = 3)
  deltaI_post <- matrix(NA, nrow = n_obs, ncol = 3)

  for (t in seq_len(n_obs)) {

    # Sanitise ensemble before forecast: replace any NaN/Inf with safe values
    for (col in c("S", "E", "I", "R")) {
      bad <- !is.finite(ens[, col])
      ens[bad, col] <- 0
    }
    bad_lb <- !is.finite(ens[, "log_beta"])
    ens[bad_lb, "log_beta"] <- log(beta_mean)

    # --- 1. Forecast ---
    beta_ens <- exp(ens[, "log_beta"])
    state_names <- c("S", "E", "I", "R", "deltaI")
    forecast_ens <- matrix(0, nrow = N_ens, ncol = 5)
    colnames(forecast_ens) <- state_names

    for (i in seq_len(N_ens)) {
      st_vals <- pmax(0, ens[i, c("S", "E", "I", "R")])
      st_vals[!is.finite(st_vals)] <- 0
      st <- c(S = st_vals[1], E = st_vals[2],
              I = st_vals[3], R = st_vals[4], deltaI = 0)
      forecast_ens[i, ] <- seir_step(st, beta_ens[i], sigma, gamma)
    }

    # Add random walk noise to log_beta (model evolution)
    ens[, "log_beta"] <- ens[, "log_beta"] + rnorm(N_ens, 0, 0.02)

    # Update states from forecast
    ens[, c("S","E","I","R")] <- forecast_ens[, c("S","E","I","R")]
    ens[, ] <- pmax(ens, 0)   # keep non-negative

    # --- 2. Analyse ---
    # Observation operator H: maps state → observed deltaI
    H_vec <- forecast_ens[, "deltaI"]   # one observation per particle

    # Observation noise variance
    obs_r <- if (is.null(obs_var)) max(obs[t], 1) else obs_var

    # Ensemble covariance of (state, Hy)
    state_full <- cbind(ens[, c("S","E","I","R")],
                        log_beta = ens[, "log_beta"])
    ensemble_mean <- colMeans(state_full)
    A <- sweep(state_full, 2, ensemble_mean)   # anomalies (N x p)
    hy_mean <- mean(H_vec)
    hy_anom <- H_vec - hy_mean                  # scalar anomaly vector (N)

    # Cross-covariance P H^T  (p x 1)  and  H P H^T + R  (scalar)
    PHT  <- crossprod(A, hy_anom) / (N_ens - 1)  # p x 1
    HPHT <- sum(hy_anom^2) / (N_ens - 1)          # scalar
    K    <- PHT / (HPHT + obs_r)                   # Kalman gain p x 1

    # Perturbed observations
    y_perturbed <- obs[t] + rnorm(N_ens, 0, sqrt(obs_r))
    innovation  <- y_perturbed - H_vec

    # Update
    update <- outer(innovation, as.vector(K))   # N x p
    update[!is.finite(update)] <- 0
    state_full <- state_full + update
    state_full[, 1:4] <- pmax(state_full[, 1:4], 0)
    # Clamp log_beta to avoid exp() overflow
    state_full[, 5] <- pmin(pmax(state_full[, 5], -5), 2)
    # Final NaN guard after update
    for (j in 1:4) {
      bad <- !is.finite(state_full[, j])
      state_full[bad, j] <- 0
    }
    bad_lb2 <- !is.finite(state_full[, 5])
    state_full[bad_lb2, 5] <- log(beta_mean)

    ens[, c("S","E","I","R")] <- state_full[, 1:4]
    ens[, "log_beta"]         <- state_full[, 5]

    # Store posterior summaries (na.rm guards against rare numerics)
    beta_samples <- exp(ens[, "log_beta"])
    beta_post[t, ]   <- quantile(beta_samples, c(0.5, 0.025, 0.975),
                                  na.rm = TRUE)
    I_post[t, ]      <- quantile(ens[, "I"],   c(0.5, 0.025, 0.975),
                                  na.rm = TRUE)
    deltaI_post[t, ] <- quantile(forecast_ens[, "deltaI"],
                                  c(0.5, 0.025, 0.975), na.rm = TRUE)
  }

  list(beta  = beta_post,
       I     = I_post,
       deltaI = deltaI_post)
}
```

### Running the filter

```{r}
#| label: run-enkf

set.seed(7)
fit <- enkf_seir(obs, N_pop = N_pop, N_ens = 300,
                 sigma = sigma_t, gamma = gamma_t,
                 beta_mean = 0.3, beta_sd = 0.15)
```

## Results

```{r}
#| label: plot-beta
#| fig-cap: "Posterior median and 95% credible interval for the transmission rate β over time. The EnKF correctly tracks the true value (dashed line) from a vague prior."

days_obs <- seq_len(n_days)

df_beta <- data.frame(
  day    = days_obs,
  median = fit$beta[, 1],
  lo     = fit$beta[, 2],
  hi     = fit$beta[, 3]
)

ggplot(df_beta, aes(x = day)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue",
              alpha = 0.3) +
  geom_line(aes(y = median), colour = "steelblue", linewidth = 1) +
  geom_hline(yintercept = beta_t, linetype = "dashed",
             colour = "firebrick") +
  annotate("text", x = 75, y = beta_t + 0.015,
           label = paste0("True β = ", beta_t),
           colour = "firebrick", size = 3.5) +
  labs(x = "Day", y = "β (transmission rate)",
       title = "EnKF: real-time β estimation") +
  coord_cartesian(ylim = c(0.1, 0.7)) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: plot-forecast
#| fig-cap: "One-step-ahead forecast of daily new cases (median and 95% CI) vs observed counts. After a short warm-up period the filter tracks the outbreak closely."

df_obs <- data.frame(
  day    = days_obs,
  median = fit$deltaI[, 1],
  lo     = fit$deltaI[, 2],
  hi     = fit$deltaI[, 3],
  obs    = obs
)

ggplot(df_obs, aes(x = day)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "orange",
              alpha = 0.35) +
  geom_line(aes(y = median), colour = "darkorange", linewidth = 1) +
  geom_point(aes(y = obs), colour = "grey30", size = 0.9) +
  labs(x = "Day", y = "New cases",
       title = "One-step-ahead forecast vs observations") +
  theme_minimal(base_size = 13)
```

The filter recovers $\beta$ close to its true value (0.35) after approximately 15 days of data, and the one-step-ahead forecasts track the observed counts closely throughout.

## Key design choices

**Ensemble size** ($N$). Larger ensembles give better covariance estimates but are slower. For a 5-state problem, $N = 200$–$500$ is usually sufficient. For high-dimensional spatial models, $N$ may be $10^3$–$10^4$.

**Process noise on parameters**. Adding a small random walk to $\log \beta$ each step allows the filter to track slowly changing transmission rates (e.g., seasonality, behaviour change). The noise magnitude controls the speed of adaptation: too large causes jitter, too small prevents tracking.

**Observation variance** $R$. We used `obs[t]` (Poisson-variance approximation). A better approach calibrates $R$ from historical data or estimates it jointly.

**Covariance inflation**. If the ensemble collapses (all particles become similar), the filter can lose the ability to update. Multiplicative inflation — multiplying ensemble anomalies by a factor slightly greater than 1 — prevents this.

## Connecting to a digital twin product

In a deployed system, the EnKF runs as a scheduled job (e.g., every night when new case counts are published):

```
New surveillance data
        │
        ▼
  Fetch from database  (Post 6: TimescaleDB)
        │
        ▼
  Run EnKF             (this post)
        │
        ▼
  Store posterior      (Post 6: TimescaleDB)
        │
        ▼
  Update dashboard     (Post 8: Streamlit)
```

The result is a model whose internal state continuously tracks reality — the defining property of an operational digital twin.

## References

::: {#refs}
:::
