---
title: "Random Forests: Ensemble Learning for Outbreak Risk Scoring"
subtitle: "Averaging many imperfect models beats any single model. Random forests apply this to decision trees and are often the strongest baseline for tabular public health data."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [machine learning, random forests, ensemble methods, R, public health]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## Why ensemble methods matter for public health

Epidemic risk scoring — flagging districts that are likely to see an outbreak — is a classification problem on tabular data: vaccination coverage, population density, age structure, prior case history. Neural networks can solve this, but they require large datasets and careful tuning. Random forests [@breiman2001] are often more accurate with smaller datasets and are far easier to interpret.

The core insight behind ensembles is **variance reduction through averaging**. A single decision tree is high-variance: small changes in training data produce very different trees. Average 500 trees, each trained on a bootstrap sample with random feature subsets, and the variance collapses while bias stays low [@hastie2009].

## Decision trees: the building block

A decision tree partitions the feature space with a sequence of binary splits. Each split is chosen to maximally reduce **Gini impurity**:

$$G = 1 - \sum_{k} p_k^2$$

where $p_k$ is the proportion of class $k$ in the node. The split that minimises the weighted average Gini of the two children is selected at each step.

The result is a step function that can capture non-linear, non-additive relationships — things like "high outbreak risk only when *both* vaccination rate is below 60% *and* population density exceeds 1000/km²."

## Simulating outbreak risk data

```{r}
#| label: rf-data

set.seed(2025)
n <- 400

df <- data.frame(
  vax_coverage  = runif(n, 0.30, 0.95),
  pop_density   = rlnorm(n, log(500), 0.6),
  prior_cases   = rpois(n, lambda = 15),
  surveillance  = runif(n, 0.2, 1.0)   # reporting completeness
)

# Outbreak occurs when coverage is low AND density high, with noise
log_odds <- -3
  + 4.0 * (1 - df$vax_coverage)
  + 0.0015 * df$pop_density
  + 0.03 * df$prior_cases
  - 1.5 * df$surveillance

prob <- 1 / (1 + exp(-log_odds))
df$outbreak <- rbinom(n, 1, prob)

cat("Outbreak rate:", round(mean(df$outbreak), 3), "\n")
```

## Implementing bagging from scratch

Bagging (bootstrap aggregating) trains many independent models and takes a majority vote. We implement a single-feature threshold classifier as the base learner — the simplest possible tree — to show the mechanics without a package dependency.

```{r}
#| label: rf-bagging

# Find the best threshold for one feature (binary classification)
best_split <- function(x, y) {
  thresholds <- quantile(x, probs = seq(0.1, 0.9, by = 0.05))
  best <- list(thr = 0, dir = 1, acc = 0)
  for (thr in thresholds) {
    for (dir in c(1, -1)) {
      pred <- as.integer(dir * x >= dir * thr)
      acc  <- mean(pred == y)
      if (acc > best$acc) best <- list(thr = thr, dir = dir, acc = acc)
    }
  }
  best
}

predict_stump <- function(split, x) {
  as.integer(split$dir * x >= split$dir * split$thr)
}

# Bagged ensemble of stumps (one randomly chosen feature per tree)
features <- c("vax_coverage", "pop_density", "prior_cases", "surveillance")

bag_train <- function(df, y_col, n_trees = 300) {
  lapply(seq_len(n_trees), function(i) {
    idx  <- sample(nrow(df), replace = TRUE)
    feat <- sample(features, 1)
    sp   <- best_split(df[idx, feat], df[idx, y_col])
    list(feat = feat, split = sp)
  })
}

bag_predict <- function(forest, df) {
  votes <- sapply(forest, function(tree) {
    predict_stump(tree$split, df[[tree$feat]])
  })
  as.integer(rowMeans(votes) >= 0.5)
}

# Train and evaluate in-sample (for illustration)
forest    <- bag_train(df, "outbreak", n_trees = 300)
pred_bag  <- bag_predict(forest, df)

# Compare single stump vs ensemble
single_feat <- "vax_coverage"
sp_single   <- best_split(df[[single_feat]], df$outbreak)
pred_single <- predict_stump(sp_single, df[[single_feat]])

cat("Single stump accuracy:", round(mean(pred_single == df$outbreak), 3), "\n")
cat("Bagged ensemble accuracy:", round(mean(pred_bag   == df$outbreak), 3), "\n")
```

## Variable importance via permutation

Permutation importance measures how much accuracy drops when a feature's values are randomly shuffled — breaking any information it carries [@breiman2001].

```{r}
#| label: rf-importance
#| fig-cap: "Permutation variable importance for the bagged ensemble. Vaccination coverage and population density matter most for outbreak prediction — consistent with the data-generating mechanism. Surveillance completeness has a modest protective effect."

base_acc <- mean(bag_predict(forest, df) == df$outbreak)

importance <- sapply(features, function(feat) {
  df_perm <- df
  df_perm[[feat]] <- sample(df_perm[[feat]])   # shuffle this feature
  acc_perm <- mean(bag_predict(forest, df_perm) == df$outbreak)
  base_acc - acc_perm                           # drop in accuracy
})

imp_df <- data.frame(
  feature    = names(importance),
  importance = as.numeric(importance)
)
imp_df <- imp_df[order(imp_df$importance, decreasing = TRUE), ]
imp_df$feature <- factor(imp_df$feature, levels = imp_df$feature)

library(ggplot2)

ggplot(imp_df, aes(x = feature, y = importance)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_hline(yintercept = 0, colour = "grey40") +
  labs(x = NULL, y = "Accuracy drop when permuted",
       title = "Permutation variable importance") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Using `ranger` in production

The implementation above illustrates the concept. In production, use the `ranger` package [@wright2017], which implements the full algorithm (continuous splits, depth control, proper OOB error estimation) efficiently in C++:

```r
# Production random forest with ranger (eval: false — package may not be installed)
library(ranger)

rf_model <- ranger(
  outbreak ~ vax_coverage + pop_density + prior_cases + surveillance,
  data        = df,
  num.trees   = 500,
  mtry        = 2,           # features per split (sqrt of total is default)
  probability = TRUE,        # return P(outbreak) not just class
  importance  = "permutation"
)

# Variable importance
rf_model$variable.importance
# Prediction on new district
predict(rf_model, new_district)$predictions[, "1"]
```

## Key takeaways

- Random forests require almost no tuning — default hyperparameters are competitive on most tabular problems.
- They naturally handle missing data, mixed feature types, and non-linear interactions.
- Permutation importance is the most reliable variable importance measure because it is model-agnostic.
- For outbreak risk scoring with < 10,000 observations, start with a random forest before reaching for a neural network.

## References

::: {#refs}
:::
