---
title: "How many infections does one isolation prevent?"
author: "Jong-Hoon Kim"
date: "2026-03-10"
categories: [epidemiology, branching process, isolation, overdetermination]
bibliography: ../../../references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

When a newly infected person successfully isolates, how many downstream infections does that prevent? The answer depends on the structure of transmission — not just the average reproduction number $\mathcal{R}_c$ (denoting a reproduction number under control), but also on how heterogeneous transmission is, and critically, on how many other infected individuals are circulating at the same time.

## Setting up the problem

Consider a large, well-mixed susceptible population. Each infected individual generates a random number of secondary infections drawn from an **offspring distribution** with probability mass function $p_k = \Pr(K = k)$ and mean $\mathcal{R}_c$. We consider two canonical choices: Poisson and negative binomial. Both share the same mean $\mathcal{R}_c$, but the negative binomial has an additional dispersion parameter $\kappa = 0.16$, meaning a small fraction of individuals cause a disproportionately large fraction of transmissions (*superspreading*).

We make use of the probability generating function (PGF) of the offspring distribution:
$$
\mu(x) = \sum_{k=0}^{\infty} p_k x^k,
$$
with $\mathcal{R}_c = \mu'(1)$.

## The final size of the epidemic

In a large well-mixed epidemic, the attack rate $\mathcal{A}(\mathcal{R}_c)$ depends only on $\mathcal{R}_c$, not on the shape of the offspring distribution [@roberts2023quantifying]. From any susceptible's perspective, the total number of transmission attempts she receives is Poisson with mean $\mathcal{R}_c \mathcal{A}$, regardless of how heterogeneous transmission is across infectors. She escapes only if she receives zero attempts, giving
$$
\mathcal{A} = 1 - e^{-\mathcal{R}_c \, \mathcal{A}}.
$$

This transcendental equation can be solved iteratively from $\mathcal{A}_0 = 1$.

An important consequence is that $\mathcal{R}_\infty = (1-\mathcal{A})\mathcal{R}_c < 1$ for all $\mathcal{R}_c > 1$, which guarantees that the effective reproduction number at epidemic's end is always subcritical. This ensures the geometric series for $\mathcal{F}$ (below) always converges.

An alternative uses the PGF fixed-point $q = \mu(q)$, whose non-trivial root $q^*$ is the branching-process **extinction probability** — the probability that a single introduction leads to a minor outbreak that dies out on its own. The complement $1-q^*$ is the probability of a major epidemic. For Poisson offspring, $1-q^*$ satisfies the same equation as the SIR attack rate, so the two coincide. For the negative binomial with small $\kappa$, however, strong superspreading means most chains die out quickly, so $1-q^* < \mathcal{A}_{\text{SIR}}$: major epidemics are less likely, even though when one does occur it infects a comparable fraction of the population. Using $1-q^*$ in place of $\mathcal{A}$ therefore overstates the remaining susceptibles, pushing $(1-\mathcal{A})\mathcal{R}_c$ higher — and near $\mathcal{R}_c = 1$ it can exceed 1, causing $\mathcal{F}$ to diverge.

```{r}
#| label: fig-attack-rate
#| fig-cap: "The attack rate $\\mathcal{A}(\\mathcal{R}_c)$ as a function of $\\mathcal{R}_c$. The curve is identical for both offspring distributions — the attack rate depends only on $\\mathcal{R}_c$, not on the shape of the offspring distribution."

library(ggplot2)
library(dplyr)
library(tidyr)
theme_set(theme_minimal(base_size = 14) + theme(legend.position = "bottom", legend.title = element_blank()))

# Attack rate: SIR mean-field formula, distribution-independent
# A = 1 - exp(-R0 * A), guarantees R_inf = (1-A)*R0 < 1 for R0 > 1
solve_attack_rate <- function(R0, kappa = Inf) {
    if (R0 <= 1) {
        return(0)
    }
    tryCatch(
        {
            uniroot(function(A) A - (1 - exp(-R0 * A)), interval = c(1e-5, 1))$root
        },
        error = function(e) 0
    )
}

R0_seq <- seq(0.5, 4, by = 0.05)

df_ar <- data.frame(R0 = R0_seq) %>%
    mutate(A = sapply(R0, solve_attack_rate))

ggplot(df_ar, aes(x = R0, y = A)) +
    geom_line(linewidth = 1) +
    labs(
        x = expression(italic(R)[c]),
        y = expression("Attack rate " * italic(A))
    )
```

## Expected infections averted: the residual offspring distribution

When individual $u$ isolates immediately after becoming infected, she prevents all infections in her **residual offspring** — offspring who would not be reached through any alternative transmission chain.

Of $u$'s $K$ potential offspring, a fraction $\mathcal{A}(\mathcal{R}_c)$ would have been infected anyway via another chain (and all their descendants too). The remaining fraction $1 - \mathcal{A}(\mathcal{R}_c)$ are truly residual. Therefore the expected number of infections averted by isolating $u$ is
$$
\mathcal{R}_\infty = \bigl[1 - \mathcal{A}(\mathcal{R}_c)\bigr] \, \mathcal{R}_c,
$$
which @roberts2023quantifying show equals the **effective reproduction number at the end of the epidemic** — the initial reproduction number scaled down by the fraction of the population still susceptible when the epidemic burns out.

```{r}
#| label: fig-R-inf
#| fig-cap: "The effective reproduction number at the end of the epidemic, $\\mathcal{R}_\\infty$."
#| echo: false

df_ar <- df_ar %>%
    mutate(R_inf = R0 * (1 - A))

ggplot(df_ar, aes(x = R0, y = R_inf)) +
    geom_line(linewidth = 1) +
    labs(
        x = expression(italic(R)[c]),
        y = expression(italic(R)[infinity])
    )
```

Now consider all descendants, not just direct offspring. The expected number of infections averted at generation $g$ is $\mathcal{R}_\infty^g$. Summing over all generations gives a geometric series:
$$
\mathcal{F}(\mathcal{R}_c) = \sum_{g=1}^{\infty} \mathcal{R}_\infty^g = \frac{\mathcal{R}_\infty}{1 - \mathcal{R}_\infty}.
$$

Each successive generation is $\mathcal{R}_\infty$ times the size of the previous one; $\mathcal{F}(\mathcal{R}_c)$ sums this over all generations.

```{r}
#| label: fig-F-R
#| fig-cap: "The expected number of infections averted $\\mathcal{F}(\\mathcal{R}_c)$."
#| echo: false

df_ar <- df_ar %>%
    mutate(F_R = R_inf / (1 - R_inf))

ggplot(df_ar, aes(x = R0, y = F_R)) +
    geom_line(linewidth = 1) +
    coord_cartesian(ylim = c(0, 50)) +
    labs(
        x = expression(italic(R)[c]),
        y = expression(italic(F)(italic(R)[c]))
    )
```

## Overdetermination limits the benefit at high $\mathcal{R}_c$

**Overdetermination** occurs when a susceptible individual has more than one sufficient cause of infection — meaning she would have been infected even if one of the transmission chains leading to her had been broken. In a high-transmission setting, most susceptibles are reached by multiple independent chains. Removing one chain — by isolating one infected individual — therefore has no effect on the eventual fate of many of those susceptibles. In other words, one's descendants are overdetermined except for one's residual descendants. 

Formally, this is reflected in $\mathcal{R}_\infty$ becoming small when $\mathcal{R}_c$ is large: a large epidemic infects almost everyone ($\mathcal{A} \approx 1$), leaving few susceptibles at the end, so $\mathcal{R}_\infty = (1 - \mathcal{A})\mathcal{R}_c \approx 0$ and $\mathcal{F}(\mathcal{R}_c) \approx 0$.

## The benefit is greatest near the epidemic threshold

When $\mathcal{R}_c$ is only slightly above 1, a large epidemic is still possible, but many potential transmission chains die out on their own. A small reduction in transmission — like a single person isolating — can tip a borderline chain from an expanding outbreak to extinction. Formally, as $\mathcal{R}_c \to 1^+$, we have $\mathcal{A} \to 0$ and $\mathcal{R}_\infty \to \mathcal{R}_c$, so $\mathcal{F}(\mathcal{R}_c) \to \mathcal{R}_c / (1 - \mathcal{R}_c) \to \infty$. A careful limiting argument using L'Hôpital's rule shows that the *expected infections averted multiplied by the probability of the epidemic*, $\mathcal{A}(\mathcal{R}_c)[1 + \mathcal{F}(\mathcal{R}_c)]$, approaches **2** as $\mathcal{R}_c \to 1^+$ [@roberts2023quantifying].

```{r}
#| label: fig-A-times-F
#| fig-cap: "The expected benefit multiplied by epidemic probability, $\\mathcal{A}(\\mathcal{R}_c)[1 + \\mathcal{F}(\\mathcal{R}_c)]$. The dashed line indicates the theoretical limit of 2 as $\\mathcal{R}_c \\to 1^+$."
#| echo: false

# Calculate A * (1 + F)
df_ar <- df_ar %>%
    mutate(A_times_1_plus_F = A * (1 + F_R))

ggplot(df_ar %>% filter(R0 > 1), aes(x = R0, y = A_times_1_plus_F)) +
    geom_line(linewidth = 1) +
    geom_hline(yintercept = 2, linetype = "dashed", color = "black") +
    coord_cartesian(ylim = c(1, 2)) +
    labs(
        x = expression(italic(R)[c]),
        y = expression(italic(A) %.% (1 + italic(F)))
    )
```

## The full distribution: rare but large cascades

@roberts2023quantifying also derive the full distribution of infections averted. The derivation rests on two ingredients.

**Ingredient 1: the Galton-Watson total-progeny identity.** The key result is
$$
T(x) = x \cdot \mu\bigl(T(x)\bigr),
$$
where $T(x) = \sum_{t=0}^\infty \Pr(T=t)\,x^t$ is the PGF of the total cascade size $T$. Here is the derivation.

Let $T$ be the total number of individuals in the cascade (root plus all descendants). The root has $K$ offspring, each seeding an independent sub-cascade of the same kind, so
$$
T = 1 + T_1 + T_2 + \cdots + T_K,
$$
where $T_1, T_2, \ldots$ are i.i.d.\ copies of $T$. Since $x^T = x \cdot x^{T_1+\cdots+T_K}$, we have $T(x) = E[x^T] = x \cdot E[x^{T_1+\cdots+T_K}]$. Conditioning on $K = k$, the $T_i$ are independent, so
$$
E\bigl[x^{T_1+\cdots+T_k} \mid K = k\bigr] = \prod_{i=1}^{k} E[x^{T_i}] = \bigl[T(x)\bigr]^k.
$$
Averaging over $K$ using $\mu(s) = \sum_{k=0}^\infty p_k s^k$ evaluated at $s = T(x)$,
$$
T(x) = x \cdot \sum_{k=0}^\infty p_k \bigl[T(x)\bigr]^k = x \cdot \mu\bigl(T(x)\bigr).
$$

**Ingredient 2: the residual offspring distribution.**

Among $u$'s $K$ potential offspring, each is *residual* — not reachable through any alternative chain — independently with probability $\mathcal{R}_\infty$. Call the surviving count $K'$. Conditional on $K = k$, we have $K' \sim \text{Binomial}(k, \mathcal{R}_\infty)$, whose PGF follows from the binomial theorem:
$$
E[s^{K'} \mid K = k] = \sum_{j=0}^{k} \binom{k}{j} \mathcal{R}_\infty^j (1-\mathcal{R}_\infty)^{k-j} s^j = \sum_{j=0}^{k} \binom{k}{j} (\mathcal{R}_\infty s)^j (1-\mathcal{R}_\infty)^{k-j} = \bigl[(1 - \mathcal{R}_\infty) + \mathcal{R}_\infty s\bigr]^k.
$$
Averaging over the random $K$ using the definition $\mu(z) = E[z^K]$, with $z = 1 - \mathcal{R}_\infty + \mathcal{R}_\infty s$:
$$
\mu_{\text{res}}(s) = E[s^{K'}] = E\!\left[(1 - \mathcal{R}_\infty + \mathcal{R}_\infty s)^K\right] = \mu\!\bigl(1 - \mathcal{R}_\infty + \mathcal{R}_\infty\, s\bigr).
$$

**Combining.** The infections averted form a Galton-Watson tree with offspring PGF $\mu_{\text{res}}$. Substituting into the total-progeny identity gives the PGF $\Omega_\infty(x)$ of the number of infections averted:
$$
\Omega_\infty(x) = x\,\mu\bigl[1 - \mathcal{R}_\infty + \mathcal{R}_\infty \Omega_\infty(x)\bigr],
$$
solved iteratively from $\Omega_0(x) = 1$. Note that $\Omega_\infty$ and $\mu$ are distinct objects: $\mu(x)$ encodes the offspring of a single individual (a one-step distribution), while $\Omega_\infty(x)$ encodes the size of an entire cascade.

Near $\mathcal{R}_c = 1$, this distribution has a heavy tail: most isolations prevent zero or very few infections, but occasionally they cut off a large cascade. The large average $\mathcal{F}(\mathcal{R}_c)$ near threshold is driven by these rare high-impact events, a direct consequence of superspreading in the negative-binomial case.

## Computing the distribution via the Cauchy integral

The Taylor coefficients $c_k = \Pr(\text{avert exactly } k \text{ infections})$ of $\Omega_\infty$ have no closed form for a general offspring distribution, but the **Cauchy integral formula** extracts them from values of $\Omega_\infty$ on the unit circle $|x| = 1$:
$$
c_k = \frac{1}{2\pi i} \oint_{|x|=1} \frac{\Omega_\infty(x)}{x^{k+1}} \, dx.
$$
Parametrising by $x = e^{i\theta}$ turns this into a Fourier coefficient,
$$
c_k = \frac{1}{2\pi} \int_0^{2\pi} \Omega_\infty(e^{i\theta})\, e^{-ik\theta}\, d\theta,
$$
so the entire probability mass function $\{c_k\}$ is the **discrete Fourier transform** of $\Omega_\infty$ sampled at $N$ equally-spaced angles. The algorithm is:

1. Set $z_j = e^{2\pi i j/N}$, $j = 0, \ldots, N-1$.
2. Iterate $\Omega_{g+1}(z_j) = z_j\,\mu\!\bigl[1 - \mathcal{R}_\infty + \mathcal{R}_\infty\,\Omega_g(z_j)\bigr]$ from $\Omega_0 \equiv 1$ until convergence (guaranteed for all $|z| \le 1$ when $\mathcal{R}_\infty < 1$).
3. Extract $c_k \approx \tfrac{1}{N}\operatorname{Re}\!\bigl[\operatorname{FFT}(\Omega_\infty)\bigr]_k$.

```{r}
#| label: fig-dist-averted
#| fig-cap: "Distribution of the number of infections averted when one infected individual isolates, computed via the Cauchy integral method. Shown on a log–log scale to reveal the power-law tail near $\\mathcal{R}_c = 1$."

# PGF of the offspring distribution, vectorised over complex x
pgf_mu <- function(x, R0, kappa = Inf) {
    if (is.infinite(kappa)) {
        exp(R0 * (x - 1)) # Poisson
    } else {
        (kappa / (kappa + R0 * (1 - x)))^kappa # Negative binomial
    }
}

# Extract PMF of infections averted via the Cauchy integral / FFT
infections_averted_pmf <- function(R0, kappa = Inf, N = 2^13,
                                   tol = 1e-10, max_iter = 1000) {
    A <- solve_attack_rate(R0)
    if (A == 0) {
        return(NULL)
    }
    R_inf <- (1 - A) * R0

    # Uniformly-spaced points on the unit circle
    z <- exp(2i * pi * seq(0, N - 1) / N)

    # Iterate until convergence
    Omega <- rep(1 + 0i, N)
    for (g in seq_len(max_iter)) {
        Omega_new <- z * pgf_mu(1 - R_inf + R_inf * Omega, R0, kappa)
        if (max(Mod(Omega_new - Omega)) < tol) break
        Omega <- Omega_new
    }

    # Cauchy integral via FFT: c_k = (1/N) * Re[ FFT[Omega] ]_k
    c_k <- Re(fft(Omega)) / N
    c_k[c_k < 0] <- 0

    data.frame(k = seq_along(c_k) - 1L, prob = c_k)
}

R0_vals <- c(1.1, 1.5, 2.0, 3.5)
dist_params <- list(Poisson = Inf, `Negative Binomial (k=0.16)` = 0.16)

df_pmf <- do.call(rbind, lapply(names(dist_params), function(dname) {
    kap <- dist_params[[dname]]
    do.call(rbind, lapply(R0_vals, function(r) {
        pmf <- infections_averted_pmf(r, kap)
        if (is.null(pmf)) {
            return(NULL)
        }
        pmf$Distribution <- dname
        pmf$R0 <- r
        pmf
    }))
}))

ggplot(
    df_pmf |> dplyr::filter(k >= 1, prob > 1e-8, k <= 1000),
    aes(x = k, y = prob, color = factor(R0), linetype = factor(R0))
) +
    geom_line(linewidth = 0.7) +
    facet_wrap(~Distribution) +
    scale_x_log10() +
    scale_y_log10() +
    labs(
        x = "Number of infections averted (k)",
        y = "Probability",
        color = expression(italic(R)[c]),
        linetype = expression(italic(R)[c])
    ) +
    scale_color_brewer(palette = "Set2")
```

At $\mathcal{R}_c = 1$ the distribution follows a power law $c_k \sim k^{-3/2}$ for the Poisson case, visible as a straight line with slope $-3/2$ on the log–log axes. The negative-binomial case has a heavier tail at the same $\mathcal{R}_c$, reflecting the additional variance from superspreading: while both distributions share the same mean $\mathcal{F}(\mathcal{R}_c)$, superspreading inflates the probability of both very small and very large cascades.

## Key takeaways

- **Near the epidemic threshold** ($\mathcal{R}_c \approx 1$), each isolation has the largest expected public-health benefit.
- **At high transmission** ($\mathcal{R}_c \gg 1$), overdetermination greatly reduces the marginal benefit of a single isolation — many infections will occur through alternative chains regardless.
- **Collective compliance has synergistic effects**: as more individuals isolate, the effective $\mathcal{R}_c$ falls, overdetermination declines, and the per-isolator benefit rises.
- The benefit of isolation is **heterogeneous**: most isolations prevent few infections, but rare events prevent large cascades — a point with important implications for how we communicate the value of individual public-health actions.

## Reference

::: {#refs}
:::
