---
title: "Observability for Digital Twin Services: Logging, Metrics, and Alerting"
subtitle: "You cannot fix what you cannot see. Structured logs, health metrics, and automated alerts for a model service that runs overnight. Skill 7 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, monitoring, logging, observability, operations]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The 2 AM problem

Your EnKF runner is scheduled for 2 AM. At 6 AM your client's epidemiologist logs in and the dashboard shows stale data. Was it a database connection failure? A malformed input file? A numerical divergence in the filter? Without structured logs and alerting [@beyer2016], you are piecing together fragments of information under time pressure.

**Observability** is the practice of instrumenting your code so that any failure leaves enough information to diagnose and fix the problem without re-running it. The three pillars are: **logs** (what happened), **metrics** (how healthy is the system), and **alerts** (notify when something crosses a threshold).

## Structured logging in R

Plain `cat()` output is hard to parse. Structured logs write JSON or key-value lines that can be ingested by log aggregators (CloudWatch, Loki, Datadog).

```{r}
#| label: structured-logger

# Minimal structured logger using base R
log_event <- function(level, message, ...) {
  extras <- list(...)
  entry  <- c(
    list(timestamp = format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ"),
         level     = toupper(level),
         message   = message),
    extras
  )
  # Compact JSON-like output
  parts <- mapply(function(k, v) paste0('"', k, '":"', v, '"'),
                  names(entry), entry)
  cat("{", paste(parts, collapse = ", "), "}\n")
}

log_info  <- function(msg, ...) log_event("info",    msg, ...)
log_warn  <- function(msg, ...) log_event("warning", msg, ...)
log_error <- function(msg, ...) log_event("error",   msg, ...)
```

```{r}
#| label: instrumented-enkf

# EnKF runner instrumented with structured logs
run_enkf_instrumented <- function(obs, location_id) {
  log_info("EnKF run started", location = location_id,
           n_obs = length(obs))

  tryCatch({
    # Simulate EnKF (abbreviated)
    set.seed(42)
    N_ens  <- 200
    beta_ens <- rnorm(N_ens, 0.3, 0.05)
    I_ens    <- pmax(1, rnorm(N_ens, tail(obs, 1), 10))

    for (t in seq_along(obs)) {
      # Update step (simplified)
      innov    <- obs[t] - mean(I_ens)
      K        <- cov(I_ens, I_ens) / (var(I_ens) + max(obs[t], 1))
      I_ens    <- I_ens + K * innov
      I_ens    <- pmax(0, I_ens)

      # Log when something looks suspicious
      if (any(!is.finite(I_ens))) {
        log_warn("NaN detected in ensemble", step = t,
                 location = location_id)
        I_ens[!is.finite(I_ens)] <- median(I_ens, na.rm = TRUE)
      }
    }

    result <- list(I_median = median(I_ens),
                   I_lo     = quantile(I_ens, 0.025),
                   I_hi     = quantile(I_ens, 0.975),
                   beta_est = mean(beta_ens))

    log_info("EnKF run completed",
             location   = location_id,
             I_median   = round(result$I_median, 1),
             beta_est   = round(result$beta_est, 4))
    result

  }, error = function(e) {
    log_error("EnKF run failed",
              location = location_id,
              error    = conditionMessage(e))
    NULL
  })
}

# Simulate observations
set.seed(7)
obs_test <- rpois(30, lambda = c(seq(10, 80, length.out = 15),
                                  seq(80, 20, length.out = 15)))

result <- run_enkf_instrumented(obs_test, "district_a")
```

## Health metrics

Beyond logs, expose numeric metrics that can be scraped by monitoring tools like Prometheus. Key metrics for a digital twin service:

```{r}
#| label: metrics
#| fig-cap: "Simulated 30-day metric history for the EnKF service. The run duration (top) and ensemble spread (bottom) are tracked over time — spikes or trends signal developing problems before a complete failure occurs."

# Simulate 30 days of service metrics
set.seed(99)
n_days <- 30
metrics <- data.frame(
  day          = seq_len(n_days),
  run_duration_s = c(rnorm(25, 45, 5), 45, 48, 90, 85, 78),  # spike at day 28
  ensemble_spread = c(rnorm(25, 15, 2), 15, 16, 32, 30, 22),  # matching spike
  obs_count    = rpois(n_days, 1),                              # 0 or 1 per day
  nans_detected = c(rep(0, 20), 0, 0, 0, 1, 2, 0, 0, 2, 3, 0) # increasing
)

library(ggplot2)
library(tidyr)

metrics_long <- pivot_longer(metrics[, c("day","run_duration_s",
                                          "ensemble_spread")],
                              -day, names_to = "metric")

ggplot(metrics_long, aes(x = day, y = value)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_hline(data = data.frame(metric = c("run_duration_s", "ensemble_spread"),
                                threshold = c(80, 25)),
             aes(yintercept = threshold), linetype = "dashed",
             colour = "firebrick") +
  facet_wrap(~ metric, scales = "free_y",
             labeller = as_labeller(c(run_duration_s = "Run duration (s)",
                                      ensemble_spread = "Ensemble spread"))) +
  labs(x = "Day", y = NULL,
       title = "EnKF service metrics",
       subtitle = "Red dashed = alert threshold") +
  theme_minimal(base_size = 13)
```

## Alert logic

```{r}
#| label: alert-logic

# Rule-based alerting
check_alerts <- function(metrics_row) {
  alerts <- character(0)

  if (metrics_row$run_duration_s > 80) {
    alerts <- c(alerts,
      sprintf("Run duration %.0fs exceeds 80s threshold",
              metrics_row$run_duration_s))
  }
  if (metrics_row$ensemble_spread > 25) {
    alerts <- c(alerts,
      sprintf("Ensemble spread %.1f exceeds 25 threshold",
              metrics_row$ensemble_spread))
  }
  if (metrics_row$obs_count == 0) {
    alerts <- c(alerts, "No observations received today")
  }
  if (metrics_row$nans_detected > 0) {
    alerts <- c(alerts,
      sprintf("%d NaN values detected in ensemble",
              metrics_row$nans_detected))
  }
  if (length(alerts) == 0) {
    cat("Day", metrics_row$day, ": OK\n")
  } else {
    cat("Day", metrics_row$day, "ALERT(S):\n")
    for (a in alerts) cat("  -", a, "\n")
  }
}

# Check recent days
for (i in 26:30) check_alerts(metrics[i, ])
```

## Production monitoring stack

| Component | Tool | Purpose |
|-----------|------|---------|
| Log collector | CloudWatch Logs / Loki | Aggregate structured JSON logs |
| Metrics store | Prometheus | Scrape numeric metrics from `/metrics` endpoint |
| Dashboards | Grafana | Visualise metric time series |
| Alerts | Alertmanager / PagerDuty | Page you when thresholds breach |
| Uptime | Pingdom / UptimeRobot | Check API `/health` every 60s |

For FastAPI, add a `/metrics` endpoint using the `prometheus-fastapi-instrumentator` Python package — Prometheus scrapes it automatically every 15 seconds.

## The "on-call" SLA calculation

Sites reliability engineering [@beyer2016] defines availability as:

$$\text{Availability} = 1 - \frac{\text{downtime}}{\text{total time}}$$

A 99.9% SLA ("three nines") allows 8.7 hours of downtime per year — roughly one major incident. A 99.5% SLA allows 43.8 hours — one serious incident per month. Know which you are promising and instrument accordingly.

## References

::: {#refs}
:::
