---
title: "Caching Surrogate Model Outputs for Sub-Second API Responses"
subtitle: "Repeated scenario queries should cost microseconds, not seconds. Redis, in-memory caching, and cache invalidation strategies for simulation services. Skill 9 of 20."
author: "Jong-Hoon Kim"
date: "2026-04-24"
categories: [business skills, caching, Redis, performance, R]
bibliography: references.bib
csl: https://www.zotero.org/styles/vancouver
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## The performance problem

An analyst opens your dashboard and selects "Run scenario: 30% transmission reduction for district_a". Your system runs the surrogate model, queries TimescaleDB, formats the response — 800 milliseconds. She tweaks the slider to 32% and clicks again — another 800 ms. Thirty interactions later, she is frustrated.

But all those slider positions are over the same data. The EnKF state was last updated 12 hours ago, and the surrogate model output for $(β_{\text{reduced}}, \text{district\_a})$ is fully deterministic. The second, third, and subsequent identical requests are **wasted compute**.

**Caching** [@redis2009] stores the result of an expensive computation and returns the stored value instantly on repeated requests — as long as the underlying data has not changed.

## Two levels of caching

**Level 1 — In-process (R session)**. The `memoise` package [@wickham2023memoise] wraps any R function so its results are stored in memory. Cost: zero infrastructure.

**Level 2 — Shared cache (Redis)**. A Redis key-value store holds results accessible by all workers and persists across process restarts. Cost: one Redis container.

## Level 1: memoise in R

```{r}
#| label: memoise-demo

library(memoise)

# Expensive surrogate model call (simulated with Sys.sleep)
slow_surrogate <- function(beta, location_id) {
  Sys.sleep(0.1)   # simulate 100ms model evaluation
  list(
    peak_cases  = round(100000 * 0.4 * beta / 0.35),
    peak_day    = round(25 / beta),
    location_id = location_id
  )
}

# Memoised version — results cached after first call
fast_surrogate <- memoise(slow_surrogate)

# Time 10 calls: first is slow, rest are instant
times <- numeric(5)
for (i in seq_len(5)) {
  t0 <- proc.time()["elapsed"]
  result <- fast_surrogate(beta = 0.35, location_id = "district_a")
  times[i] <- proc.time()["elapsed"] - t0
}
cat(sprintf("Call times (s): %s\n",
            paste(round(times, 4), collapse = ", ")))
cat("First call: model evaluated. Calls 2-5: cache hit.\n")
```

```{r}
#| label: cache-speedup
#| fig-cap: "Cache performance: first call (cold) takes ~100ms; subsequent identical calls (warm) take under 1ms. This is the same pattern whether the cache is in-process (memoise) or Redis."

library(ggplot2)

# Simulate cache performance over many repeated parameter combinations
set.seed(42)
n_queries   <- 200
beta_values <- sample(seq(0.2, 0.5, by = 0.05), n_queries, replace = TRUE)
cache_hits  <- !duplicated(beta_values)   # first occurrence = cold

df_cache <- data.frame(
  query     = seq_len(n_queries),
  beta      = beta_values,
  cold      = cache_hits,
  latency_ms = ifelse(!cache_hits, 0.5,  # cache hit: ~0.5ms
                      100 + rnorm(n_queries, 0, 5))  # cold: ~100ms
)
df_cache$latency_ms[df_cache$latency_ms < 0] <- 0.5

# Running average latency
df_cache$avg_latency <- cumsum(df_cache$latency_ms) / seq_len(n_queries)

ggplot(df_cache, aes(x = query, y = avg_latency)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(data = df_cache[df_cache$cold, ],
             aes(y = latency_ms), colour = "firebrick",
             size = 1.5, alpha = 0.6) +
  labs(x = "Query number", y = "Average latency (ms)",
       title = "Cache performance: cumulative average latency",
       subtitle = "Red dots = cold misses (first occurrence of that β value)") +
  theme_minimal(base_size = 13)
```

## Level 2: Redis cache with Python

In the production FastAPI service, Redis [@redis2009] provides a shared cache across multiple worker processes:

```python
import redis
import json
import hashlib
from functools import wraps

r = redis.Redis(host="localhost", port=6379, db=0)

def cache_result(ttl_seconds: int = 3600):
    """Decorator: cache function result in Redis for ttl_seconds."""
    def decorator(fn):
        @wraps(fn)
        def wrapper(*args, **kwargs):
            # Create deterministic cache key from arguments
            key_data = json.dumps({"args": args, "kwargs": kwargs},
                                   sort_keys=True)
            cache_key = f"{fn.__name__}:{hashlib.md5(key_data.encode()).hexdigest()}"

            # Try cache first
            cached = r.get(cache_key)
            if cached is not None:
                return json.loads(cached)

            # Cache miss: compute and store
            result = fn(*args, **kwargs)
            r.setex(cache_key, ttl_seconds, json.dumps(result))
            return result
        return wrapper
    return decorator

@cache_result(ttl_seconds=3600)   # cache for 1 hour
def run_surrogate(beta: float, location_id: str) -> dict:
    # ... expensive surrogate model evaluation ...
    pass
```

## Cache invalidation: the hard part

The classic programming joke: "There are only two hard things in computer science: cache invalidation and naming things."

For an epidemic digital twin, the cache becomes stale when:

1. **New EnKF update** — state estimate changed → all scenario projections for that location are invalid
2. **New surveillance data** — observations changed → forecasts derived from them are invalid
3. **Model refit** — surrogate retrained on new simulator runs → all surrogate outputs are invalid

**Invalidation strategy**: tag each cache entry with the `run_id` of the last EnKF update. When a new EnKF run completes, increment a counter and use it in all future cache keys — old keys naturally expire (TTL) without explicit deletion.

```{r}
#| label: invalidation

# Cache versioning: key includes run_id
make_cache_key <- function(fn_name, beta, location, run_id) {
  paste0(fn_name, ":", location, ":", beta, ":", run_id)
}

# When EnKF updates, new run_id invalidates all old cache entries
run_id_v1 <- "enkf-2024-04-01"
run_id_v2 <- "enkf-2024-04-02"   # new update

key_v1 <- make_cache_key("surrogate", 0.35, "district_a", run_id_v1)
key_v2 <- make_cache_key("surrogate", 0.35, "district_a", run_id_v2)

cat("Old key:", key_v1, "\n")
cat("New key:", key_v2, "\n")
cat("Old entries expire naturally; no explicit flush needed.\n")
```

## Sizing the cache

For a typical digital twin with 20 districts × 50 scenario parameter combinations × 1KB per result = 1MB per EnKF run. Redis defaults to 256MB — more than enough for years of runs. Use `MAXMEMORY-POLICY allkeys-lru` to evict the least-recently-used entries when memory fills.

## References

::: {#refs}
:::
